---
title: "03-subclustering.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(presto)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "main-lineage-processing")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```


## Load data
```{r load-seurat-object}
obj <- readRDS(file.path(results_dir, "cart_annotated.rds"))
```

## Visualize annotated Seurat object
```{r visualize-annotated-object, fig.height=6, fig.width=7}
DimPlot(obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Annotated Lineages")
```
## Separated by CAR-T conditions
```{r visualize-annotated-object-by-condition, fig.height=6, fig.width=10}
DimPlot(obj, reduction = "umap", group.by = "cell_type", split.by = "condition", ncol = 3) + 
  ggtitle("Annotated Lineages by Condition")
```

# Myeloid subclustering

## Subset myeloid cells
```{r subset-myeloid-cells}
myeloid_cells <- c("Macrophages", "Microglia", "Monocytes")
myeloid_obj <- subset(obj, subset = cell_type %in% myeloid_cells)
```


## Reprocess myeloid cells
```{r reprocess-myeloid-cells, verbose=FALSE}
myeloid_obj[['RNA']] <- split(myeloid_obj[['RNA']], myeloid_obj$sample)
myeloid_obj <- NormalizeData(myeloid_obj)
myeloid_obj <- FindVariableFeatures(myeloid_obj, selection.method = "vst", nfeatures = 2000)
myeloid_obj <- ScaleData(myeloid_obj)
myeloid_obj <- RunPCA(myeloid_obj, npcs = 30)
ElbowPlot(myeloid_obj, ndims = 30) # use 20 dims

myeloid_obj <- IntegrateLayers(
  object = myeloid_obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca", verbose=FALSE
)

myeloid_obj [["RNA"]] <- JoinLayers(myeloid_obj[["RNA"]])

myeloid_obj <- FindNeighbors(myeloid_obj, reduction = "integrated.cca", dims = 1:20)
myeloid_obj <- FindClusters(myeloid_obj, resolution = 0.5)

myeloid_obj <- RunUMAP(myeloid_obj, reduction = "integrated.cca", dims = 1:20)
```

## Visualize myeloid subclusters
```{r visualize-myeloid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Cell Lineages")

p1 + p2
```

## Split by condition
```{r visualize-myeloid-subclusters-by-condition, fig.height=6, fig.width=8}
DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3) + 
  ggtitle("Myeloid Subclusters by Condition")
```
## Save myeloid object
```{r save-myeloid-object}
saveRDS(myeloid_obj, file.path(results_dir, "cart_myeloid_subclusters.rds"))
```

# NK/T subclustering

## Subset NK/T cells
```{r subset-lymphoid-cells}
nkt_obj <- subset(obj, subset = cell_type %in% c("NK/T cells"))
```

## Reprocess lymphoid cells
```{r reprocess-lymphoid-cells, verbose=FALSE}
nkt_obj[['RNA']] <- split(nkt_obj[['RNA']], nkt_obj$sample)
nkt_obj <- NormalizeData(nkt_obj)
nkt_obj <- FindVariableFeatures(nkt_obj, selection.method = "vst", nfeatures = 2000)
nkt_obj <- ScaleData(nkt_obj)
nkt_obj <- RunPCA(nkt_obj, npcs = 30)
ElbowPlot(nkt_obj, ndims = 30) # use 20 dims

nkt_obj <- IntegrateLayers(
  object = nkt_obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca", verbose=FALSE
)
nkt_obj [["RNA"]] <- JoinLayers(nkt_obj[["RNA"]])

nkt_obj <- FindNeighbors(nkt_obj, reduction = "integrated.cca", dims = 1:20)
nkt_obj <- FindClusters(nkt_obj, resolution = 0.5)

nkt_obj <- RunUMAP(nkt_obj, reduction = "integrated.cca", dims = 1:20)
```

## Visualize lymphoid subclusters
```{r visualize-lymphoid-subclusters, fig.height=6, fig.width=7}
DimPlot(nkt_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) +
  ggtitle("Subclusters")
```

## Split by condition
```{r visualize-lymphoid-subclusters-by-condition, fig.height=6, fig.width=8}
DimPlot(nkt_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3) + 
  ggtitle("Lymphoid Subclusters by Condition")
```

## Save lymphoid object
```{r save-lymphoid-object}
saveRDS(nkt_obj, file.path(results_dir, "cart_lymphoid_subclusters.rds"))
```

# Dendritic cell subclustering

## Subset dendritic cells
```{r subset-dendritic-cells}
dc_obj <- subset(obj, subset = cell_type %in% c("Dendritic cells"))
```

## Reprocess dendritic cells
```{r reprocess-dendritic-cells, verbose=FALSE}
dc_obj[['RNA']] <- split(dc_obj[['RNA']], dc_obj$sample)
dc_obj <- NormalizeData(dc_obj)
dc_obj <- FindVariableFeatures(dc_obj, selection.method = "vst", nfeatures = 2000)
dc_obj <- ScaleData(dc_obj)
dc_obj <- RunPCA(dc_obj, npcs = 30)
ElbowPlot(dc_obj, ndims = 30) # use 20 dims

dc_obj <- IntegrateLayers(
  object = dc_obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca", k.weight=50, verbose=FALSE
)
dc_obj [["RNA"]] <- JoinLayers(dc_obj[["RNA"]])

dc_obj <- FindNeighbors(dc_obj, reduction = "integrated.cca", dims = 1:20)
dc_obj <- FindClusters(dc_obj, resolution = 0.5)
dc_obj <- RunUMAP(dc_obj, reduction = "integrated.cca", dims = 1:20)
```

## Visualize dendritic cell subclusters
```{r visualize-dc-subclusters, fig.height=6, fig.width=7}
DimPlot(dc_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) +
  ggtitle("Dendritic Cell Subclusters")
```

## Split by condition
```{r visualize-dc-subclusters-by-condition, fig.height=6, fig.width=8}
DimPlot(dc_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3) + 
  ggtitle("Dendritic Cell Subclusters by Condition")
```

## Save dendritic cell object
```{r save-dc-object}
saveRDS(dc_obj, file.path(results_dir, "cart_dc_subclusters.rds"))
```
