---
title: "03-myeloid-functional-marker-dotplots"
author: "Bicna Song"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(msigdbr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")
input_dir <- file.path(analysis_dir, "input")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "dotplot_helpers.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(data_dir, "cart_myeloid_subtypes.rds"))

# Define condition order and colors once
condition_levels <- c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")
myeloid_obj$condition <- factor(myeloid_obj$condition, levels = condition_levels)

unique(myeloid_obj$myeloid_subtype)
```

## Define manual gene lists
```{r define-manual-genes}
panelE_categories <- list(
  "Lineage" = c("Adgre1", "Itgam", "Itgax", "Itga4", "Aif1"),
  "Pro-tumorigenic (M1)" = c(
    "Trem1", "Arg1", "Mrc1", "Il4ra", "Chil3", "Ccl6", "Fn1", "Il13ra1", "Ccl24",
    "Stat3", "Egr2", "Hif1a", "Socs2", "Msr1", "Top2a", "Sparc", "Birc5", "Gpnmb", "Cxcl10"
  ),
  "Anti-tumorigenic (M2)" = c(
    "Lyz2", "Il18bp", "Ms4a7", "Thbs1", "Pf4", "Stat1", "Ifi27", "Apoc2", "Il1b",
    "Il1a", "Ifngr2", "Cxcl9", "H2-Eb1", "H2-Ab1"
  ),
  "Differentiation" = c("Cd3g", "Cd3d", "Cd4", "Cd8a")
)

panelF_categories <- list(
  "Immunosuppressive" = c("Cd74", "Ccl4", "Cxcl16", "Lgals3", "Mif", "Apoe", "Id2", "Stat3", "Stat6"),
  "Invasion" = c("Mt1", "Ctsb", "Ctsd"),
  "Recruitment" = c("Cxcl10", "Cxcl9", "Ccr5", "Ccl5"),
  "M1-like" = c("H2-Ab1", "H2-Eb1", "H2-Aa", "Irf7", "Il1b", "Cd86", "Lyz2", "Socs1", "Ifngr2", "Il18bp", "Ms4a7", "Adgre1"),
  "M2-like" = c("Msr1", "Tgfb1", "Pkm", "Axl", "Arg1", "Mrc1", "Il4ra", "Il13ra1", "Hif1a")
)

hallmark <- msigdbr(species = "Mus musculus", category = "H")
```

## Subcluster dot plots

### Visualize top markers per subcluster by dot plot
```{r visualize-subcluster-markers-dotplot, fig.height=16, fig.width=8, warning=FALSE}
myeloid_subcluster_markers <- read.csv(file.path(input_dir, "myeloid_subcluster_markers.csv"))

myeloid_subcluster_markers <- myeloid_subcluster_markers %>%
  filter(p_val_adj < 0.05)

top10_markers <- myeloid_subcluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)

subcluster_hallmark <- prepare_hallmark_categories(
  marker_df = top10_markers,
  hallmark = hallmark,
  categories_to_remove = c(
    "HALLMARK_ADIPOGENESIS",
    "HALLMARK_ALLOGRAFT_REJECTION",
    "HALLMARK_ANDROGEN_RESPONSE",
    "HALLMARK_APICAL_JUNCTION",
    "HALLMARK_CHOLESTEROL_HOMEOSTASIS",
    "HALLMARK_COAGULATION",
    "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"
  )
)

anyDuplicated(subcluster_hallmark$gene_order)

make_dotplot_with_categories(
  obj = myeloid_obj,
  group_by = "seurat_clusters",
  gene_order = subcluster_hallmark$gene_order,
  category_bounds = subcluster_hallmark$category_bounds,
  category_labels = subcluster_hallmark$category_labels,
  file_name = "myeloid_dotplot_subcluster_markers.pdf",
  ylab_text = NULL,
  y_levels = rev(as.character(0:15)),
  width = 20,
  height = 6,
  cat_label_size = 4,
  cat_height = 2.2,
  cat_y_limits = c(-0.55, 0.05),
  show_legend_titles = FALSE
)
```

### Visualize subcluster dot plot using manual gene list from panel E/F
```{r visualize-subcluster-markers-manual-dotplot, fig.height=12, fig.width=8, warning=FALSE}
make_manual_dotplot(
  myeloid_obj,
  group_by = "seurat_clusters",
  categories_list = panelE_categories,
  file_name = "myeloid_dotplot_subcluster_markers_panelE.pdf",
  ylab_text = NULL,
  y_levels = rev(as.character(0:15)),
  width = 15,
  height = 5,
  cat_label_size = 5,
  cat_height = 0.6,
  cat_y_limits = c(-0.25, 0.05),
  show_legend_titles = FALSE
)

make_manual_dotplot(
  myeloid_obj,
  group_by = "seurat_clusters",
  categories_list = panelF_categories,
  file_name = "myeloid_dotplot_subcluster_markers_panelF.pdf",
  ylab_text = NULL,
  y_levels = rev(as.character(0:15)),
  width = 15,
  height = 5,
  cat_label_size = 5,
  cat_height = 0.6,
  cat_y_limits = c(-0.25, 0.05),
  show_legend_titles = TRUE
)
```

## Subtype dot plots
```{r generate-umap}
## Reannotate clusters by their expression profiles
new_cluster_ids <- c(
  "0" = "Early activated microglia",
  "1" = "TAM-like phagocytic microglia",
  "2" = "Stress-response microglia",
  "3" = "Stress-response microglia",
  "4" = "Inflammatory activated microglia",
  "5" = "Angiogenic TAMs",
  "6" = "Antigen-presenting TAMs",
  "7" = "Homeostatic microglia",
  "8" = "Perivascular macrophages",
  "9" = "Lipid-associated TAMs",
  "10" = "Perivascular macrophages (chemokine-producing)",
  "11" = "Antigen-presenting monocytes",
  "12" = "Inflammatory TAMs",
  "13" = "Homeostatic microglia",
  "14" = "cDC1",
  "15" = "Inflammatory monocytes"
)

cluster_label_df <- tibble(
  seurat_cluster = names(new_cluster_ids),
  myeloid_subtype = unname(new_cluster_ids)
) %>%
  group_by(myeloid_subtype) %>%
  summarise(
    cluster_label = paste0("C", paste(seurat_cluster, collapse = "_")),
    .groups = "drop"
  ) %>%
  mutate(
    subtype_label = paste0(cluster_label, ": ", myeloid_subtype)
  )

cluster_label_df <- cluster_label_df %>%
  arrange(as.numeric(gsub("^C([0-9]+).*", "\\1", cluster_label)))

subtype_label_map <- setNames(
  cluster_label_df$subtype_label,
  cluster_label_df$myeloid_subtype
)

myeloid_obj$myeloid_subtype_label <- unname(
  subtype_label_map[as.character(myeloid_obj$myeloid_subtype)]
)

myeloid_obj$myeloid_subtype_label <- factor(
  myeloid_obj$myeloid_subtype_label,
  levels = cluster_label_df$subtype_label
)

cluster_short_map <- setNames(
  cluster_label_df$cluster_label,
  cluster_label_df$myeloid_subtype
)

myeloid_obj$cluster_short_label <- unname(
  cluster_short_map[as.character(myeloid_obj$myeloid_subtype)]
)

p1 <- DimPlot(
  myeloid_obj,
  reduction = "umap",
  group.by = "myeloid_subtype_label",
  label = FALSE,
  repel = TRUE,
  raster = TRUE
) +
  theme(plot.title = element_blank())

pdf(
  file = file.path(plot_dir, "myeloid_umap_subtype_cluster_labels.pdf"),
  width = 10,
  height = 5
)
print(p1)
dev.off()

p2 <- DimPlot(
  myeloid_obj,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  repel = TRUE,
  raster = TRUE
) +
  theme(plot.title = element_blank())

pdf(
  file = file.path(plot_dir, "myeloid_umap_subcluster.pdf"),
  width = 6,
  height = 5
)
print(p2)
dev.off()
```

### Visualize top markers per subtype by dot plot
```{r visualize-subtype-markers-dotplot, fig.height=16, fig.width=8, warning=FALSE}
subtype_axis_labels <- setNames(
  cluster_label_df$cluster_label,
  cluster_label_df$myeloid_subtype
)

subtype_y_levels <- cluster_label_df %>%
  arrange(as.numeric(gsub("^C([0-9]+).*", "\\1", cluster_label))) %>%
  pull(myeloid_subtype)

subtype_y_levels_no_inflam <- subtype_y_levels[subtype_y_levels != "Inflammatory monocytes"]

myeloid_subtype_markers <- read.csv(file.path(input_dir, "myeloid_subtype_markers.csv"))

myeloid_subtype_markers <- myeloid_subtype_markers %>%
  filter(p_val_adj < 0.05)

top10_markers <- myeloid_subtype_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)

subtype_hallmark <- prepare_hallmark_categories(
  marker_df = top10_markers,
  hallmark = hallmark,
  categories_to_remove = c(
    "HALLMARK_ADIPOGENESIS",
    "HALLMARK_ALLOGRAFT_REJECTION",
    "HALLMARK_ANDROGEN_RESPONSE",
    "HALLMARK_APICAL_JUNCTION",
    "HALLMARK_CHOLESTEROL_HOMEOSTASIS",
    "HALLMARK_COAGULATION",
    "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"
  )
)

make_dotplot_with_categories(
  obj = myeloid_obj,
  group_by = "myeloid_subtype",
  gene_order = subtype_hallmark$gene_order,
  category_bounds = subtype_hallmark$category_bounds,
  category_labels = subtype_hallmark$category_labels,
  file_name = "myeloid_dotplot_subtype_markers.pdf",
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels),
  y_axis_labels = subtype_axis_labels,
  width = 20,
  height = 8,
  cat_label_size = 4,
  cat_height = 2.2,
  cat_y_limits = c(-0.55, 0.05),
  show_legend_titles = FALSE
)

make_dotplot_with_categories(
  obj = myeloid_obj,
  group_by = "myeloid_subtype",
  gene_order = subtype_hallmark$gene_order,
  category_bounds = subtype_hallmark$category_bounds,
  category_labels = subtype_hallmark$category_labels,
  file_name = "myeloid_dotplot_subtype_markers_exclude_inflammatory_monocytes.pdf",
  exclude_subtypes = c("Inflammatory monocytes"),
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels_no_inflam),
  y_axis_labels = subtype_axis_labels,
  width = 20,
  height = 8,
  cat_label_size = 4,
  cat_height = 2.2,
  cat_y_limits = c(-0.55, 0.05),
  show_legend_titles = FALSE
)
```

### Visualize subtype dot plot using manual gene lists from panel E/F
```{r visualize-subtype-markers-manual-dotplot}
make_manual_dotplot(
  myeloid_obj,
  group_by = "myeloid_subtype",
  categories_list = panelE_categories,
  file_name = "myeloid_dotplot_subtype_markers_panelE.pdf",
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels),
  y_axis_labels = subtype_axis_labels,
  width = 15,
  height = 5
)

make_manual_dotplot(
  myeloid_obj,
  group_by = "myeloid_subtype",
  categories_list = panelF_categories,
  file_name = "myeloid_dotplot_subtype_markers_panelF.pdf",
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels),
  y_axis_labels = subtype_axis_labels,
  width = 15,
  height = 6
)

make_manual_dotplot(
  myeloid_obj,
  group_by = "myeloid_subtype",
  categories_list = panelE_categories,
  file_name = "myeloid_dotplot_subtype_markers_panelE_exclude_inflammatory_monocytes.pdf",
  exclude_subtypes = c("Inflammatory monocytes"),
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels_no_inflam),
  y_axis_labels = subtype_axis_labels,
  width = 15,
  height = 5
)

make_manual_dotplot(
  myeloid_obj,
  group_by = "myeloid_subtype",
  categories_list = panelF_categories,
  file_name = "myeloid_dotplot_subtype_markers_panelF_exclude_inflammatory_monocytes.pdf",
  exclude_subtypes = c("Inflammatory monocytes"),
  ylab_text = NULL,
  y_levels = rev(subtype_y_levels_no_inflam),
  y_axis_labels = subtype_axis_labels,
  width = 15,
  height = 6
)
```

## Treatment dot plots

### Marker genes on treatment groups
```{r subset-treatment-markers, message=FALSE, warning=FALSE}
myeloid_treatment_markers <- FindAllMarkers(myeloid_obj,
                                            assay = "RNA",
                                            group.by = "condition",
                                            slot = "data",
                                            only.pos = TRUE,
                                            min.pct = 0.5,
                                            logfc.threshold = 0.5)

myeloid_treatment_markers <- myeloid_treatment_markers %>%
  filter(p_val_adj < 0.05)

write.csv(myeloid_treatment_markers, file.path(results_dir, "myeloid_treatment_markers.csv"), row.names = FALSE)
```

### Visualize top markers per treatment by dot plot
```{r visualize-treatment-markers-dotplot, fig.height=16, fig.width=8, warning=FALSE}
myeloid_treatment_markers <- read.csv(file.path(results_dir, "myeloid_treatment_markers.csv"))

treatment_hallmark <- prepare_hallmark_categories(
  marker_df = myeloid_treatment_markers,
  hallmark = hallmark,
  categories_to_remove = c(
    "HALLMARK_ADIPOGENESIS",
    "HALLMARK_ALLOGRAFT_REJECTION",
    "HALLMARK_ANDROGEN_RESPONSE",
    "HALLMARK_APICAL_JUNCTION"
  )
)

make_dotplot_with_categories(
  obj = myeloid_obj,
  group_by = "condition",
  gene_order = treatment_hallmark$gene_order,
  category_bounds = treatment_hallmark$category_bounds,
  category_labels = treatment_hallmark$category_labels,
  file_name = "myeloid_dotplot_treatment_markers.pdf",
  ylab_text = NULL,
  width = 15,
  height = 5,
  y_levels = rev(condition_levels),
  cat_label_size = 4,
  cat_height = 2.2,
  cat_y_limits = c(-0.55, 0.05),
  show_legend_titles = TRUE
)
```

### Visualize treatment dot plot using manual gene list from panel E/F
```{r visualize-treatment-markers-manual-dotplot}
make_manual_dotplot(
  myeloid_obj,
  group_by = "condition",
  categories_list = panelE_categories,
  file_name = "myeloid_dotplot_treatment_markers_panelE.pdf",
  ylab_text = NULL,
  width = 15,
  height = 5,
  y_levels = rev(condition_levels)
)

make_manual_dotplot(
  myeloid_obj,
  group_by = "condition",
  categories_list = panelF_categories,
  file_name = "myeloid_dotplot_treatment_markers_panelF.pdf",
  ylab_text = NULL,
  width = 15,
  height = 5,
  y_levels = rev(condition_levels)
)
```

## Session info
```{r session-info}
sessionInfo()
```
