---
title: "04-myeloid-pseudobulk-gsea-by-treatment"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
  subtype_label: "myeloid"
  condition_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")
  car_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB")
  minSize: 15
  maxSize: 500
  volcano_lfc_cutoff: 1
  volcano_p_cutoff: 0.05
  gsea_p_cutoff: 0.05
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(tibble)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(msigdbr)
library(fgsea)
library(Matrix)
library(DESeq2)
library(stringr)
library(EnhancedVolcano)
library(purrr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "pseudobulk_gsea_helpers.R"))

# set random seed
set.seed(1234)
```

## Load data and define identities
```{r load-data}
subtype_label <- params$subtype_label

subtype_title <- dplyr::case_when(
  subtype_label == "myeloid" ~ "Myeloid",
  subtype_label == "tcell" ~ "T cell",
  subtype_label == "dc" ~ "DC",
  TRUE ~ stringr::str_to_title(subtype_label)
)

obj <- readRDS(file.path(data_dir, params$input_rds))

# Make condition a factor with explicit order so contrasts are stable/reproducible
condition_levels <- intersect(params$condition_levels, unique(as.character(obj$condition)))

obj$condition <- factor(obj$condition, levels = condition_levels)
Idents(obj) <- "condition"

all_conditions <- levels(obj$condition)
car_levels <- intersect(params$car_levels, all_conditions)

cat("Subtype label:", subtype_label, "\n")
cat("Input object:", params$input_rds, "\n")
cat("Conditions:", paste(all_conditions, collapse = ", "), "\n")
cat("CAR conditions:", paste(car_levels, collapse = ", "), "\n")

table(obj$condition)
```

## Load MSigDB Hallmark gene sets
```{r load-msigdb-hallmark}
hallmark_pathways <- msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)

length(hallmark_pathways)
```

## Build pseudobulk counts per sample
This section aggregates raw UMI counts across all cells in the input subtype object belonging to the same sample, yielding a genes × samples count matrix.
```{r build-pseudobulk, message=FALSE, warning=FALSE}
stopifnot(all(c("sample", "condition") %in% colnames(obj@meta.data)))

counts <- GetAssayData(obj, assay = "RNA", layer = "counts")

meta <- obj@meta.data %>%
  dplyr::select(sample, condition) %>%
  mutate(
    sample = as.character(sample),
    condition = factor(condition, levels = levels(obj$condition))
  )

# Aggregate counts by sample (genes x samples)
pb_counts <- Matrix::t(rowsum(Matrix::t(counts), group = meta$sample))

# Sample-level metadata (one row per sample)
coldata <- meta %>%
  as.data.frame() %>%
  { rownames(.) <- NULL; . } %>%
  distinct(sample, condition) %>%
  tibble::column_to_rownames("sample")

# Align matrices
pb_counts <- pb_counts[, rownames(coldata)]

table(coldata$condition)
```

## 1) Each condition vs rest

```{r each-condition-vs-rest, message=FALSE, warning=FALSE}
pb_rest_results <- list()

for (cond in all_conditions) {
  ref_conditions <- setdiff(all_conditions, cond)
  out_prefix <- paste0(cond, "_vs_rest")

  cat("\n===== Running each-vs-rest:", cond, "vs rest =====\n")

  pb_rest_results[[cond]] <- run_pseudobulk_contrast(
    pb_counts = pb_counts,
    coldata = coldata,
    case_condition = cond,
    ref_conditions = ref_conditions,
    ref_label = "rest",
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_", out_prefix),
    subtype_label = subtype_label,
    subtype_title = subtype_title,
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = params$minSize,
    maxSize = params$maxSize,
    volcano_lfc_cutoff = params$volcano_lfc_cutoff,
    volcano_p_cutoff = params$volcano_p_cutoff,
    gsea_p_cutoff = params$gsea_p_cutoff
  )
}
```

## 2) Each condition vs other CARs

For CAR conditions:

- compare to the other CAR conditions

For non-CAR conditions:

- compare to all CAR conditions

```{r each-condition-vs-other-cars, message=FALSE, warning=FALSE}
pb_car_results <- list()

for (cond in all_conditions) {

  if (cond %in% car_levels) {
    ref_conditions <- setdiff(car_levels, cond)
    ref_label <- "otherCAR"
    out_prefix <- paste0(cond, "_vs_otherCAR")
  } else {
    ref_conditions <- car_levels
    ref_label <- "allCAR"
    out_prefix <- paste0(cond, "_vs_allCAR")
  }

  if (length(ref_conditions) == 0) next

  cat("\n===== Running condition-vs-CARs:", cond, "=====\n")

  pb_car_results[[cond]] <- run_pseudobulk_contrast(
    pb_counts = pb_counts,
    coldata = coldata,
    case_condition = cond,
    ref_conditions = ref_conditions,
    ref_label = ref_label,
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_", out_prefix),
    subtype_label = subtype_label,
    subtype_title = subtype_title,
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = params$minSize,
    maxSize = params$maxSize,
    volcano_lfc_cutoff = params$volcano_lfc_cutoff,
    volcano_p_cutoff = params$volcano_p_cutoff,
    gsea_p_cutoff = params$gsea_p_cutoff
  )
}
```

## Session info
```{r session-info}
sessionInfo()
```
