---
title: "06-tcell-reclustering.Rmd"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
---

## Load packages
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tibble)
  library(Seurat)
  library(patchwork)
  library(SeuratObject)
  library(ggplot2)
  library(msigdbr)
  library(fgsea)
  library(Matrix)
  library(DESeq2)
  library(stringr)
  library(EnhancedVolcano)
  library(purrr)
})

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "tcell")
plot_dir <- file.path(analysis_dir, "plots", "tcell")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "pseudobulk_gsea_helpers.R"))

# Set seed for reproducibility
set.seed(1234)

# Analysis parameters
subtype_label <- "tcell"
```

## Load tcell object
```{r load-data}
tcell_obj <- readRDS(file.path(data_dir, "cart_nkt_subtypes.rds"))

table(tcell_obj$tcell_subtype)
```

## Exclude myeloid and NK cells
```{r subset-tcells}
exclude_subtypes <- c("Myeloid cells", "NK cells")

tcell_obj <- subset(
  tcell_obj,
  subset = !tcell_subtype %in% exclude_subtypes
)

tcell_obj$tcell_subtype <- droplevels(tcell_obj$tcell_subtype)

table(tcell_obj$tcell_subtype)
```

## Reprocess and re-cluster T cells
```{r recluster-tcells, message=FALSE, warning=FALSE}
set.seed(1234)

tcell_obj[["RNA"]] <- split(tcell_obj[["RNA"]], tcell_obj$sample)

tcell_obj <- NormalizeData(tcell_obj)
tcell_obj <- FindVariableFeatures(tcell_obj, selection.method = "vst", nfeatures = 2000)
tcell_obj <- ScaleData(tcell_obj)
tcell_obj <- RunPCA(tcell_obj, npcs = 30)

ElbowPlot(tcell_obj, ndims = 30)

tcell_obj <- IntegrateLayers(
  object = tcell_obj,
  method = CCAIntegration,
  orig.reduction = "pca",
  new.reduction = "integrated.cca",
  verbose = FALSE
)

tcell_obj[["RNA"]] <- JoinLayers(tcell_obj[["RNA"]])

tcell_obj <- FindNeighbors(tcell_obj, reduction = "integrated.cca", dims = 1:20)
tcell_obj <- FindClusters(tcell_obj, resolution = 0.5)
tcell_obj <- RunUMAP(tcell_obj, reduction = "integrated.cca", dims = 1:20)
```

## UMAP of re-clustered T cells
```{r umap-reclustered, fig.height=6, fig.width=7}
p_umap_clusters <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  label.size = 4
) +
  ggtitle("Re-clustered T cell subclusters")

p_umap_clusters

ggsave(
  filename = file.path(plot_dir, "tcell_reclustered_umap.pdf"),
  plot = p_umap_clusters,
  width = 7,
  height = 6,
  device = cairo_pdf
)
```

## UMAP split by treatment
```{r umap-by-treatment, fig.height=6, fig.width=12}
p_umap_split <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "seurat_clusters",
  split.by = "condition",
  ncol = 3,
  label = FALSE
) +
  ggtitle("Re-clustered T cell subclusters by treatment")

p_umap_split

ggsave(
  filename = file.path(plot_dir, "tcell_reclustered_umap_by_treatment.pdf"),
  plot = p_umap_split,
  width = 12,
  height = 6,
  device = cairo_pdf
)
```

## Generate cluster markers
```{r find-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(
  tcell_obj,
  assay = "RNA",
  slot = "data",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.5,
  random.seed = 1234,
  max.cells.per.ident = Inf
)

write.csv(
  cluster_markers,
  file.path(results_dir, "tcell_reclustered_markers.csv"),
  row.names = FALSE
)
```

## Top 30 markers per cluster
```{r top30-markers}
top30_markers <- cluster_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 30, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

write.csv(
  top30_markers,
  file.path(results_dir, "tcell_reclustered_top30_markers_per_cluster.csv"),
  row.names = FALSE
)

top30_markers
```

## UMAP colored by previous subtype labels
```{r umap-old-labels, fig.height=6, fig.width=8}
p_old_labels <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "tcell_subtype",
  label = TRUE,
  label.size = 3
) +
  ggtitle("T cells colored by previous subtype labels")

p_old_labels

ggsave(
  filename = file.path(plot_dir, "tcell_old_subtype_labels_on_reclustered_umap.pdf"),
  plot = p_old_labels,
  width = 8,
  height = 6,
  device = cairo_pdf
)
```

## New clusters vs old subtype labels
```{r cluster-vs-old-labels}
cluster_label_df <- tcell_obj@meta.data %>%
  dplyr::filter(!is.na(seurat_clusters), !is.na(tcell_subtype)) %>%
  dplyr::count(seurat_clusters, tcell_subtype, name = "n") %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop_within_cluster = n / sum(n)) %>%
  dplyr::ungroup()

cluster_label_df
```

## Top old label per new cluster
```{r top-old-label-per-cluster}
cluster_label_summary <- cluster_label_df %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::slice_max(order_by = prop_within_cluster, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::rename(
    matched_old_label = tcell_subtype,
    matched_count = n,
    matched_prop = prop_within_cluster
  ) %>%
  dplyr::arrange(as.numeric(as.character(seurat_clusters)))

write.table(
  cluster_label_summary,
  file = file.path(results_dir, "tcell_top_old_label_per_new_cluster.tsv"),
  sep = "	",
  quote = FALSE,
  row.names = FALSE
)

cluster_label_summary
```

## Barplot of top old label match per new cluster
```{r top-old-label-barplot, fig.height=5, fig.width=9}
p_cluster_label_summary <- ggplot(
  cluster_label_summary,
  aes(
    x = seurat_clusters,
    y = matched_prop,
    fill = matched_old_label
  )
) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(matched_prop, accuracy = 0.1)),
    vjust = -0.3,
    size = 3
  ) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  ) +
  xlab("New Seurat cluster") +
  ylab("Matched proportion") +
  labs(fill = "Top old label") +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1.05)
  ) +
  ggtitle("Top old subtype label match per new cluster")

p_cluster_label_summary

ggsave(
  filename = file.path(plot_dir, "tcell_top_old_label_per_new_cluster_barplot.pdf"),
  plot = p_cluster_label_summary,
  width = 9,
  height = 5,
  device = cairo_pdf
)
```

## Heatmap of new clusters vs old subtype labels
```{r cluster-vs-old-label-heatmap, fig.height=6, fig.width=10}
p_cluster_label_heatmap <- ggplot(
  cluster_label_df,
  aes(x = tcell_subtype, y = seurat_clusters, fill = prop_within_cluster)
) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#2C7FB8") +
  theme_Publication() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  xlab("Previous subtype label") +
  ylab("New Seurat cluster") +
  labs(fill = "Within-cluster
proportion") +
  ggtitle("Mapping of new clusters to previous subtype labels")

p_cluster_label_heatmap

ggsave(
  filename = file.path(plot_dir, "tcell_cluster_vs_old_subtype_heatmap.pdf"),
  plot = p_cluster_label_heatmap,
  width = 10,
  height = 6,
  device = cairo_pdf
)
```

## Old subtype key markers
```{r old-subtype-key-markers}
old_subtype_markers <- list(
  "Heat-shocked T cells" = c("Hspa1a", "Hspa1b", "Dnajb1", "Jun", "Cd69", "Ms4a4b", "Rasal3", "Btg2"),
  "CD4+ Regulatory T cells" = c("Foxp3", "Ctla4", "Il2ra", "Ikzf2", "Tnfrsf4"),
  "CD8+ Cytotoxic T cells" = c("Cd8a", "Cd8b1", "Gzmb", "Nkg7", "Prf1", "Pdcd1"),
  "CD4+ Helper T cells" = c("Cd4", "Icos", "Tnfrsf4", "Bhlhe40"),
  "Proliferating" = c("Mki67", "Cdk1", "Ccnb1", "Ccnb2", "Top2a"),
  "iNKT1 cells" = c("Eomes", "Tbx21", "Prf1", "Gzma", "Nkg7"),
  "Naive T cells" = c("Satb1", "Txk", "Cd7", "Tcf7", "Lef1"),
  "Central memory T cells" = c("Sell", "S1pr1", "Lef1", "Foxp1"),
  "IFN-stimulated T cells" = c("Ifit3", "Isg15", "Oas3", "Zbp1"),
  "iNKT17 cells" = c("Rorc", "Il23r", "Il1r1", "Cxcr6", "Zbtb16"),
  "iNKT2 cells" = c("Gata3", "Il17rb", "Il1rl1", "Areg")
)
```

## Marker validation plots for each old subtype
```{r old-subtype-marker-validation, fig.height=6, fig.width=8}
vln_plot_list <- list()

for (subtype_name in names(old_subtype_markers)) {
  markers_use <- old_subtype_markers[[subtype_name]]
  markers_use <- markers_use[markers_use %in% rownames(tcell_obj)]

  if (length(markers_use) == 0) next

  p_vln <- VlnPlot(
    tcell_obj,
    features = markers_use,
    group.by = "seurat_clusters",
    pt.size = 0,
    ncol = 2
  ) +
    plot_annotation(title = paste0(subtype_name, " key markers across new clusters"))

  vln_plot_list[[subtype_name]] <- p_vln
}

pdf(
  file = file.path(plot_dir, "tcell_old_subtype_marker_violinplots.pdf"),
  width = 12,
  height = 8,
  onefile = TRUE
)
for (plot_name in names(vln_plot_list)) {
  print(vln_plot_list[[plot_name]])
}
dev.off()

# Print the last plot in the knitted document as a visual check
if (length(vln_plot_list) > 0) {
  vln_plot_list[[length(vln_plot_list)]]
}
```

## Rename clusters
```{r rename-clusters}
new_cluster_ids <- c(
  "0"  = "Quiescent T cells",
  "1"  = "Treg",
  "2"  = "Activated effector CD8 T cells (stress)",
  "3"  = "NK-like cytotoxic lymphocytes",
  "4"  = "Cycling / proliferating CAR T cells",
  "5"  = "Activated CD4 effector / helper-like T cells",
  "6"  = "Antigen-presenting myeloid cells",
  "7"  = "Naïve / central-memory T cells",
  "8"  = "Terminally differentiated cytotoxic T cells",
  "9"  = "γδ T cells",
  "10" = "IFN-responsive T cells",
  "11" = "γδ / Th17-like T cells",
  "12" = "Th2-like inflammatory T cells"
)

tcell_obj$tcell_subtype <- plyr::mapvalues(
  x = tcell_obj$seurat_clusters,
  from = names(new_cluster_ids),
  to = new_cluster_ids
) 
```

## UMAP colored by renmaed clusters
```{r umap-rename-clusters, fig.height=6, fig.width=8}
legend_labels <- paste0(
  "C", sort(unique(tcell_obj$seurat_clusters)), ": ",
  new_cluster_ids[as.character(sort(unique(tcell_obj$seurat_clusters)))]
)

p_rename_clusters <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "tcell_subtype",
  label = FALSE
) +
  scale_color_discrete(labels = legend_labels) +  # ✅ legend only
  theme(
    plot.title = element_blank()
  )

p_rename_clusters

ggsave(
  filename = file.path(plot_dir, "tcell_rename_clustered_umap.pdf"),
  plot = p_rename_clusters,
  width = 10,
  height = 6,
  device = cairo_pdf
)
```

## Stacked barplot of subtype proportions
```{r subtype-stacked-barplot}
prop_df <- tcell_obj@meta.data %>%
  dplyr::filter(!is.na(condition), !is.na(tcell_subtype)) %>%
  dplyr::count(condition, cell_group = tcell_subtype, name = "n") %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

write.table(
  prop_df,
  file = file.path(results_dir, "tcell_reclustered_proportions.tsv"),
  sep = "	",
  quote = FALSE,
  row.names = FALSE
)

p_prop <- ggplot(prop_df, aes(x = condition, y = prop, fill = cell_group)) +
  geom_bar(stat = "identity") +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background  = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_manual(
    values = DiscretePalette(n = length(unique(prop_df$cell_group)))
  ) +
  labs(fill = NULL) +
  ylab("Proportion") +
  xlab("Condition") +
  scale_y_continuous(labels = scales::percent_format())

p_prop

ggsave(
  filename = file.path(plot_dir, "tcell_reclustered_proportions_stacked_barplot.pdf"),
  plot = p_prop,
  width = 8,
  height = 5,
  device = cairo_pdf
)
```

## Load MSigDB Hallmark gene sets
```{r load-msigdb-hallmark}
hallmark_pathways <- msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)

length(hallmark_pathways)
```

## Pseudobulk GSEA: each tcell_subtype vs rest

```{r}
make_safe_filename <- function(x) {
  x %>%
    stringr::str_replace_all("γδ", "gamma_delta") %>%      # explicit handling
    stringi::stri_trans_general("Latin-ASCII") %>%         # ✅ Naïve → Naive
    stringr::str_replace_all("[^A-Za-z0-9]", "_") %>%
    stringr::str_replace_all("_+", "_") %>%
    stringr::str_remove("^_+|_+$")
}
```

```{r each-tcell-subtype-vs-rest, message=FALSE, warning=FALSE}
if (!"tcell_subtype" %in% colnames(tcell_obj@meta.data)) {
  stop("The object does not contain a tcell_subtype metadata column.")
}

counts <- GetAssayData(tcell_obj, assay = "RNA", layer = "counts")

subtype_levels <- tcell_obj@meta.data %>%
  dplyr::pull(tcell_subtype) %>%
  as.character() %>%
  unique() %>%
  sort()

tcell_subtype_gsea_results <- list()

for (subtype in subtype_levels) {
  sample_subtype <- tcell_obj@meta.data %>%
    dplyr::select(sample, tcell_subtype) %>%
    mutate(
      sample = as.character(sample),
      subtype_group = ifelse(tcell_subtype == subtype, subtype, "rest"),
      sample_group = paste(sample, subtype_group, sep = "__")
    )

  pb_counts_subtype <- Matrix::t(
    rowsum(Matrix::t(counts), group = sample_subtype$sample_group)
  )

  coldata_subtype <- sample_subtype %>%
    distinct(sample_group, subtype_group) %>%
    dplyr::rename(condition = subtype_group) %>%
    tibble::remove_rownames() %>%
    tibble::column_to_rownames("sample_group")

  coldata_subtype$condition <- factor(coldata_subtype$condition, levels = c("rest", subtype))
  pb_counts_subtype <- pb_counts_subtype[, rownames(coldata_subtype), drop = FALSE]

  if (sum(coldata_subtype$condition == subtype) < 1 || sum(coldata_subtype$condition == "rest") < 1) {
    message("Skipping subtype ", subtype, " because there are not enough samples in one of the groups.")
    next
  }
  
  cat("\n===== Running tcell_subtype vs rest for:", subtype, "====\n")
  print(table(coldata_subtype$condition))

  safe_subtype <- make_safe_filename(subtype)
  out_prefix <- paste0(safe_subtype, "_vs_rest")

  tcell_subtype_gsea_results[[subtype]] <- run_pseudobulk_contrast(
    pb_counts = pb_counts_subtype,
    coldata = coldata_subtype,
    case_condition = subtype,
    ref_conditions = "rest",
    ref_label = "rest",
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_", out_prefix),
    subtype_label = subtype_label,
    subtype_title = "T cell",
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = 15,
    maxSize = 500,
    volcano_lfc_cutoff = 1,
    volcano_p_cutoff = 0.05,
    gsea_p_cutoff = 0.05
  )
}
```

## Save object
- Save updated T-cell object (v2) with refined subtype annotations and subtype-specific GSEA results.
```{r save-object}
saveRDS(
  tcell_obj,
  file.path(results_dir, "cart_tcell_subtypes_v2.rds")
)
```

## Session info
```{r session-info}
sessionInfo()
```
