---
title: "10-tcell-milor-analysis"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_tcell_subtypes.rds"
  subtype_label: "tcell"
  condition_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")
  reference_condition: "tumor"
  milo_reduction: "INTEGRATED.CCA"
  milo_plot_reduction: "UMAP"
  milo_dims: 20
  milo_k: 30
  milo_prop: 0.1
  save_neighborhood_graphs: false
---

## Setup
This block loads the packages required for converting the reclustered T cell Seurat object to a `SingleCellExperiment`, running miloR, and saving outputs.
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tibble)
  library(Seurat)
  library(SeuratObject)
  library(SingleCellExperiment)
  library(miloR)
  library(ggplot2)
})

root_dir <- find_root(has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "tcell")
plot_dir <- file.path(analysis_dir, "plots", "tcell")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(analysis_dir, "util", "milor_helpers.R"))

set_seed(1234)
```

## Parameters
This block records the T cell input object, metadata columns, miloR graph settings, and tumor reference used for all contrasts.
```{r parameters}
input_rds <- params$input_rds
subtype_label <- params$subtype_label
condition_levels <- params$condition_levels
reference_condition <- params$reference_condition
milo_reduction <- params$milo_reduction
milo_plot_reduction <- params$milo_plot_reduction
milo_dims <- params$milo_dims
milo_k <- params$milo_k
milo_prop <- params$milo_prop
save_neighborhood_graphs <- params$save_neighborhood_graphs

cell_group_col <- paste0(subtype_label, "_subtype")
analysis_label <- paste0(subtype_label, "_subtype")

cat("Input RDS:", input_rds, "\n")
cat("Cell group column:", cell_group_col, "\n")
cat("Condition levels:", paste(condition_levels, collapse = ", "), "\n")
cat("Reference condition:", reference_condition, "\n")
cat("Milo reduction:", milo_reduction, "\n")
cat("Milo plotting reduction:", milo_plot_reduction, "\n")
cat("Milo dims:", milo_dims, "\n")
cat("Milo k:", milo_k, "\n")
cat("Milo neighborhood sampling proportion:", milo_prop, "\n")
cat("Save native miloR neighborhood graphs:", save_neighborhood_graphs, "\n")
```

## Load T Cell Object
This block loads the refined T cell Seurat object from the downstream results folder and checks that required metadata is present.
```{r load-data}
obj <- readRDS(file.path(results_dir, input_rds))

stopifnot(cell_group_col %in% colnames(obj@meta.data))
stopifnot("sample" %in% colnames(obj@meta.data))
stopifnot("condition" %in% colnames(obj@meta.data))
stopifnot(reference_condition %in% condition_levels)

table(obj$condition)
table(obj@meta.data[[cell_group_col]])
```

## Prepare SingleCellExperiment
This block converts the Seurat object to a `SingleCellExperiment`, standardizes metadata names, and copies Seurat reductions if they were not transferred automatically.
```{r prepare-single-cell-experiment}
milo_prepared <- prepare_milo_sce(
  obj = obj,
  cell_group_col = cell_group_col,
  condition_levels = condition_levels,
  milo_reduction = milo_reduction,
  milo_plot_reduction = milo_plot_reduction,
  milo_dims = milo_dims
)

milo_sce <- milo_prepared$sce
milo_d <- milo_prepared$d
milo_layout <- milo_prepared$layout

cat("Milo graph reduction:", milo_reduction, "\n")
cat("Milo graph dimensions:", milo_d, "\n")
cat("Milo plot layout:", milo_layout, "\n")
```

## Build Milo Neighborhoods
This block builds the KNN graph, samples representative neighborhoods, and counts cells per biological sample for differential abundance testing.
```{r build-milo-neighborhoods, message=FALSE, warning=FALSE}
milo_obj <- build_milo_neighborhoods(
  milo_sce = milo_sce,
  milo_reduction = milo_reduction,
  milo_d = milo_d,
  milo_k = milo_k,
  milo_prop = milo_prop,
  plot_size_hist = TRUE
)
```

## Build Design Matrix
This block creates one row per sample for miloR and aligns the design table to the neighborhood count matrix.
```{r build-design-matrix}
milo_design <- build_milo_design(milo_obj, reference_condition)

table(milo_design$condition)
```

## Test Neighborhood Differential Abundance
This block tests each treatment condition against tumor and annotates neighborhoods by the dominant T cell subtype.
```{r test-neighborhood-da, message=FALSE, warning=FALSE}
milo_test <- run_milo_da(
  milo_obj = milo_obj,
  milo_design = milo_design,
  condition_levels = condition_levels,
  reference_condition = reference_condition,
  milo_reduction = milo_reduction,
  milo_d = milo_d
)

milo_obj <- milo_test$milo_obj
milo_da_results <- milo_test$da_results

write.table(
  milo_da_results,
  file = file.path(results_dir, paste0(analysis_label, "_milor_", safe_label(reference_condition), "_baseline_results.tsv")),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)
```

## Summarize MiloR Trends by T Cell Subtype
This block summarizes FDR-adjusted and nominal neighborhoods by annotated T cell subtype. The plots emphasize nominal `PValue` trends and effect sizes, and should not be interpreted as FDR-significant DA calls unless neighborhoods pass the adjusted thresholds.
```{r summarize-neighborhood-da, fig.height=6, fig.width=8}
milo_summary <- summarize_milo_da(milo_da_results)

milo_da_results <- milo_summary$da_results
milo_da_threshold_summary <- milo_summary$threshold_summary
milo_da_effect_summary <- milo_summary$effect_summary

write.table(
  milo_da_threshold_summary,
  file = file.path(results_dir, paste0(analysis_label, "_milor_", safe_label(reference_condition), "_baseline_threshold_summary.tsv")),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

write.table(
  milo_da_effect_summary,
  file = file.path(results_dir, paste0(analysis_label, "_milor_", safe_label(reference_condition), "_baseline_effect_summary.tsv")),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE
)

print(milo_da_threshold_summary)

if (all(milo_da_threshold_summary$n_nominal == 0)) {
  cat("No neighborhoods pass nominal PValue < 0.05 in any contrast.\n")
}
if (all(milo_da_threshold_summary$n_spatial_fdr_0_05 == 0)) {
  cat("No neighborhoods pass SpatialFDR < 0.05 in any contrast.\n")
}

p_milo_effect_dot <- plot_milo_effect_dot(
  milo_da_effect_summary,
  title = "T cell miloR exploratory nominal trends"
)

p_milo_effect_dot

ggsave(
  filename = file.path(plot_dir, paste0(analysis_label, "_milor_", safe_label(reference_condition), "_baseline_effect_dotplot.pdf")),
  plot = p_milo_effect_dot,
  width = 8,
  height = 6,
  device = cairo_pdf
)

p_milo_logfc_box <- plot_milo_logfc_box(
  milo_da_results,
  milo_da_effect_summary,
  title = "T cell miloR logFC distributions"
)

p_milo_logfc_box

ggsave(
  filename = file.path(plot_dir, paste0(analysis_label, "_milor_", safe_label(reference_condition), "_baseline_logFC_boxplot.pdf")),
  plot = p_milo_logfc_box,
  width = 15,
  height = 8,
  device = cairo_pdf
)
```

## Optional Native MiloR Neighborhood Graphs
This optional block saves one native miloR neighborhood graph per contrast. These graphs are useful for exploratory QC but can be visually sparse when no neighborhoods pass the FDR threshold, so they are skipped by default.
```{r visualize-neighborhood-da, fig.height=6, fig.width=7}
if (isTRUE(save_neighborhood_graphs)) {
  save_optional_milo_graphs(
    milo_obj = milo_obj,
    milo_da_results = milo_da_results,
    milo_layout = milo_layout,
    analysis_label = analysis_label,
    reference_condition = reference_condition,
    population_title = "T cell",
    plot_dir = plot_dir
  )
} else {
  cat("Skipping native miloR neighborhood graphs. Set params$save_neighborhood_graphs to TRUE to save them.\n")
}
```

## Session Info
This block records package versions and the R environment used for the miloR analysis.
```{r session-info}
sessionInfo()
```
