---
title: "11-tcell-composition-analysis"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_tcell_subtypes.rds"
  subtype_label: "tcell"
  condition_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")
---

## Setup
This block loads packages and defines module paths for the T cell composition analysis.
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tibble)
  library(Seurat)
  library(ggplot2)
  library(sccomp)
})

root_dir <- find_root(has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "tcell")
plot_dir <- file.path(analysis_dir, "plots", "tcell")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(analysis_dir, "util", "sccomp_helpers.R"))
```

## Parameters
This block records the reclustered T cell input object, subtype metadata column, condition order, and sccomp reference levels.
```{r parameters}
input_rds <- params$input_rds
subtype_label <- params$subtype_label
condition_levels <- params$condition_levels

cell_group_col <- paste0(subtype_label, "_subtype")
analysis_label <- cell_group_col

condition_levels_tumor_reference <- c("tumor", setdiff(condition_levels, "tumor"))

cat("Input RDS:", input_rds, "\n")
cat("Subtype label:", subtype_label, "\n")
cat("Cell group column:", cell_group_col, "\n")
cat("Condition levels:", paste(condition_levels, collapse = ", "), "\n")
```

## Load Reclustered T Cell Object
This block loads the refined T cell object generated by `06-tcell-reclustering.Rmd` and verifies that the metadata required for composition testing is present.
```{r load-data}
obj <- readRDS(file.path(results_dir, input_rds))

stopifnot(cell_group_col %in% colnames(obj@meta.data))
stopifnot("sample" %in% colnames(obj@meta.data))
stopifnot("condition" %in% colnames(obj@meta.data))

table(obj$condition)
table(obj@meta.data[[cell_group_col]])
```

## Composition Analysis Using sccomp
## 41BB-L is set as the reference condition
This block tests refined T cell subtype composition changes across conditions using 41BB-L as the reference condition.
```{r composition-analysis-sccomp-41bbl-reference, message=FALSE, warning=FALSE}
sccomp_41bbl <- run_sccomp_composition(
  obj = obj,
  cell_group_col = cell_group_col,
  condition_levels = condition_levels,
  analysis_label = analysis_label,
  output_suffix = "41BBL_baseline",
  reference_label = "41BB-L reference",
  results_dir = results_dir
)

sccomp_input <- sccomp_41bbl$input
sccomp_result <- sccomp_41bbl$result
```

## Plot 41BB-L Reference sccomp Results
This block saves the sccomp interval plot using FDR as the significance statistic.
```{r visualize-sccomp-41bbl-reference-results, fig.height=6, fig.width=12}
sccomp_result |>
  plot_1D_intervals(significance_statistic = c("FDR"))

ggsave(
  file.path(plot_dir, paste0(analysis_label, "_composition_sccomp_41BBL_baseline.pdf")),
  width = 14,
  height = 6,
  device = cairo_pdf
)
```

## Composition Analysis Using sccomp
## tumor is set as the reference condition
This block repeats sccomp with tumor as the reference condition so treatment effects can be read directly as each condition relative to tumor.
```{r composition-analysis-sccomp-tumor-reference, message=FALSE, warning=FALSE}
sccomp_tumor_reference <- run_sccomp_composition(
  obj = obj,
  cell_group_col = cell_group_col,
  condition_levels = condition_levels_tumor_reference,
  analysis_label = analysis_label,
  output_suffix = "tumor_baseline",
  reference_label = "tumor-reference",
  results_dir = results_dir
)

sccomp_input_tumor_reference <- sccomp_tumor_reference$input
sccomp_result_tumor_reference <- sccomp_tumor_reference$result
```

## Plot Tumor Reference sccomp Results
This block saves the tumor-reference sccomp interval plot using FDR as the significance statistic.
```{r visualize-sccomp-tumor-reference-results, fig.height=6, fig.width=12}
sccomp_result_tumor_reference |>
  plot_1D_intervals(significance_statistic = c("FDR"))

ggsave(
  file.path(plot_dir, paste0(analysis_label, "_composition_sccomp_tumor_baseline.pdf")),
  width = 14,
  height = 6,
  device = cairo_pdf
)
```

## Session Info
```{r session-info}
sessionInfo()
```
