---
title: "12-tcell-cd4-cd8-annotation"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

This module prepares CD4-like and CD8-like T-cell trajectory inputs by separating cells using marker expression and ScGate support, annotating them with ProjecTILs and marker programs, clustering each compartment, and saving the prepared CD4-like and CD8-like Seurat objects for Slingshot analysis in script 13.

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(purrr)
  library(Seurat)
  library(SingleCellExperiment)
  library(SeuratExtend)
  library(ggplot2)
  library(patchwork)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "tcell")
plot_dir <- file.path(analysis_dir, "plots", "tcell")

source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "trajectory_helpers.R"))
source(file.path(analysis_dir, "util", "tcell_cd4_cd8_trajectory_helpers.R"))

set.seed(1234)
```

## Parameters
This chunk records the input file, marker thresholds, and marker programs used throughout the analysis.

CD4 and CD8 marker scores are calculated from average marker expression and used to identify CD4-like and CD8-like cells. ProjecTILs labels and functional-state marker programs are kept as supporting evidence for CD4-like and CD8-like cluster annotation and downstream trajectory root selection in script 13.

```{r parameters}
input_rds <- "cart_tcell_subtypes.rds"
min_score_delta <- 0.15
min_marker_score <- 0.10
marker_programs <- list(
  CD4 = c("Cd4"),
  CD8 = c("Cd8a", "Cd8b1"),
  naive_memory = c("Tcf7", "Lef1", "Sell", "Ccr7", "Il7r"),
  activated = c("Cd69", "Il2ra", "Icos", "Tnfrsf4", "Tnfrsf9"),
  treg = c("Foxp3", "Il2ra", "Ctla4", "Tigit", "Ikzf2"),
  cytotoxic = c("Nkg7", "Gzmb", "Gzma", "Prf1", "Ifng", "Ccl5"),
  exhausted = c("Pdcd1", "Lag3", "Havcr2", "Tox", "Tigit", "Ctla4"),
  proliferating = c("Mki67", "Top2a", "Cdk1", "Ccnb1", "Ccnb2"),
  ifn_response = c("Ifit1", "Ifit3", "Isg15", "Oas1a", "Oas3", "Zbp1"),
  stress_response = c("Hspa1a", "Hspa1b", "Dnajb1", "Jun", "Fos")
)

aucell_programs <- marker_programs[c(
  "exhausted",
  "proliferating",
  "naive_memory",
  "cytotoxic"
)]

cat("Input RDS:", input_rds, "\n")
cat("Minimum CD4/CD8 score delta:", min_score_delta, "\n")
cat("Minimum marker score:", min_marker_score, "\n")
```

## Load T-cell object
This chunk loads the renamed T-cell object and confirms that condition metadata is available for downstream pseudotime summaries.

```{r load-data}
input_path <- dplyr::case_when(
  file.exists(file.path(results_dir, input_rds)) ~ file.path(results_dir, input_rds),
  file.exists(file.path(data_dir, input_rds)) ~ file.path(data_dir, input_rds),
  TRUE ~ NA_character_
)

if (is.na(input_path)) {
  stop("Input RDS was not found in results/tcell or data/v5: ", input_rds)
}

cat("Reading:", input_path, "\n")
tcell_obj <- readRDS(input_path)

stopifnot("condition" %in% colnames(tcell_obj@meta.data))

if ("tcell_subtype" %in% colnames(tcell_obj@meta.data)) {
  table(tcell_obj$tcell_subtype)
  tcell_obj$previous_tcell_subtype <- tcell_obj$tcell_subtype
}

table(tcell_obj$condition)

condition_levels <- if (is.factor(tcell_obj$condition)) {
  levels(droplevels(tcell_obj$condition))
} else {
  sort(unique(as.character(tcell_obj$condition)))
}

cat("Conditions:", paste(condition_levels, collapse = ", "), "\n")
```

## CD4/CD8 marker classification
This chunk classifies cells as CD4-like, CD8-like, or ambiguous using CD4 and CD8 marker-expression scores. The marker-score summary checks whether the split captures clear CD4-like and CD8-like populations while leaving uncertain cells as ambiguous.

```{r marker-classification, fig.height=7, fig.width=10}
tcell_obj <- classify_cd4_cd8(tcell_obj)

score_cols <- purrr::map_chr(names(marker_programs), marker_score_column) %>%
  intersect(colnames(tcell_obj@meta.data))

marker_score_summary <- tcell_obj@meta.data %>%
  dplyr::group_by(condition, cd4_cd8_marker_group) %>%
  dplyr::summarise(
    n = dplyr::n(),
    dplyr::across(
      dplyr::all_of(score_cols),
      ~ median(.x, na.rm = TRUE),
      .names = "median_{.col}"
    ),
    .groups = "drop"
  )

write.csv(
  marker_score_summary,
  file.path(results_dir, "tcell_cd4_cd8_marker_score_summary.csv"),
  row.names = FALSE
)

p_marker_feature <- FeaturePlot(
  tcell_obj,
  features = available_features(
    tcell_obj,
    c(marker_programs$CD4, marker_programs$CD8, marker_programs$naive_memory)
  ),
  reduction = "umap",
  ncol = 3
)

p_marker_group <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "cd4_cd8_marker_group"
) +
  ggtitle("CD4/CD8 marker-based groups")

p_marker_scatter <- ggplot(
  tcell_obj@meta.data,
  aes(x = cd4_score, y = cd8_score, color = cd4_cd8_marker_group)
) +
  geom_point(size = 0.25, alpha = 0.45) +
  geom_abline(slope = 1, intercept = min_score_delta, linetype = "dashed", color = "grey35") +
  geom_abline(slope = 1, intercept = -min_score_delta, linetype = "dashed", color = "grey35") +
  theme_Publication() +
  labs(
    title = "CD4/CD8 marker-score split",
    x = "CD4 marker score",
    y = "CD8 marker score",
    color = "Marker group"
  )

p_marker_feature
p_marker_group
p_marker_scatter

ggsave(
  file.path(plot_dir, "tcell_cd4_cd8_marker_expression_umap.pdf"),
  plot = p_marker_feature,
  width = 10,
  height = 7,
  device = cairo_pdf
)

ggsave(
  file.path(plot_dir, "tcell_cd4_cd8_marker_groups_and_score_scatter.pdf"),
  plot = p_marker_group | p_marker_scatter,
  width = 12,
  height = 5,
  device = cairo_pdf
)
```

## ScGate CD4/CD8 support
This chunk runs generic mouse ScGate CD4T/CD8T models on the marker-classified T-cell object. ScGate is used here as supporting evidence for CD4/CD8 identity before creating the clean CD4-like and CD8-like trajectory inputs.

```{r scgate-cd4-cd8-support, fig.height=5, fig.width=7}
tcell_obj <- run_scgate_cd4_cd8(tcell_obj)

scgate_marker_counts <- tcell_obj@meta.data %>%
  dplyr::count(
    condition,
    cd4_cd8_marker_group,
    scgate_cd4_cd8_group,
    name = "n"
  ) %>%
  dplyr::arrange(condition, cd4_cd8_marker_group, scgate_cd4_cd8_group)

write.csv(
  scgate_marker_counts,
  file.path(results_dir, "tcell_cd4_cd8_marker_scgate_concordance.csv"),
  row.names = FALSE
)
```

## ProjecTILs state annotation
This chunk runs ProjecTILs on the marker-classified T-cell object using the default mouse tumor-infiltrating T-cell reference. These labels are used as reference-based support for ambiguous cells, cluster annotation, and naive-like Slingshot root selection.

```{r projectils-annotation, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
tcell_obj <- run_projectils_annotation(tcell_obj)

projectils_counts <- tcell_obj@meta.data %>%
  dplyr::count(
    condition,
    cd4_cd8_marker_group,
    scgate_cd4_cd8_group,
    projectils_label,
    name = "n"
  ) %>%
  dplyr::arrange(condition, cd4_cd8_marker_group, scgate_cd4_cd8_group, dplyr::desc(n))

write.csv(
  projectils_counts,
  file.path(results_dir, "tcell_cd4_cd8_projectils_labels_by_group.csv"),
  row.names = FALSE
)

p_projectils <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "projectils_label",
  label = TRUE,
  repel = TRUE
) +
  ggtitle("ProjecTILs mouse TIL labels")

p_projectils

ggsave(
  file.path(plot_dir, "tcell_cd4_cd8_projectils_labels_umap.pdf"),
  plot = p_projectils,
  width = 8,
  height = 6,
  device = cairo_pdf
)
```

## Prepare clean CD4/CD8 cells
This chunk keeps cells with concordant marker-based and ScGate-supported CD4/CD8 identity and carries ProjecTILs labels into the clean CD4-like and CD8-like trajectory inputs.

```{r prepare-clean-cd4-cd8-cells}
clean_obj <- subset(
  tcell_obj,
  subset = cd4_cd8_marker_group == scgate_cd4_cd8_group &
    cd4_cd8_marker_group %in% c("CD4-like", "CD8-like")
)

clean_obj$cd4_cd8_marker_group <- droplevels(clean_obj$cd4_cd8_marker_group)

clean_cd4_obj <- subset(clean_obj, subset = cd4_cd8_marker_group == "CD4-like")
clean_cd8_obj <- subset(clean_obj, subset = cd4_cd8_marker_group == "CD8-like")
```

## Cluster and annotate CD4-like and CD8-like cells separately
This chunk clusters the clean CD4-like and CD8-like objects once using regular PCA. The clusters are annotated with ProjecTILs, average-expression functional-state marker programs, previous T-cell subtype labels, and SingleR with both mouse references, and a compact cluster annotation table is written.

```{r cluster-cd4-cd8, warning=FALSE, fig.height=6, fig.width=10}
compartment_objects <- list()
compartment_cluster_annotations <- list()
for (compartment in c("CD4-like", "CD8-like")) {
  obj_compartment <- if (compartment == "CD4-like") clean_cd4_obj else clean_cd8_obj

  obj_compartment <- cluster_subset(
    obj_compartment,
    safe_label(tolower(compartment))
  )

  annotated_compartment <- annotate_clustered_object(
    obj_compartment,
    compartment = compartment
  )

  obj_compartment <- annotated_compartment$obj
  obj_compartment <- add_aucell_program_scores(obj_compartment, aucell_programs)
  compartment_cluster_annotations[[compartment]] <- annotated_compartment$cluster_annotation

  obj_compartment$seurat_cluster_projectils_label <- paste0(
    as.character(obj_compartment$seurat_clusters),
    ": ",
    dplyr::coalesce(
      obj_compartment$cluster_projectils_label,
      obj_compartment$top_previous_tcell_subtype,
      "Unannotated"
    )
  )
  projectils_colors <- make_tcell_subtype_palette(
    obj_compartment$seurat_cluster_projectils_label
  )
  compartment_objects[[compartment]] <- obj_compartment

  p <- DimPlot(
    obj_compartment,
    reduction = "umap",
    group.by = "seurat_clusters",
    label = TRUE,
    repel = TRUE
  ) +
    ggtitle(paste(compartment, "clustered UMAP"))

  p_condition <- DimPlot(
    obj_compartment,
    reduction = "umap",
    group.by = "condition"
  ) +
    ggtitle(paste(compartment, "UMAP by condition"))

  p_projectils <- DimPlot(
    obj_compartment,
    reduction = "umap",
    group.by = "seurat_cluster_projectils_label",
    label = TRUE,
    repel = TRUE
  ) +
    scale_color_manual(
      values = projectils_colors,
      breaks = names(projectils_colors),
      drop = FALSE
    ) +
    ggtitle(paste(compartment, "UMAP by cluster ProjecTILs annotation"))

  print(p)
  print(p_condition)
  print(p_projectils)

  ggsave(
    file.path(
      plot_dir,
      paste0("tcell_", safe_label(tolower(compartment)), "_clustered_umap.pdf")
    ),
    plot = p,
    width = 7,
    height = 6,
    device = cairo_pdf
  )

  ggsave(
    file.path(
      plot_dir,
      paste0("tcell_", safe_label(tolower(compartment)), "_clustered_umap_by_condition.pdf")
    ),
    plot = p_condition,
    width = 7,
    height = 6,
    device = cairo_pdf
  )

  ggsave(
    file.path(
      plot_dir,
      paste0("tcell_", safe_label(tolower(compartment)), "_clustered_umap_by_projectils.pdf")
    ),
    plot = p_projectils,
    width = 8,
    height = 6,
    device = cairo_pdf
  )

}

compartment_cluster_annotation_df <- purrr::map_dfr(
  compartment_cluster_annotations,
  ~ .x,
  .id = "analysis"
)

compartment_cluster_annotation_summary <- compartment_cluster_annotation_df %>%
  dplyr::select(
    analysis,
    seurat_clusters,
    n,
    dplyr::any_of(c(
      "functional_state_label",
      "top_functional_state_program",
      "top_functional_state_score",
      "projectils_label",
      "projectils_label_prop",
      "median_projectils_confidence",
      "top_previous_tcell_subtype",
      "top_previous_tcell_subtype_prop",
      "singler_immgen_label",
      "singler_mousernaseq_label"
    ))
  )

write.csv(
  compartment_cluster_annotation_summary,
  file.path(results_dir, "tcell_cd4_cd8_cluster_annotations.csv"),
  row.names = FALSE
)
```

## Save trajectory input objects
This chunk saves the prepared CD4-like and CD8-like objects used by script 13 for Slingshot trajectory analysis.

```{r save-trajectory-inputs}
trajectory_input_files <- tibble::tibble(
  compartment = c("CD4-like", "CD8-like"),
  rds_file = file.path(
    results_dir,
    c(
      "tcell_cd4_like_trajectory_input.rds",
      "tcell_cd8_like_trajectory_input.rds"
    )
  )
)

saveRDS(
  compartment_objects[["CD4-like"]],
  trajectory_input_files$rds_file[trajectory_input_files$compartment == "CD4-like"]
)

saveRDS(
  compartment_objects[["CD8-like"]],
  trajectory_input_files$rds_file[trajectory_input_files$compartment == "CD8-like"]
)

trajectory_input_files
```

## Session info
```{r session-info}
sessionInfo()
```
