---
title: "12-tcell-cd4-cd8-trajectory"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

This module extends the T-cell trajectory analysis by first separating CD4-like and CD8-like cells using marker expression and ScGate support. Clean CD4-like and CD8-like objects are then created, clustered, and annotated separately. Slingshot is then run on condition-specific CD4-like and CD8-like subsets.

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(purrr)
  library(Seurat)
  library(SingleCellExperiment)
  library(SeuratExtend)
  library(ggplot2)
  library(patchwork)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "tcell")
plot_dir <- file.path(analysis_dir, "plots", "tcell")

source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "trajectory_helpers.R"))
source(file.path(analysis_dir, "util", "tcell_cd4_cd8_trajectory_helpers.R"))

set.seed(1234)
```

## Parameters
This chunk records the input file, marker thresholds, Slingshot thresholds, and marker programs used throughout the analysis.

CD4 and CD8 marker scores are used to identify CD4-like and CD8-like cells. The naive/memory score is used to choose the Slingshot root cluster. Functional-state marker programs are used to annotate CD4-like and CD8-like clusters.

```{r parameters}
input_rds <- "cart_tcell_subtypes.rds"
min_score_delta <- 0.15
min_marker_score <- 0.10
min_cells_for_slingshot <- 100
min_clusters_for_slingshot <- 3
min_cells_per_cluster <- 10

cd4_markers <- c("Cd4")
cd8_markers <- c("Cd8a", "Cd8b1")
naive_memory_markers <- c("Tcf7", "Lef1", "Sell", "Ccr7", "Il7r")

marker_programs <- list(
  CD4 = cd4_markers,
  CD8 = cd8_markers,
  naive_memory = naive_memory_markers
)

functional_state_programs <- list(
  naive_memory = naive_memory_markers,
  activated = c("Cd69", "Il2ra", "Icos", "Tnfrsf4", "Tnfrsf9"),
  treg = c("Foxp3", "Il2ra", "Ctla4", "Tigit", "Ikzf2"),
  cytotoxic = c("Nkg7", "Gzmb", "Gzma", "Prf1", "Ifng", "Ccl5"),
  exhausted = c("Pdcd1", "Lag3", "Havcr2", "Tox", "Tigit", "Ctla4"),
  proliferating = c("Mki67", "Top2a", "Cdk1", "Ccnb1", "Ccnb2"),
  ifn_response = c("Ifit1", "Ifit3", "Isg15", "Oas1a", "Oas3", "Zbp1"),
  stress_response = c("Hspa1a", "Hspa1b", "Dnajb1", "Jun", "Fos")
)

cat("Input RDS:", input_rds, "\n")
cat("Minimum CD4/CD8 score delta:", min_score_delta, "\n")
cat("Minimum marker score:", min_marker_score, "\n")
cat("Minimum cells for Slingshot:", min_cells_for_slingshot, "\n")
cat("Minimum clusters for Slingshot:", min_clusters_for_slingshot, "\n")
cat("Minimum cells per Slingshot cluster:", min_cells_per_cluster, "\n")
```

## Load T-cell object
This chunk loads the renamed T-cell object and confirms that condition metadata is available for downstream pseudotime summaries.

```{r load-data}
input_path <- dplyr::case_when(
  file.exists(file.path(results_dir, input_rds)) ~ file.path(results_dir, input_rds),
  file.exists(file.path(data_dir, input_rds)) ~ file.path(data_dir, input_rds),
  TRUE ~ NA_character_
)

if (is.na(input_path)) {
  stop("Input RDS was not found in results/tcell or data/v5: ", input_rds)
}

cat("Reading:", input_path, "\n")
tcell_obj <- readRDS(input_path)

stopifnot("condition" %in% colnames(tcell_obj@meta.data))

if ("tcell_subtype" %in% colnames(tcell_obj@meta.data)) {
  table(tcell_obj$tcell_subtype)
  tcell_obj$previous_tcell_subtype <- tcell_obj$tcell_subtype
}

table(tcell_obj$condition)

condition_levels <- if (is.factor(tcell_obj$condition)) {
  levels(droplevels(tcell_obj$condition))
} else {
  sort(unique(as.character(tcell_obj$condition)))
}

cat("Conditions:", paste(condition_levels, collapse = ", "), "\n")
```

## CD4/CD8 marker classification
This chunk classifies cells as CD4-like, CD8-like, or ambiguous using CD4 and CD8 marker-expression scores. The marker-score summary checks whether the split captures clear CD4-like and CD8-like populations while leaving uncertain cells as ambiguous.

```{r marker-classification, fig.height=7, fig.width=10}
tcell_obj <- classify_cd4_cd8(tcell_obj)

marker_score_summary <- tcell_obj@meta.data %>%
  dplyr::group_by(condition, cd4_cd8_marker_group) %>%
  dplyr::summarise(
    n = dplyr::n(),
    median_cd4_score = median(cd4_score, na.rm = TRUE),
    median_cd8_score = median(cd8_score, na.rm = TRUE),
    median_naive_memory_score = median(naive_memory_score, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(
  marker_score_summary,
  file.path(results_dir, "tcell_cd4_cd8_marker_score_summary.csv"),
  row.names = FALSE
)

p_marker_feature <- FeaturePlot(
  tcell_obj,
  features = available_features(tcell_obj, c(cd4_markers, cd8_markers, naive_memory_markers)),
  reduction = "umap",
  ncol = 3
)

p_marker_group <- DimPlot(
  tcell_obj,
  reduction = "umap",
  group.by = "cd4_cd8_marker_group"
) +
  ggtitle("CD4/CD8 marker-based groups")

p_marker_scatter <- ggplot(
  tcell_obj@meta.data,
  aes(x = cd4_score, y = cd8_score, color = cd4_cd8_marker_group)
) +
  geom_point(size = 0.25, alpha = 0.45) +
  geom_abline(slope = 1, intercept = min_score_delta, linetype = "dashed", color = "grey35") +
  geom_abline(slope = 1, intercept = -min_score_delta, linetype = "dashed", color = "grey35") +
  theme_Publication() +
  labs(
    title = "CD4/CD8 marker-score split",
    x = "CD4 marker score",
    y = "CD8 marker score",
    color = "Marker group"
  )

p_marker_feature
p_marker_group
p_marker_scatter

ggsave(
  file.path(plot_dir, "tcell_cd4_cd8_marker_expression_and_groups.pdf"),
  plot = p_marker_feature / (p_marker_group | p_marker_scatter),
  width = 12,
  height = 14,
  device = cairo_pdf
)
```

## ScGate CD4/CD8 support
This chunk runs generic mouse ScGate CD4T/CD8T models on the marker-classified T-cell object. ScGate is used here as supporting evidence for CD4/CD8 identity before creating the clean CD4-like and CD8-like trajectory inputs.

```{r scgate-cd4-cd8-support, fig.height=5, fig.width=7}
tcell_obj <- run_scgate_cd4_cd8(tcell_obj)

scgate_marker_counts <- tcell_obj@meta.data %>%
  dplyr::count(
    condition,
    cd4_cd8_marker_group,
    scgate_cd4_cd8_group,
    name = "n"
  ) %>%
  dplyr::arrange(condition, cd4_cd8_marker_group, scgate_cd4_cd8_group)

write.csv(
  scgate_marker_counts,
  file.path(results_dir, "tcell_cd4_cd8_marker_scgate_concordance.csv"),
  row.names = FALSE
)
```

## Prepare clean CD4/CD8 cells
This chunk keeps cells with concordant marker-based and ScGate-supported CD4/CD8 identity and creates clean CD4-like and CD8-like Seurat objects for separate downstream analysis.

```{r prepare-clean-cd4-cd8-cells}
clean_obj <- subset(
  tcell_obj,
  subset = cd4_cd8_marker_group == scgate_cd4_cd8_group &
    cd4_cd8_marker_group %in% c("CD4-like", "CD8-like")
)

clean_obj$cd4_cd8_marker_group <- droplevels(clean_obj$cd4_cd8_marker_group)

clean_cd4_obj <- subset(clean_obj, subset = cd4_cd8_marker_group == "CD4-like")
clean_cd8_obj <- subset(clean_obj, subset = cd4_cd8_marker_group == "CD8-like")
```

## Cluster and annotate CD4-like and CD8-like cells separately
This chunk clusters the clean CD4-like and CD8-like objects once, annotates the new clusters using functional-state marker programs, previous T-cell subtype labels, and SingleR with both mouse references, and writes a compact cluster annotation table.

```{r cluster-cd4-cd8, fig.height=6, fig.width=10}
compartment_objects <- list()
compartment_cluster_annotations <- list()
for (compartment in c("CD4-like", "CD8-like")) {
  obj_compartment <- if (compartment == "CD4-like") clean_cd4_obj else clean_cd8_obj
  
  obj_compartment <- cluster_subset(
    obj_compartment,
    safe_label(tolower(compartment))
  )
  
  annotated_compartment <- annotate_clustered_object(
    obj_compartment,
    compartment = compartment
  )
  
  obj_compartment <- annotated_compartment$obj
  compartment_objects[[compartment]] <- obj_compartment
  compartment_cluster_annotations[[compartment]] <- annotated_compartment$cluster_annotation
  
  p <- DimPlot(
    obj_compartment,
    reduction = "umap",
    group.by = "seurat_clusters",
    label = TRUE,
    repel = TRUE
  ) +
    ggtitle(paste(compartment, "clustered UMAP"))
  
  print(p)
  
  ggsave(
    file.path(
      plot_dir,
      paste0("tcell_", safe_label(tolower(compartment)), "_clustered_umap.pdf")
    ),
    plot = p,
    width = 7,
    height = 6,
    device = cairo_pdf
  )
  
}

compartment_cluster_annotation_df <- purrr::map_dfr(
  compartment_cluster_annotations,
  ~ .x,
  .id = "analysis"
)

compartment_cluster_annotation_summary <- compartment_cluster_annotation_df %>%
  dplyr::select(
    analysis,
    seurat_clusters,
    n,
    dplyr::any_of(c(
      "functional_state_label",
      "top_functional_state_program",
      "top_functional_state_score",
      "top_previous_tcell_subtype",
      "top_previous_tcell_subtype_prop",
      "singler_immgen_label",
      "singler_mousernaseq_label"
    ))
  )

write.csv(
  compartment_cluster_annotation_summary,
  file.path(results_dir, "tcell_cd4_cd8_cluster_annotations.csv"),
  row.names = FALSE
)

compartment_cluster_annotation_summary
```

## Run condition-specific Slingshot by CD4/CD8 group
This chunk subsets the annotated CD4-like and CD8-like objects by condition and runs Slingshot using the CD4/CD8 Seurat clusters.

```{r slingshot-by-compartment-condition, message=FALSE, warning=FALSE}
slingshot_results <- list()

for (compartment in c("CD4-like", "CD8-like")) {
  compartment_obj <- compartment_objects[[compartment]]
  
  for (condition in condition_levels) {
    subset_cells <- colnames(compartment_obj)[compartment_obj$condition == condition]
    
    if (length(subset_cells) == 0) {
      next
    }
    
    result_name <- paste(compartment, condition, sep = "_")
    slingshot_results[[result_name]] <- run_condition_slingshot(
      subset(compartment_obj, cells = subset_cells),
      compartment = compartment,
      condition = condition
    )
  }
}

slingshot_cluster_counts <- purrr::map_dfr(slingshot_results, "cluster_counts", .id = "analysis") %>%
  dplyr::select(
    dplyr::any_of(c(
      "analysis",
      "seurat_clusters",
      "n",
      "retained_for_slingshot"
    )
  ))

slingshot_cluster_counts

slingshot_summaries <- purrr::map_dfr(
  slingshot_results,
  function(x) {
    if (is.null(x$sling_summary)) {
      return(NULL)
    }
    x$sling_summary
  },
  .id = "analysis"
)

slingshot_summaries <- slingshot_summaries %>%
  dplyr::select(
    dplyr::any_of(c(
      "analysis",
      "seurat_clusters",
      "n",
      "lineage",
      "median_pseudotime",
      "mean_pseudotime",
      "functional_state_label",
      "singler_immgen_label",
      "singler_mousernaseq_label",
      "top_previous_tcell_subtype",
      "seurat_cluster_top_previous_tcell_subtype"
    ))
  )

write.csv(
  slingshot_summaries,
  file.path(results_dir, "tcell_cd4_cd8_condition_slingshot_pseudotime_summary.csv"),
  row.names = FALSE
)
```

## Session info
```{r session-info}
sessionInfo()
```
