---
title: "13-tcell-cd4-cd8-slingshot"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

This module loads the CD4-like and CD8-like trajectory input objects saved by script 12 and runs Slingshot on condition-specific subsets and all conditions together. PDFs are written under `plots/tcell/{CD4,CD8}-trajectory`; CSV and RDS files are written under `results/tcell/{CD4,CD8}-trajectory`.

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(purrr)
  library(Seurat)
  library(SingleCellExperiment)
  library(SeuratExtend)
  library(ggplot2)
  library(patchwork)
})

root_dir <- find_root(has_dir(".git"))
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
plot_base_dir <- file.path(analysis_dir, "plots", "tcell")
results_base_dir <- file.path(analysis_dir, "results", "tcell")

source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "trajectory_helpers.R"))
source(file.path(analysis_dir, "util", "tcell_cd4_cd8_trajectory_helpers.R"))

set.seed(1234)
```

## Parameters
This chunk records the Slingshot thresholds, cache controls, input RDS files, marker genes for gene trends, and AUCell programs used for trajectory summaries. Set `force_rerun_slingshot` to `TRUE` to ignore valid caches and recompute trajectories.

```{r parameters}
cd4_input_rds <- file.path(results_base_dir, "tcell_cd4_like_trajectory_input.rds")
cd8_input_rds <- file.path(results_base_dir, "tcell_cd8_like_trajectory_input.rds")

trajectory_plot_dirs <- list(
  `CD4-like` = file.path(plot_base_dir, "CD4-trajectory"),
  `CD8-like` = file.path(plot_base_dir, "CD8-trajectory")
)
trajectory_results_dirs <- list(
  `CD4-like` = file.path(results_base_dir, "CD4-trajectory"),
  `CD8-like` = file.path(results_base_dir, "CD8-trajectory")
)

purrr::walk(trajectory_plot_dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
purrr::walk(trajectory_results_dirs, dir.create, recursive = TRUE, showWarnings = FALSE)

min_cells_for_slingshot <- 100
min_clusters_for_slingshot <- 3
min_cells_per_cluster <- 10
force_rerun_slingshot <- FALSE
slingshot_cache_version <- 2L

marker_programs <- list(
  CD4 = c("Cd4"),
  CD8 = c("Cd8a", "Cd8b1"),
  naive_memory = c("Tcf7", "Lef1", "Sell", "Ccr7", "Il7r"),
  activated = c("Cd69", "Il2ra", "Icos", "Tnfrsf4", "Tnfrsf9"),
  treg = c("Foxp3", "Il2ra", "Ctla4", "Tigit", "Ikzf2"),
  cytotoxic = c("Nkg7", "Gzmb", "Gzma", "Prf1", "Ifng", "Ccl5"),
  exhausted = c("Pdcd1", "Lag3", "Havcr2", "Tox", "Tigit", "Ctla4"),
  proliferating = c("Mki67", "Top2a", "Cdk1", "Ccnb1", "Ccnb2"),
  ifn_response = c("Ifit1", "Ifit3", "Isg15", "Oas1a", "Oas3", "Zbp1"),
  stress_response = c("Hspa1a", "Hspa1b", "Dnajb1", "Jun", "Fos")
)

slingshot_gene_trend_genes <- unique(c(
  "Pdcd1", "Tox", "Lag3", "Havcr2", "Ctla4",
  "Tcf7", "Lef1", "Ccr7", "Il7r",
  "Mki67", "Top2a",
  "Gzmb", "Prf1", "Ifng", "Ccl5"
))

aucell_programs <- marker_programs[c(
  "exhausted",
  "proliferating",
  "naive_memory",
  "cytotoxic"
)]

cat("CD4 input RDS:", cd4_input_rds, "\n")
cat("CD8 input RDS:", cd8_input_rds, "\n")
cat("CD4 trajectory plot dir:", trajectory_plot_dirs[["CD4-like"]], "\n")
cat("CD4 trajectory results dir:", trajectory_results_dirs[["CD4-like"]], "\n")
cat("CD8 trajectory plot dir:", trajectory_plot_dirs[["CD8-like"]], "\n")
cat("CD8 trajectory results dir:", trajectory_results_dirs[["CD8-like"]], "\n")
cat("Minimum cells for Slingshot:", min_cells_for_slingshot, "\n")
cat("Minimum clusters for Slingshot:", min_clusters_for_slingshot, "\n")
cat("Minimum cells per Slingshot cluster:", min_cells_per_cluster, "\n")
cat("Force Slingshot rerun:", force_rerun_slingshot, "\n")
cat("Slingshot cache version:", slingshot_cache_version, "\n")
```

## Load trajectory input objects
This chunk loads the prepared CD4-like and CD8-like objects from script 12 and records checksums used to invalidate stale Slingshot caches.

```{r load-trajectory-inputs}
if (!file.exists(cd4_input_rds)) {
  stop("CD4 trajectory input RDS was not found: ", cd4_input_rds)
}

if (!file.exists(cd8_input_rds)) {
  stop("CD8 trajectory input RDS was not found: ", cd8_input_rds)
}

trajectory_input_md5 <- c(
  `CD4-like` = unname(tools::md5sum(cd4_input_rds)),
  `CD8-like` = unname(tools::md5sum(cd8_input_rds))
)

compartment_objects <- list(
  `CD4-like` = readRDS(cd4_input_rds),
  `CD8-like` = readRDS(cd8_input_rds)
)

condition_levels <- compartment_objects %>%
  purrr::map(~ {
    condition <- .x$condition
    if (is.factor(condition)) {
      levels(droplevels(condition))
    } else {
      unique(as.character(condition))
    }
  }) %>%
  unlist(use.names = FALSE) %>%
  unique() %>%
  sort()

cat("Conditions:", paste(condition_levels, collapse = ", "), "\n")
```

## Helper for one compartment
This helper loads a valid cached Slingshot result when available or runs and caches Slingshot otherwise. Figures and tables are regenerated from the loaded or newly computed result.

```{r slingshot-compartment-helper}
run_compartment_trajectories <- function(
  compartment,
  compartment_obj,
  output_plot_dir,
  output_results_dir,
  input_rds_md5
) {
  compartment_label <- safe_label(compartment)
  subtype_colors <- make_tcell_subtype_palette(
    compartment_obj$seurat_cluster_projectils_label
  )
  cache_context <- list(
    analysis_cache_version = slingshot_cache_version,
    input_rds_md5 = input_rds_md5
  )
  slingshot_results <- list()

  for (condition in condition_levels) {
    subset_cells <- colnames(compartment_obj)[compartment_obj$condition == condition]

    if (length(subset_cells) == 0) {
      next
    }

    result_name <- paste(compartment, condition, sep = "_")
    condition_cache_file <- file.path(
      output_results_dir,
      paste0("tcell_", safe_label(result_name), "_slingshot_result.rds")
    )
    slingshot_results[[result_name]] <- run_condition_slingshot(
      subset(compartment_obj, cells = subset_cells),
      compartment = compartment,
      condition = condition,
      gene_trend_genes = slingshot_gene_trend_genes,
      aucell_programs = aucell_programs,
      plot_aucell_pseudotime = TRUE,
      subtype_colors = subtype_colors,
      cache_file = condition_cache_file,
      cache_context = cache_context,
      force_rerun_slingshot = force_rerun_slingshot,
      output_plot_dir = output_plot_dir,
      output_results_dir = output_results_dir
    )
  }

  slingshot_qc <- purrr::imap_dfr(
    slingshot_results,
    function(x, analysis_name) {
      x$cluster_counts %>%
        dplyr::mutate(analysis = analysis_name) %>%
        dplyr::left_join(
          x$eligibility,
          by = c("compartment", "condition")
        )
    }
  ) %>%
    dplyr::select(
      dplyr::any_of(c(
        "analysis",
        "seurat_clusters",
        "n",
        "retained_for_slingshot",
        "root_cluster",
        "root_cluster_selection",
        "is_root_cluster",
        "root_cluster_projectils_label",
        "skip_reason"
      ))
    )

  write.csv(
    slingshot_qc,
    file.path(output_results_dir, paste0("tcell_", compartment_label, "_condition_slingshot_qc.csv")),
    row.names = FALSE
  )

  slingshot_summaries <- purrr::map_dfr(
    slingshot_results,
    function(x) {
      if (is.null(x$sling_summary)) {
        return(NULL)
      }
      x$sling_summary
    },
    .id = "analysis"
  )

  slingshot_summaries <- slingshot_summaries %>%
    dplyr::select(
      dplyr::any_of(c(
        "analysis",
        "seurat_clusters",
        "n",
        "lineage",
        "median_pseudotime",
        "mean_pseudotime",
        "functional_state_label",
        "projectils_label",
        "projectils_label_prop",
        "singler_immgen_label",
        "singler_mousernaseq_label",
        "top_previous_tcell_subtype",
        "seurat_cluster_top_previous_tcell_subtype",
        "seurat_cluster_projectils_label"
      ))
    )

  write.csv(
    slingshot_summaries,
    file.path(output_results_dir, paste0("tcell_", compartment_label, "_condition_slingshot_pseudotime_summary.csv")),
    row.names = FALSE
  )

  all_conditions_name <- paste(compartment, "all_conditions", sep = "_")
  all_conditions_cache_file <- file.path(
    output_results_dir,
    paste0("tcell_", safe_label(all_conditions_name), "_slingshot_result.rds")
  )
  slingshot_combined_results <- list()
  slingshot_combined_results[[all_conditions_name]] <- run_condition_slingshot(
    compartment_obj,
    compartment = compartment,
    condition = "all_conditions",
    gene_trend_genes = slingshot_gene_trend_genes,
    aucell_programs = aucell_programs,
    plot_aucell_pseudotime = TRUE,
    subtype_colors = subtype_colors,
    cache_file = all_conditions_cache_file,
    cache_context = cache_context,
    force_rerun_slingshot = force_rerun_slingshot,
    output_plot_dir = output_plot_dir,
    output_results_dir = output_results_dir
  )

  slingshot_combined_qc <- purrr::imap_dfr(
    slingshot_combined_results,
    function(x, analysis_name) {
      x$cluster_counts %>%
        dplyr::mutate(analysis = analysis_name) %>%
        dplyr::left_join(
          x$eligibility,
          by = c("compartment", "condition")
        )
    }
  ) %>%
    dplyr::select(
      dplyr::any_of(c(
        "analysis",
        "seurat_clusters",
        "n",
        "retained_for_slingshot",
        "root_cluster",
        "root_cluster_selection",
        "is_root_cluster",
        "root_cluster_projectils_label",
        "skip_reason"
      ))
    )

  write.csv(
    slingshot_combined_qc,
    file.path(output_results_dir, paste0("tcell_", compartment_label, "_all_conditions_slingshot_qc.csv")),
    row.names = FALSE
  )

  slingshot_combined_summaries <- purrr::map_dfr(
    slingshot_combined_results,
    function(x) {
      if (is.null(x$sling_summary)) {
        return(NULL)
      }
      x$sling_summary
    },
    .id = "analysis"
  )

  slingshot_combined_summaries <- slingshot_combined_summaries %>%
    dplyr::select(
      dplyr::any_of(c(
        "analysis",
        "seurat_clusters",
        "n",
        "lineage",
        "median_pseudotime",
        "mean_pseudotime",
        "functional_state_label",
        "projectils_label",
        "projectils_label_prop",
        "singler_immgen_label",
        "singler_mousernaseq_label",
        "top_previous_tcell_subtype",
        "seurat_cluster_top_previous_tcell_subtype",
        "seurat_cluster_projectils_label"
      ))
    )

  write.csv(
    slingshot_combined_summaries,
    file.path(output_results_dir, paste0("tcell_", compartment_label, "_all_conditions_slingshot_pseudotime_summary.csv")),
    row.names = FALSE
  )

  list(
    condition_specific = slingshot_results,
    all_conditions = slingshot_combined_results
  )
}
```

## Run Slingshot trajectories
This chunk runs the cache-aware compartment-specific trajectory workflow, writing PDFs to the compartment plot directory and CSV/RDS files to the matching results directory.

```{r run-slingshot-trajectories, message=FALSE, warning=FALSE}
trajectory_results <- purrr::imap(
  compartment_objects,
  function(compartment_obj, compartment) {
    run_compartment_trajectories(
      compartment = compartment,
      compartment_obj = compartment_obj,
      output_plot_dir = trajectory_plot_dirs[[compartment]],
      output_results_dir = trajectory_results_dirs[[compartment]],
      input_rds_md5 = trajectory_input_md5[[compartment]]
    )
  }
)
```

## Session info
```{r session-info}
sessionInfo()
```
