---
title: "14-kdm6b-expression-subcluster-definition"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

This analysis addresses:

1. compare pseudobulk `Kdm6b` expression across treatments in all cells, T cells, and myeloid cells
2. visualize `Kdm6b` in the myeloid compartment
3. define Kdm6b-high and Kdm6b-low myeloid subclusters.

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(Seurat)
  library(ggplot2)
  library(patchwork)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "myeloid", "kdm6b")
plot_dir <- file.path(analysis_dir, "plots", "myeloid", "kdm6b")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "kdm6b_expression_helpers.R"))

set.seed(1234)
```

## Analysis parameters

```{r parameters}
gene <- "Kdm6b"
compartment_levels <- c("All cells", "T cells", "Myeloid cells")

input_files <- list(
  all_cells = file.path(data_dir, "cart_annotated.rds"),
  t_cells = file.path(data_dir, "cart_tcell_subtypes.rds"),
  myeloid = file.path(data_dir, "cart_myeloid_subtypes.rds")
)
missing_input_files <- unlist(input_files)[!file.exists(unlist(input_files))]
if (length(missing_input_files) > 0) {
  stop("Missing input file(s): ", paste(missing_input_files, collapse = ", "))
}

condition_levels <- c(
  "41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor"
)

condition_colors <- c(
  "41BB-L" = "#1B9E77",
  "B7H3" = "#7570B3",
  "CD28-41BB" = "#E7298A",
  "CD8-41BB" = "#D95F02",
  "STOP" = "#BDBDBD",
  "tumor" = "#66A61E"
)

# Minimum cells required for a sample/subcluster expression profile.
min_cells_per_profile <- 20L

# Assign the most and least expressed quartiles to the high and low groups.
tail_fraction <- 0.25
```

## 1. Compare pseudobulk Kdm6b expression across treatment groups

Raw RNA counts are aggregated within each biological sample. DESeq2 size
factors are estimated from the full pseudobulk transcriptome separately for
each compartment. The plots show individual samples; no pooled cell-level
test is used. Log2-transformed normalized counts are shown to compare treatment
conditions within each compartment. Because normalization is performed
separately, the y-axis magnitudes should not be compared across compartments.

### All cells

```{r pseudobulk-all-cells, message=FALSE, warning=FALSE}
all_obj <- readRDS(input_files$all_cells)

all_pseudobulk <- aggregate_kdm6b_by_sample(
  obj = all_obj,
  compartment = "All cells",
  condition_levels = condition_levels,
  gene = gene
)

rm(all_obj)
invisible(gc())
```

### T cells

```{r pseudobulk-t-cells, message=FALSE, warning=FALSE}
tcell_obj <- readRDS(input_files$t_cells)

tcell_pseudobulk <- aggregate_kdm6b_by_sample(
  obj = tcell_obj,
  compartment = "T cells",
  condition_levels = condition_levels,
  gene = gene
)

rm(tcell_obj)
invisible(gc())
```

### Myeloid cells

```{r pseudobulk-myeloid-cells, message=FALSE, warning=FALSE}
myeloid_obj <- readRDS(input_files$myeloid)
observed_seurat_clusters <- unique(as.character(myeloid_obj$seurat_clusters))
if (!setequal(observed_seurat_clusters, as.character(0:15))) {
  stop("Expected Seurat clusters 0-15 in the myeloid object.")
}

myeloid_obj$seurat_clusters <- factor(
  as.character(myeloid_obj$seurat_clusters),
  levels = as.character(0:15)
)

myeloid_pseudobulk <- aggregate_kdm6b_by_sample(
  obj = myeloid_obj,
  compartment = "Myeloid cells",
  condition_levels = condition_levels,
  gene = gene
)
```

### Pseudobulk tables and plots

```{r pseudobulk-summary-tables}
pseudobulk_sample_table <- dplyr::bind_rows(
  all_pseudobulk,
  tcell_pseudobulk,
  myeloid_pseudobulk
) %>%
  dplyr::mutate(
    compartment = factor(compartment, levels = compartment_levels),
    condition = factor(condition, levels = condition_levels),
    log2_kdm6b_deseq2_normalized_count = log2(
      kdm6b_deseq2_normalized_count + 1
    )
  )

readr::write_tsv(
  pseudobulk_sample_table,
  file.path(results_dir, "kdm6b_pseudobulk_by_sample.tsv")
)
```

```{r plot-pseudobulk-expression, fig.width=9, fig.height=10}
p_pseudobulk <- ggplot(
  pseudobulk_sample_table,
  aes(
    x = condition,
    y = log2_kdm6b_deseq2_normalized_count,
    color = condition,
    shape = sample_number
  )
) +
  geom_point(
    position = position_jitter(width = 0.08, height = 0),
    size = 3,
    alpha = 0.9
  ) +
  facet_wrap(~ compartment, scales = "free_y", ncol = 1) +
  scale_color_manual(values = condition_colors, drop = FALSE) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    strip.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    title = "Pseudobulk Kdm6b expression",
    x = NULL,
    y = "log2(DESeq2-normalized Kdm6b pseudobulk count + 1)",
    color = "Condition",
    shape = "Biological sample"
  )

p_pseudobulk

ggsave(
  filename = file.path(
    plot_dir,
    "kdm6b_pseudobulk_expression_by_compartment.pdf"
  ),
  plot = p_pseudobulk,
  width = 9,
  height = 10,
  device = cairo_pdf
)
```

## 2. Visualize Kdm6b expression within the myeloid compartment

The integrated reduction is used only for visualization. `Kdm6b` expression is
drawn from the normalized RNA assay.

```{r myeloid-kdm6b-featureplot, fig.width=9, fig.height=7}
p_kdm6b_feature <- FeaturePlot(
  myeloid_obj,
  features = gene,
  reduction = "umap",
  min.cutoff = "q05",
  max.cutoff = "q95",
  order = TRUE,
  raster = TRUE
) +
  ggtitle("Kdm6b expression in myeloid cells")

p_kdm6b_feature

ggsave(
  filename = file.path(plot_dir, "kdm6b_myeloid_featureplot.pdf"),
  plot = p_kdm6b_feature,
  width = 9,
  height = 7,
  device = cairo_pdf
)
```

```{r myeloid-kdm6b-violin, fig.width=12, fig.height=6}
p_kdm6b_violin <- VlnPlot(
  myeloid_obj,
  features = gene,
  group.by = "seurat_clusters",
  assay = "RNA",
  pt.size = 0
) &
  theme_Publication() &
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) &
  labs(
    title = "Kdm6b expression by myeloid subcluster",
    x = "Seurat cluster",
    y = "Log-normalized RNA expression"
  )

p_kdm6b_violin

ggsave(
  filename = file.path(plot_dir, "kdm6b_myeloid_violin_by_subcluster.pdf"),
  plot = p_kdm6b_violin,
  width = 12,
  height = 6,
  device = cairo_pdf
)
```

## 3. Define Kdm6b-high and Kdm6b-low myeloid subclusters

### Sample-level subcluster ranking

For each qualifying sample/subcluster profile, calculate mean log-normalized
Kdm6b expression. Rank subclusters by the median across samples, using detection
fraction only to break ties.

```{r aggregate-rank-subclusters, message=FALSE, warning=FALSE}
kdm6b_sample_subcluster <- aggregate_kdm6b_by_sample_cluster(
  obj = myeloid_obj,
  condition_levels = condition_levels,
  gene = gene
)

kdm6b_subcluster_ranking <- rank_kdm6b_subclusters(
  sample_cluster_table = kdm6b_sample_subcluster,
  min_cells_per_profile = min_cells_per_profile,
  tail_fraction = tail_fraction
)

readr::write_tsv(
  kdm6b_subcluster_ranking,
  file.path(results_dir, "kdm6b_subcluster_ranking_and_groups.tsv")
)
```

```{r plot-subcluster-ranking, fig.width=11, fig.height=8}
group_colors <- c(
  "Kdm6b-high" = "#B2182B",
  "Intermediate" = "#BDBDBD",
  "Kdm6b-low" = "#2166AC"
)

ranking_plot_data <- kdm6b_subcluster_ranking %>%
  dplyr::mutate(
    kdm6b_group = factor(
      kdm6b_group,
      levels = c(
        "Kdm6b-high",
        "Intermediate",
        "Kdm6b-low"
      )
    )
  ) %>%
  dplyr::arrange(
    kdm6b_group,
    kdm6b_rank,
    as.integer(subcluster)
  ) %>%
  dplyr::mutate(
    seurat_cluster = factor(
      subcluster,
      levels = rev(unique(as.character(subcluster)))
    )
  )

p_ranking <- ggplot(
  ranking_plot_data,
  aes(
    x = seurat_cluster,
    y = median_log_expression,
    color = kdm6b_group
  )
) +
  geom_errorbar(
    aes(
      ymin = q25_log_expression,
      ymax = q75_log_expression
    ),
    width = 0.25,
    linewidth = 0.5
  ) +
  geom_point(size = 3) +
  scale_x_discrete(drop = FALSE) +
  coord_flip() +
  scale_color_manual(values = group_colors, drop = FALSE) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    legend.position = "right"
  ) +
  labs(
    title = "Sample-level Kdm6b ranking of myeloid subclusters",
    subtitle = "Points are medians; error bars are sample-level interquartile ranges",
    x = "Seurat cluster",
    y = "Median sample-level mean log-normalized Kdm6b expression",
    color = "Kdm6b group"
  )

p_ranking

ggsave(
  filename = file.path(plot_dir, "kdm6b_subcluster_ranking.pdf"),
  plot = p_ranking,
  width = 11,
  height = 8,
  device = cairo_pdf,
  bg = "white"
)
```

## Session info
```{r session-info}
sessionInfo()
```
