---
title: "15-kdm6b-subcluster-abundance"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
  ranking_tsv: "myeloid_kdm6b_subcluster_ranking_and_groups.tsv"
  subtype_label: "myeloid"
  condition_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")
---

This analysis quantifies the abundance of numeric subclusters in the selected compartment and
tests their differential abundance across treatment conditions. The sccomp model
uses biological samples as replicates and 41BB-L as the reference condition.

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(ggplot2)
  library(Seurat)
  library(sccomp)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
analysis_label <- params$subtype_label
results_dir <- file.path(analysis_dir, "results", analysis_label, "kdm6b")
plot_dir <- file.path(analysis_dir, "plots", analysis_label, "kdm6b")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "sccomp_helpers.R"))
```

## Analysis parameters

```{r parameters}
input_file <- file.path(data_dir, params$input_rds)
ranking_file <- file.path(results_dir, params$ranking_tsv)
condition_levels <- params$condition_levels

missing_input_files <- c(input_file, ranking_file)[
  !file.exists(c(input_file, ranking_file))
]
if (length(missing_input_files) > 0) {
  stop("Missing input file(s): ", paste(missing_input_files, collapse = ", "))
}
```

## Load data

```{r load-data}
obj <- readRDS(input_file)
kdm6b_ranking <- readr::read_tsv(ranking_file, show_col_types = FALSE) %>%
  dplyr::mutate(subcluster = as.character(subcluster))

required_metadata <- c("sample", "condition", "seurat_clusters")
ranking_subtype_col <- paste0(analysis_label, "_subtype")
kdm6b_ranking <- kdm6b_ranking %>%
  dplyr::rename(subtype = dplyr::all_of(ranking_subtype_col))

if (analysis_label == "tcell") {
  kdm6b_ranking <- kdm6b_ranking %>%
    dplyr::mutate(
      subtype = dplyr::if_else(
        subcluster == "4",
        "Cycling / proliferating T cells",
        subtype
      )
    )
}

required_ranking_columns <- c(
  "subcluster", "subtype", "kdm6b_rank", "kdm6b_group"
)

stopifnot(all(required_metadata %in% colnames(obj@meta.data)))
stopifnot(all(required_ranking_columns %in% colnames(kdm6b_ranking)))

observed_conditions <- unique(as.character(obj$condition))
if (!setequal(observed_conditions, condition_levels)) {
  stop("The observed treatment conditions do not match condition_levels.")
}

observed_subclusters <- unique(as.character(obj$seurat_clusters))
if (!setequal(observed_subclusters, kdm6b_ranking$subcluster)) {
  stop("The ranking table and myeloid object contain different subclusters.")
}

subcluster_levels <- as.character(
  sort(as.integer(observed_subclusters))
)
obj$condition <- factor(
  as.character(obj$condition),
  levels = condition_levels
)
obj$seurat_clusters <- factor(
  as.character(obj$seurat_clusters),
  levels = subcluster_levels
)

sample_condition_check <- obj@meta.data %>%
  dplyr::distinct(sample, condition) %>%
  dplyr::count(sample, name = "n_conditions")
if (any(sample_condition_check$n_conditions != 1)) {
  stop("Each biological sample must map to exactly one condition.")
}

table(obj$seurat_clusters, obj$condition)
```

## Subcluster proportions

```{r plot-subcluster-proportions, fig.width=8, fig.height=5}
subcluster_proportions <- obj@meta.data %>%
  dplyr::count(condition, cell_group = seurat_clusters, name = "n_cells") %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(proportion = n_cells / sum(n_cells)) %>%
  dplyr::ungroup()

p_subcluster_proportions <- ggplot(
  subcluster_proportions,
  aes(x = condition, y = proportion, fill = cell_group)
) +
  geom_col() +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_manual(values = DiscretePalette(n = length(subcluster_levels))) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = "Condition", y = "Proportion", fill = "Subcluster")

p_subcluster_proportions

ggsave(
  file.path(plot_dir, paste0(analysis_label, "_subcluster_proportions_stacked_barplot.png")),
  plot = p_subcluster_proportions,
  width = 8,
  height = 5,
  bg = "white"
)
```

## All subcluster differential abundance

This model tests all 16 numeric subclusters. B7H3, CD28-41BB, CD8-41BB, STOP,
and tumor are each compared with the 41BB-L reference condition.

```{r sccomp-subclusters, message=FALSE, warning=FALSE}
sccomp_subclusters <- run_sccomp_composition(
  obj = obj,
  cell_group_col = "seurat_clusters",
  condition_levels = condition_levels,
  analysis_label = paste0(analysis_label, "_subcluster"),
  output_suffix = "41BBL_baseline",
  reference_label = "41BB-L reference",
  results_dir = results_dir
)

subcluster_sccomp_result <- sccomp_subclusters$result
```

```{r plot-sccomp-subclusters, fig.width=16, fig.height=9}
p_sccomp_subclusters <- subcluster_sccomp_result %>%
  sccomp::plot_1D_intervals(significance_statistic = "FDR")

p_sccomp_subclusters

ggsave(
  file.path(
    plot_dir,
    paste0(analysis_label, "_subcluster_composition_sccomp_41BBL_baseline.pdf")
  ),
  plot = p_sccomp_subclusters,
  width = 16,
  height = 9,
  bg = "white"
)
```

## Kdm6b-high and Kdm6b-low subcluster patterns

This summary uses the individual-subcluster sccomp results. Intermediate
subclusters are excluded from this presentation but remain included in the
fitted compositional model. Effects and intervals are reoriented to report
41BB-L relative to each comparison condition.

```{r summarize-kdm6b-high-low-subclusters}
kdm6b_high_low_sccomp <- subcluster_sccomp_result %>%
  dplyr::filter(parameter != "(Intercept)") %>%
  dplyr::mutate(
    cell_group = as.character(cell_group),
    comparison_condition = sub("^condition", "", parameter)
  ) %>%
  dplyr::left_join(
    kdm6b_ranking %>%
      dplyr::select(
        cell_group = subcluster,
        subtype,
        kdm6b_rank,
        kdm6b_group
      ),
    by = "cell_group"
  ) %>%
  dplyr::filter(kdm6b_group %in% c("Kdm6b-high", "Kdm6b-low")) %>%
  dplyr::mutate(
    comparison = paste("41BB-L vs", comparison_condition),
    sccomp_reference_condition = "41BB-L",
    effect_41bbl_vs_condition = -c_effect,
    lower_41bbl_vs_condition = -c_upper,
    upper_41bbl_vs_condition = -c_lower,
    significant_direction = dplyr::case_when(
      c_FDR >= 0.05 ~ "Not significantly different",
      effect_41bbl_vs_condition > 0 ~ "Enriched in 41BB-L",
      effect_41bbl_vs_condition < 0 ~ paste("Enriched in", comparison_condition),
      TRUE ~ "Not significantly different"
    ),
    status_in_41bbl = dplyr::case_when(
      c_FDR >= 0.05 ~ "Not significantly different",
      effect_41bbl_vs_condition > 0 ~ "Enriched in 41BB-L",
      effect_41bbl_vs_condition < 0 ~ "Depleted in 41BB-L",
      TRUE ~ "Not significantly different"
    )
  ) %>%
  dplyr::select(
    subcluster = cell_group,
    subtype,
    kdm6b_rank,
    kdm6b_group,
    comparison,
    comparison_condition,
    sccomp_reference_condition,
    effect_41bbl_vs_condition,
    lower_41bbl_vs_condition,
    upper_41bbl_vs_condition,
    c_pH0,
    c_FDR,
    significant_direction,
    status_in_41bbl,
    c_rhat,
    c_ess_bulk,
    c_ess_tail
  ) %>%
  dplyr::arrange(
    kdm6b_group,
    kdm6b_rank,
    match(comparison_condition, condition_levels[-1])
  )

readr::write_tsv(
  kdm6b_high_low_sccomp,
  file.path(
    results_dir,
    paste0(analysis_label, "_kdm6b_high_low_subcluster_differential_abundance.tsv")
  )
)

kdm6b_high_low_sccomp
```

```{r plot-kdm6b-high-low-subclusters, fig.width=10, fig.height=7}
subcluster_label_levels <- kdm6b_ranking %>%
  dplyr::filter(kdm6b_group %in% c("Kdm6b-high", "Kdm6b-low")) %>%
  dplyr::arrange(kdm6b_group, kdm6b_rank) %>%
  dplyr::transmute(
    label = paste0("Cluster ", subcluster, ": ", subtype)
  ) %>%
  dplyr::pull(label)

kdm6b_high_low_plot_data <- kdm6b_high_low_sccomp %>%
  dplyr::mutate(
    kdm6b_group = factor(
      kdm6b_group,
      levels = c("Kdm6b-high", "Kdm6b-low")
    ),
    comparison = factor(
      comparison,
      levels = paste("41BB-L vs", condition_levels[-1])
    ),
    subcluster_label = factor(
      paste0("Cluster ", subcluster, ": ", subtype),
      levels = rev(subcluster_label_levels)
    ),
    significance_label = ifelse(c_FDR < 0.05, "*", "")
  )

p_kdm6b_high_low_enrichment <- ggplot(
  kdm6b_high_low_plot_data,
  aes(
    x = comparison,
    y = subcluster_label,
    fill = effect_41bbl_vs_condition
  )
) +
  geom_tile(color = "grey85") +
  geom_text(aes(label = significance_label), size = 5) +
  facet_grid(
    rows = vars(kdm6b_group),
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "sccomp effect\n(41BB-L vs comparison)"
  ) +
  labs(
    x = "Comparison",
    y = NULL,
    caption = paste(
      "* FDR < 0.05. Positive (red): enriched in 41BB-L;",
      "negative (blue): enriched in the comparison condition."
    )
  ) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "white", color = "grey70")
  )

p_kdm6b_high_low_enrichment

ggsave(
  file.path(
    plot_dir,
    paste0(analysis_label, "_kdm6b_high_low_subcluster_enrichment_across_conditions.pdf")
  ),
  plot = p_kdm6b_high_low_enrichment,
  width = 10,
  height = 7,
  device = cairo_pdf,
  bg = "white"
)
```

## Session info

```{r session-info}
sessionInfo()
```
