---
title: "16-kdm6b-transcriptional-comparison"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
  ranking_tsv: "kdm6b_subcluster_ranking_and_groups.tsv"
  subtype_label: "myeloid"
  min_total_count: 10
  de_fdr: 0.05
  heatmap_genes_per_direction: 25
  gsea_min_size: 15
  gsea_max_size: 500
  gsea_pathways_per_direction: 10
---

This analysis compares the transcriptional profiles of Kdm6b-high and
Kdm6b-low subclusters within the selected compartment. Intermediate subclusters are excluded from the
comparison.

Raw RNA counts are aggregated into one Kdm6b-high and one Kdm6b-low pseudobulk
profile per biological sample. The paired DESeq2 design, `~ sample +
kdm6b_group`, estimates the high-versus-low difference within samples and then
combines that evidence across all samples. Treatment is not included as a tested
effect.

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(SeuratObject)
  library(Matrix)
  library(DESeq2)
  library(msigdbr)
  library(fgsea)
  library(ComplexHeatmap)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
analysis_label <- params$subtype_label
compartment_label <- dplyr::case_when(
  analysis_label == "tcell" ~ "T-cell",
  analysis_label == "myeloid" ~ "Myeloid",
  TRUE ~ analysis_label
)
results_dir <- file.path(analysis_dir, "results", analysis_label, "kdm6b")
plot_dir <- file.path(analysis_dir, "plots", analysis_label, "kdm6b")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(root_dir, "figures", "theme.R"))
set.seed(1234)
```

## Analysis parameters

```{r parameters}
input_file <- file.path(data_dir, params$input_rds)
ranking_file <- file.path(results_dir, params$ranking_tsv)
min_total_count <- params$min_total_count
de_fdr <- params$de_fdr
heatmap_genes_per_direction <- params$heatmap_genes_per_direction
gsea_min_size <- params$gsea_min_size
gsea_max_size <- params$gsea_max_size
gsea_pathways_per_direction <- params$gsea_pathways_per_direction

kdm6b_group_levels <- c("Kdm6b-low", "Kdm6b-high")
condition_levels <- c(
  "41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor"
)
kdm6b_group_colors <- c(
  "Kdm6b-low" = "#2166AC",
  "Kdm6b-high" = "#B2182B"
)
missing_input_files <- c(input_file, ranking_file)[
  !file.exists(c(input_file, ranking_file))
]
if (length(missing_input_files) > 0) {
  stop("Missing input file(s): ", paste(missing_input_files, collapse = ", "))
}
```

## Load data and map Kdm6b groups

The previously defined high and low assignments are imported from the
subcluster-ranking table and are not recalculated here.

```{r load-and-map-groups}
myeloid_obj <- readRDS(input_file)
kdm6b_ranking <- readr::read_tsv(ranking_file, show_col_types = FALSE) %>%
  dplyr::mutate(subcluster = as.character(subcluster))

required_metadata <- c("sample", "condition", "seurat_clusters")
required_ranking_columns <- c("subcluster", "kdm6b_group")
stopifnot(all(required_metadata %in% colnames(myeloid_obj@meta.data)))
stopifnot(all(required_ranking_columns %in% colnames(kdm6b_ranking)))

if (anyDuplicated(kdm6b_ranking$subcluster)) {
  stop("Each numeric subcluster must have exactly one Kdm6b group assignment.")
}

observed_subclusters <- unique(as.character(myeloid_obj$seurat_clusters))
if (!setequal(observed_subclusters, kdm6b_ranking$subcluster)) {
  stop("The ranking table and myeloid object contain different subclusters.")
}

cell_metadata <- myeloid_obj@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::transmute(
    cell,
    sample = as.character(sample),
    condition = as.character(condition),
    subcluster = as.character(seurat_clusters)
  ) %>%
  dplyr::left_join(
    kdm6b_ranking %>% dplyr::select(subcluster, kdm6b_group),
    by = "subcluster"
  )

if (anyNA(cell_metadata$kdm6b_group)) {
  stop("At least one cell could not be assigned to a Kdm6b group.")
}

analysis_metadata <- cell_metadata %>%
  dplyr::filter(kdm6b_group %in% kdm6b_group_levels) %>%
  dplyr::mutate(
    condition = factor(condition, levels = condition_levels),
    kdm6b_group = factor(kdm6b_group, levels = kdm6b_group_levels),
    profile_id = paste(sample, kdm6b_group, sep = "__")
  )

if (anyNA(analysis_metadata$condition)) {
  stop("At least one cell has a condition outside condition_levels.")
}

cat("Cells retained for the high-versus-low comparison:",
    nrow(analysis_metadata), "\n")
cat("Intermediate cells excluded:",
    nrow(cell_metadata) - nrow(analysis_metadata), "\n")
table(analysis_metadata$kdm6b_group, analysis_metadata$condition)
```

## Build paired sample-level pseudobulks

Each sample must contribute both a Kdm6b-high and a Kdm6b-low profile. Raw RNA
counts are summed across all cells belonging to each sample and group.

```{r build-pseudobulk, message=FALSE, warning=FALSE}
counts <- SeuratObject::GetAssayData(
  myeloid_obj,
  assay = "RNA",
  layer = "counts"
)

if (!all(analysis_metadata$cell %in% colnames(counts))) {
  stop("At least one retained cell is absent from the RNA count matrix.")
}

analysis_metadata <- analysis_metadata %>%
  dplyr::arrange(match(cell, colnames(counts)))
analysis_counts <- counts[, analysis_metadata$cell, drop = FALSE]
stopifnot(identical(colnames(analysis_counts), analysis_metadata$cell))

profile_metadata <- analysis_metadata %>%
  dplyr::count(
    profile_id,
    sample,
    condition,
    kdm6b_group,
    name = "n_cells"
  ) %>%
  dplyr::arrange(sample, kdm6b_group)

pair_check <- profile_metadata %>%
  dplyr::count(sample, name = "n_group_profiles")
if (any(pair_check$n_group_profiles != 2)) {
  stop("Every sample must contain one Kdm6b-high and one Kdm6b-low profile.")
}

profile_indicator <- Matrix::sparseMatrix(
  i = seq_len(ncol(analysis_counts)),
  j = match(analysis_metadata$profile_id, profile_metadata$profile_id),
  x = 1,
  dims = c(ncol(analysis_counts), nrow(profile_metadata)),
  dimnames = list(colnames(analysis_counts), profile_metadata$profile_id)
)
pseudobulk_counts <- analysis_counts %*% profile_indicator

profile_metadata <- profile_metadata %>%
  dplyr::mutate(
    raw_library_size = as.numeric(Matrix::colSums(pseudobulk_counts)),
    sample_number = sub("^.*_(D[0-9]+)$", "\\1", sample)
  )

profile_metadata
```

## Paired differential-expression analysis

The sample term blocks differences among biological samples. The model tests one
common Kdm6b-high versus Kdm6b-low effect across the 12 paired samples. Treatment
is retained only as provenance in the profile table and is not included in the
design, contrast, or plots.

Genes with fewer than `r min_total_count` total counts across the 24 profiles are
excluded before model fitting. The output table contains every gene tested by
DESeq2, rather than only statistically significant genes.

```{r paired-deseq2, message=FALSE, warning=FALSE}
coldata <- profile_metadata %>%
  dplyr::mutate(
    sample_block = factor(match(sample, unique(sample))),
    kdm6b_group_model = factor(
      ifelse(kdm6b_group == "Kdm6b-high", "high", "low"),
      levels = c("low", "high")
    )
  ) %>%
  as.data.frame()
rownames(coldata) <- coldata$profile_id

pseudobulk_counts <- pseudobulk_counts[, rownames(coldata), drop = FALSE]
keep_genes <- Matrix::rowSums(pseudobulk_counts) >= min_total_count
cat("Genes retained for DESeq2:", sum(keep_genes), "of",
    length(keep_genes), "\n")

design_matrix <- stats::model.matrix(
  ~ sample_block + kdm6b_group_model,
  data = coldata
)
if (qr(design_matrix)$rank != ncol(design_matrix)) {
  stop("The paired DESeq2 design matrix is not full rank.")
}

dds <- DESeq2::DESeqDataSetFromMatrix(
  countData = round(as.matrix(pseudobulk_counts[keep_genes, , drop = FALSE])),
  colData = coldata,
  design = ~ sample_block + kdm6b_group_model
)
dds <- DESeq2::DESeq(dds, quiet = TRUE)

de_result <- DESeq2::results(
  dds,
  contrast = c("kdm6b_group_model", "high", "low"),
  alpha = de_fdr
)

de_table <- as.data.frame(de_result) %>%
  tibble::rownames_to_column("gene") %>%
  dplyr::mutate(
    comparison = "Kdm6b-high vs Kdm6b-low",
    direction = dplyr::case_when(
      is.na(log2FoldChange) ~ NA_character_,
      log2FoldChange > 0 ~ "Higher in Kdm6b-high",
      log2FoldChange < 0 ~ "Higher in Kdm6b-low",
      TRUE ~ "No direction"
    ),
    significant_fdr = !is.na(padj) & padj < de_fdr
  ) %>%
  dplyr::arrange(is.na(padj), padj, pvalue)

readr::write_tsv(
  de_table,
  file.path(results_dir, paste0(analysis_label, "_kdm6b_high_vs_low_paired_deseq2.tsv"))
)

summary(de_result)
table(de_table$direction, de_table$significant_fdr, useNA = "ifany")
```

## Hallmark gene-set enrichment analysis

All genes with finite DESeq2 Wald statistics are ranked from higher in
Kdm6b-high to higher in Kdm6b-low. Positive normalized enrichment scores indicate
Hallmark pathways associated with the Kdm6b-high subclusters.

```{r hallmark-gsea, message=FALSE, warning=FALSE}
hallmark_pathways <- msigdbr::msigdbr(
  species = "Mus musculus",
  collection = "H"
) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)

gene_ranks <- de_table %>%
  dplyr::filter(!is.na(stat), is.finite(stat)) %>%
  dplyr::select(gene, stat) %>%
  tibble::deframe() %>%
  sort(decreasing = TRUE)

set.seed(1234)
hallmark_gsea <- fgsea::fgsea(
  pathways = hallmark_pathways,
  stats = gene_ranks,
  minSize = gsea_min_size,
  maxSize = gsea_max_size
) %>%
  dplyr::arrange(padj, pval) %>%
  dplyr::mutate(
    direction = ifelse(
      NES > 0,
      "Enriched in Kdm6b-high",
      "Enriched in Kdm6b-low"
    )
  )

hallmark_gsea_output <- hallmark_gsea %>%
  dplyr::mutate(
    leadingEdge = vapply(
      leadingEdge,
      paste,
      collapse = ";",
      FUN.VALUE = character(1)
    )
  )

readr::write_tsv(
  hallmark_gsea_output,
  file.path(results_dir, paste0(analysis_label, "_kdm6b_high_vs_low_hallmark_gsea.tsv"))
)

hallmark_gsea_output
```

```{r plot-hallmark-gsea, fig.width=10, fig.height=7}
gsea_plot_data <- hallmark_gsea %>%
  dplyr::filter(!is.na(padj), is.finite(padj), is.finite(NES)) %>%
  dplyr::group_by(direction) %>%
  dplyr::slice_min(order_by = padj, n = gsea_pathways_per_direction) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    pathway_label = stringr::str_remove(pathway, "^HALLMARK_"),
    pathway_label = stringr::str_replace_all(pathway_label, "_", " "),
    pathway_label = factor(
      pathway_label,
      levels = rev(unique(pathway_label[order(NES)]))
    ),
    minus_log10_fdr = -log10(pmax(padj, .Machine$double.xmin))
  )

if (nrow(gsea_plot_data) == 0) {
  stop("No finite Hallmark GSEA results are available for plotting.")
}

p_hallmark_gsea <- ggplot(
  gsea_plot_data,
  aes(
    x = NES,
    y = pathway_label,
    color = NES,
    size = minus_log10_fdr
  )
) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.5) +
  geom_point(alpha = 0.9) +
  scale_color_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0
  ) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title = paste(
      "Hallmark GSEA:",
      compartment_label,
      "Kdm6b-high vs Kdm6b-low"
    ),
    subtitle = "Positive NES indicates enrichment in Kdm6b-high subclusters",
    x = "Normalized enrichment score",
    y = NULL,
    color = "NES",
    size = "-log10(FDR)"
  )

p_hallmark_gsea

ggsave(
  file.path(plot_dir, paste0(analysis_label, "_kdm6b_high_vs_low_hallmark_gsea_dotplot.pdf")),
  plot = p_hallmark_gsea,
  width = 10,
  height = 7,
  device = cairo_pdf,
  bg = "white"
)
```

## Heatmap of top differential genes

The heatmap displays variance-stabilized pseudobulk expression. Genes are ranked
separately by direction, prioritizing FDR-significant genes. If fewer than the
requested number pass the FDR threshold, the remaining positions are filled by
the next best-ranked tested genes and identified in the accompanying table.
Columns are ordered with all Kdm6b-high profiles first, followed by all
Kdm6b-low profiles; individual sample labels are hidden.

```{r select-heatmap-genes}
heatmap_gene_table <- de_table %>%
  dplyr::filter(
    direction %in% c("Higher in Kdm6b-high", "Higher in Kdm6b-low"),
    !is.na(pvalue),
    is.finite(pvalue),
    is.finite(log2FoldChange)
  ) %>%
  dplyr::group_by(direction) %>%
  dplyr::arrange(
    dplyr::desc(significant_fdr),
    padj,
    pvalue,
    dplyr::desc(abs(log2FoldChange)),
    .by_group = TRUE
  ) %>%
  dplyr::slice_head(n = heatmap_genes_per_direction) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    heatmap_direction = factor(
      direction,
      levels = c("Higher in Kdm6b-high", "Higher in Kdm6b-low")
    )
  ) %>%
  dplyr::arrange(heatmap_direction, padj, pvalue)

if (nrow(heatmap_gene_table) == 0) {
  stop("No genes are available for the differential-expression heatmap.")
}

readr::write_tsv(
  heatmap_gene_table,
  file.path(results_dir, paste0(analysis_label, "_kdm6b_high_vs_low_heatmap_genes.tsv"))
)

heatmap_gene_table %>%
  dplyr::select(
    gene,
    log2FoldChange,
    padj,
    significant_fdr,
    direction
  )
```

```{r plot-differential-gene-heatmap, fig.width=13, fig.height=10}
vsd <- DESeq2::varianceStabilizingTransformation(dds, blind = FALSE)
column_order <- order(
  match(
    as.character(coldata$kdm6b_group),
    c("Kdm6b-high", "Kdm6b-low")
  ),
  coldata$sample_block
)
coldata_plot <- coldata[column_order, , drop = FALSE]

heatmap_matrix <- SummarizedExperiment::assay(vsd)[
  heatmap_gene_table$gene,
  rownames(coldata_plot),
  drop = FALSE
]

heatmap_z <- t(scale(t(heatmap_matrix)))
heatmap_z[!is.finite(heatmap_z)] <- 0
heatmap_z <- pmax(pmin(heatmap_z, 2.5), -2.5)

column_annotation <- ComplexHeatmap::HeatmapAnnotation(
  Kdm6b_group = factor(
    as.character(coldata_plot$kdm6b_group),
    levels = c("Kdm6b-high", "Kdm6b-low")
  ),
  col = list(
    Kdm6b_group = kdm6b_group_colors
  ),
  annotation_legend_param = list(
    Kdm6b_group = list(direction = "vertical")
  ),
  show_annotation_name = FALSE
)

heatmap_plot <- ComplexHeatmap::Heatmap(
  heatmap_z,
  name = "Row z-score",
  col = circlize::colorRamp2(
    c(-2.5, 0, 2.5),
    c("#2166AC", "white", "#B2182B")
  ),
  top_annotation = column_annotation,
  row_split = heatmap_gene_table$heatmap_direction,
  cluster_columns = FALSE,
  cluster_rows = TRUE,
  cluster_row_slices = FALSE,
  show_row_dend = FALSE,
  show_column_names = FALSE,
  column_title = paste(
    compartment_label,
    "Kdm6b-high vs Kdm6b-low"
  ),
  column_title_gp = grid::gpar(fontsize = 14, fontface = "bold"),
  row_title = NULL,
  border = TRUE,
  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2.5, 0, 2.5),
    labels = c("-2.5", "0", "2.5"),
    direction = "vertical"
  )
)

grDevices::pdf(
  file.path(plot_dir, paste0(analysis_label, "_kdm6b_high_vs_low_pseudobulk_gene_heatmap.pdf")),
  width = 10,
  height = 8
)
ComplexHeatmap::draw(
  heatmap_plot,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE
)
grDevices::dev.off()

heatmap_plot
```

## Session info

```{r session-info}
sessionInfo()
```
