---
title: "17-kdm6b-myeloid-gene-correlations"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
---

This analysis measures the association between `Kdm6b` and selected immune-program
genes across the full myeloid compartment. It does not subset cells by
Kdm6b-high or Kdm6b-low subcluster.

Mean log-normalized expression is calculated for `Kdm6b` and each gene of
interest within each biological sample. Spearman correlation is then calculated
across the 12 sample-level means. Results are reported separately for antigen
presentation, inflammatory activation, and T-cell recruitment programs.

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(SeuratObject)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
results_dir <- file.path(analysis_dir, "results", "myeloid", "kdm6b")
plot_dir <- file.path(analysis_dir, "plots", "myeloid", "kdm6b")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
```

## Genes of interest

```{r analysis-parameters}
input_file <- file.path(data_dir, params$input_rds)
if (!file.exists(input_file)) {
  stop("Missing input file: ", input_file)
}

condition_levels <- c(
  "41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor"
)
program_levels <- c(
  "Antigen presentation",
  "Inflammatory activation",
  "T-cell recruitment"
)

program_genes <- tibble::tribble(
  ~program, ~gene,
  "Antigen presentation", "H2-Ab1",
  "Antigen presentation", "Cd74",
  "Antigen presentation", "H2-Eb1",
  "Antigen presentation", "B2m",
  "Antigen presentation", "Psmb8",
  "Antigen presentation", "Tap1",
  "Antigen presentation", "Tapbp",
  "Inflammatory activation", "Il12b",
  "Inflammatory activation", "Ifngr1",
  "Inflammatory activation", "Stat1",
  "Inflammatory activation", "Irf1",
  "Inflammatory activation", "Irf7",
  "Inflammatory activation", "Cxcl9",
  "Inflammatory activation", "Cxcl10",
  "Inflammatory activation", "Tnf",
  "T-cell recruitment", "Ccl5",
  "T-cell recruitment", "Cxcl9",
  "T-cell recruitment", "Cxcl10",
  "T-cell recruitment", "Xcl1",
  "T-cell recruitment", "Xcr1",
  "T-cell recruitment", "Ccr7",
  "T-cell recruitment", "Ccl19",
  "T-cell recruitment", "Ccl21a"
) %>%
  dplyr::mutate(program = factor(program, levels = program_levels))
```

## Load all myeloid cells

Correlations use sample-level mean log-normalized RNA expression.

```{r load-expression-data, message=FALSE, warning=FALSE}
myeloid_obj <- readRDS(input_file)
required_metadata <- c("sample", "condition")
stopifnot(all(required_metadata %in% colnames(myeloid_obj@meta.data)))

log_expression <- SeuratObject::GetAssayData(
  myeloid_obj,
  assay = "RNA",
  layer = "data"
)
if (!"Kdm6b" %in% rownames(log_expression)) {
  stop("Kdm6b is absent from the RNA assay.")
}

missing_genes <- setdiff(unique(program_genes$gene), rownames(log_expression))
if (length(missing_genes) > 0) {
  message("Genes absent from the RNA assay: ", paste(missing_genes, collapse = ", "))
}
program_genes <- program_genes %>%
  dplyr::filter(gene %in% rownames(log_expression))
target_genes <- unique(program_genes$gene)

cell_metadata <- myeloid_obj@meta.data[colnames(log_expression), , drop = FALSE] %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::transmute(
    cell,
    sample = as.character(sample),
    condition = factor(as.character(condition), levels = condition_levels)
  )

if (anyNA(cell_metadata$condition)) {
  stop("At least one cell has a condition outside condition_levels.")
}
if (!identical(cell_metadata$cell, colnames(log_expression))) {
  stop("Expression columns and metadata rows are not aligned.")
}

correlation_expression <- as.matrix(
  log_expression[c("Kdm6b", target_genes), , drop = FALSE]
)
cat("Myeloid cells included:", nrow(cell_metadata), "\n")
cat("Biological samples included:", dplyr::n_distinct(cell_metadata$sample), "\n")
cat("Target genes detected:", length(target_genes), "\n")
```

## Calculate sample-level mean expression and correlations

```{r correlation-functions}
safe_sample_correlation <- function(x, y) {
  if (length(x) < 3 || length(unique(x)) < 2 || length(unique(y)) < 2) {
    return(NA_real_)
  }
  stats::cor(x, y, method = "spearman", use = "complete.obs")
}

sample_metadata <- cell_metadata %>%
  dplyr::distinct(sample, condition) %>%
  dplyr::arrange(condition, sample)

calculate_sample_means <- function(sample_id) {
  cell_index <- which(cell_metadata$sample == sample_id)
  sample_condition <- as.character(cell_metadata$condition[cell_index][1])
  sample_means <- rowMeans(correlation_expression[, cell_index, drop = FALSE])
  tibble::tibble(
    sample = sample_id,
    condition = sample_condition,
    n_cells = length(cell_index)
  ) %>%
    dplyr::bind_cols(tibble::as_tibble_row(as.list(sample_means)))
}
```

```{r calculate-correlations, message=FALSE, warning=FALSE}
sample_means_wide <- dplyr::bind_rows(
  lapply(sample_metadata$sample, calculate_sample_means)
) %>%
  dplyr::mutate(condition = factor(condition, levels = condition_levels))

sample_mean_output <- sample_means_wide %>%
  dplyr::select(sample, condition, n_cells, Kdm6b, dplyr::all_of(target_genes)) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(target_genes),
    names_to = "gene",
    values_to = "gene_mean_expression"
  ) %>%
  dplyr::rename(kdm6b_mean_expression = Kdm6b) %>%
  dplyr::arrange(gene, condition, sample)

correlation_summary <- sample_mean_output %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    n_samples = sum(
      is.finite(kdm6b_mean_expression) & is.finite(gene_mean_expression)
    ),
    spearman_rho = if (n_samples >= 3) {
      safe_sample_correlation(kdm6b_mean_expression, gene_mean_expression)
    } else {
      NA_real_
    },
    p_value = if (n_samples >= 3) {
      stats::cor.test(
        kdm6b_mean_expression,
        gene_mean_expression,
        method = "spearman",
        exact = FALSE
      )$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    adjusted_p_value = stats::p.adjust(
      p_value,
      method = "BH"
    )
  )

correlation_summary <- correlation_summary %>%
  dplyr::left_join(program_genes, by = "gene") %>%
  dplyr::relocate(program, gene) %>%
  dplyr::arrange(program, dplyr::desc(spearman_rho), gene)

sample_correlation_output <- sample_mean_output %>%
  dplyr::left_join(
    program_genes,
    by = "gene",
    relationship = "many-to-many"
  ) %>%
  dplyr::relocate(program, gene, sample, condition) %>%
  dplyr::arrange(program, gene, condition, sample)

readr::write_tsv(
  correlation_summary,
  file.path(results_dir, "myeloid_kdm6b_gene_correlation_summary.tsv")
)
readr::write_tsv(
  sample_correlation_output,
  file.path(results_dir, "myeloid_kdm6b_gene_mean_expression_by_sample.tsv")
)

correlation_summary
```

## Program-specific sample-level scatter plots

Each facet shows sample-level mean `Kdm6b` expression against sample-level mean
expression of one target gene. Each point represents one biological sample. The
line is a fitted linear trend, and each facet reports the Spearman correlation
and BH-adjusted p-value. Genes present in more than one program are shown in each
relevant plot.

```{r program-correlation-scatterplots, fig.width=12, fig.height=8}
for (program_name in program_levels) {
  program_gene_levels <- program_genes %>%
    dplyr::filter(program == program_name) %>%
    dplyr::pull(gene)

  plot_data <- sample_correlation_output %>%
    dplyr::filter(program == program_name) %>%
    dplyr::mutate(
      gene = factor(gene, levels = program_gene_levels)
    )

  annotation_data <- correlation_summary %>%
    dplyr::filter(program == program_name) %>%
    dplyr::mutate(
      gene = factor(gene, levels = program_gene_levels),
      label = paste0(
        "rho = ", sprintf("%.2f", spearman_rho),
        "\nFDR = ", format.pval(adjusted_p_value, digits = 2, eps = 1e-3)
      )
    )

  p <- ggplot(
    plot_data,
    aes(x = kdm6b_mean_expression, y = gene_mean_expression)
  ) +
    geom_smooth(
      data = plot_data,
      aes(x = kdm6b_mean_expression, y = gene_mean_expression),
      method = "lm",
      formula = y ~ x,
      se = FALSE,
      color = "#B2182B",
      linewidth = 0.6,
      inherit.aes = FALSE
    ) +
    geom_point(size = 2.2, color = "#2166AC", alpha = 0.85) +
    geom_text(
      data = annotation_data,
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.15,
      size = 3
    ) +
    facet_wrap(~gene, scales = "free_y", ncol = 3) +
    theme_bw(base_size = 11) +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      strip.background = element_rect(fill = "grey95", color = "grey70")
    ) +
    labs(
      title = paste("Kdm6b and immune-program genes:", program_name),
      x = "Sample mean log-normalized Kdm6b expression",
      y = "Sample mean log-normalized gene expression"
    )

  print(p)

  output_label <- gsub("[^a-z0-9]+", "_", tolower(program_name))
  ggsave(
    file.path(
      plot_dir,
      paste0("myeloid_kdm6b_correlation_", output_label, "_scatterplots.pdf")
    ),
    plot = p,
    width = 12,
    height = 8,
    device = grDevices::pdf,
    bg = "white"
  )
}
```

## Program-specific correlation dot plots

The dot plots provide compact summaries of the same sample-level correlations
shown in the scatter plots.

```{r program-correlation-dotplots, fig.width=6, fig.height=4}
rho_limit <- max(abs(correlation_summary$spearman_rho), na.rm = TRUE)
rho_limit <- max(rho_limit, 0.1)

for (program_name in program_levels) {
  output_label <- gsub("[^a-z0-9]+", "_", tolower(program_name))
  program_gene_levels <- program_genes %>%
    dplyr::filter(program == program_name) %>%
    dplyr::pull(gene)

  plot_data <- correlation_summary %>%
    dplyr::filter(program == program_name) %>%
    dplyr::mutate(
      gene = factor(gene, levels = rev(program_gene_levels)),
      minus_log10_fdr = -log10(
        pmax(adjusted_p_value, .Machine$double.xmin)
      )
    )

  p_dot <- ggplot(
    plot_data,
    aes(
      x = spearman_rho,
      y = gene,
      color = spearman_rho,
      size = minus_log10_fdr
    )
  ) +
    geom_vline(xintercept = 0, color = "grey70", linewidth = 0.5) +
    geom_point(alpha = 0.9) +
    scale_color_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-rho_limit, rho_limit),
      oob = scales::squish
    ) +
    theme_bw(base_size = 11) +
    theme(
      panel.grid = element_blank(),
      plot.margin = ggplot2::margin(2, 2, 2, 2)
    ) +
    labs(
      title = paste("Kdm6b correlation:", program_name),
      x = "Spearman rho",
      y = NULL,
      color = "Spearman rho",
      size = "-log10(FDR)"
    )

  ggsave(
    file.path(
      plot_dir,
      paste0("myeloid_kdm6b_correlation_", output_label, "_dotplot.pdf")
    ),
    plot = p_dot,
    width = 6,
    height = 4,
    device = grDevices::pdf,
    bg = "white"
  )

}
```

## Session info

```{r session-info}
sessionInfo()
```
