---
title: "18-kdm6b-immune-programs"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
  ranking_tsv: "myeloid_kdm6b_subcluster_ranking_and_groups.tsv"
  subtype_label: "myeloid"
---

This analysis evaluates predefined immune programs across selected-compartment
subclusters. For myeloid cells, the programs are antigen presentation,
inflammatory activation, and T-cell recruitment. For T cells, the programs are
cytotoxicity, expansion/proliferation, memory/stem-like, exhaustion/dysfunction,
IFN response, and chemokine responsiveness. Gene-level expression and detection
are summarized by numeric subcluster, and UCell scores are calculated for each
selected-compartment cell.

Program scores are summarized across all subclusters, including intermediate
subclusters. The Kdm6b-high versus Kdm6b-low comparison excludes intermediate
subclusters and uses paired median scores from each biological sample. These
program differences describe the predefined subcluster groups and should not be
interpreted as effects caused by Kdm6b.

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(Seurat)
  library(SeuratObject)
  library(UCell)
})

root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v5")
analysis_dir <- file.path(root_dir, "analyses", "downstream-analyses")
analysis_label <- params$subtype_label
compartment_label <- dplyr::case_when(
  analysis_label == "tcell" ~ "T-cell",
  analysis_label == "myeloid" ~ "myeloid",
  TRUE ~ analysis_label
)
subtype_col <- paste0(analysis_label, "_subtype")
if (!analysis_label %in% c("myeloid", "tcell")) {
  stop("subtype_label must be either myeloid or tcell.")
}
results_dir <- file.path(analysis_dir, "results", analysis_label, "kdm6b")
plot_dir <- file.path(analysis_dir, "plots", analysis_label, "kdm6b")
output_prefix <- paste0(analysis_label, "_kdm6b_immune_program_")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

source(file.path(root_dir, "figures", "theme.R"))
```

## Gene programs

```{r analysis-parameters}
input_file <- file.path(data_dir, params$input_rds)
ranking_file <- file.path(results_dir, params$ranking_tsv)
missing_input_files <- c(input_file, ranking_file)[
  !file.exists(c(input_file, ranking_file))
]
if (length(missing_input_files) > 0) {
  stop("Missing input file(s): ", paste(missing_input_files, collapse = ", "))
}

program_signatures_by_compartment <- list(
  myeloid = list(
    Antigen_presentation = c(
      "H2-Ab1", "Cd74", "H2-Eb1", "B2m", "Psmb8", "Tap1", "Tapbp"
    ),
    Inflammatory_activation = c(
      "Il12b", "Ifngr1", "Stat1", "Irf1", "Irf7", "Cxcl9", "Cxcl10", "Tnf"
    ),
    T_cell_recruitment = c(
      "Ccl5", "Cxcl9", "Cxcl10", "Xcl1", "Xcr1", "Ccr7", "Ccl19", "Ccl21a"
    )
  ),
  tcell = list(
    Cytotoxicity = c(
      "Gzmb", "Prf1", "Nkg7", "Ifng", "Ccl3", "Ccl4"
    ),
    Expansion_proliferation = c(
      "Mki67", "Top2a", "Stmn1"
    ),
    Memory_stem_like = c(
      "Tcf7", "Il7r", "Ccr7", "Lef1"
    ),
    Exhaustion_dysfunction = c(
      "Pdcd1", "Lag3", "Havcr2", "Tox", "Entpd1"
    ),
    IFN_response = c(
      "Isg15", "Ifit3", "Stat1", "Irf7"
    ),
    Chemokine_responsiveness = c(
      "Cxcr3", "Ccr5", "Ccr7"
    )
  )
)
program_labels_by_compartment <- list(
  myeloid = c(
    Antigen_presentation = "Antigen presentation",
    Inflammatory_activation = "Inflammatory activation",
    T_cell_recruitment = "T-cell recruitment"
  ),
  tcell = c(
    Cytotoxicity = "Cytotoxicity",
    Expansion_proliferation = "Expansion/proliferation",
    Memory_stem_like = "Memory/stem-like",
    Exhaustion_dysfunction = "Exhaustion/dysfunction",
    IFN_response = "IFN response",
    Chemokine_responsiveness = "Chemokine responsiveness"
  )
)
program_signatures <- program_signatures_by_compartment[[analysis_label]]
program_labels <- program_labels_by_compartment[[analysis_label]]
program_levels <- unname(program_labels[names(program_signatures)])
program_key <- tibble::tibble(
  score_name = names(program_signatures),
  program = factor(program_levels, levels = program_levels)
)
program_genes <- dplyr::bind_rows(lapply(names(program_signatures), function(x) {
  tibble::tibble(score_name = x, gene = program_signatures[[x]])
})) %>%
  dplyr::left_join(program_key, by = "score_name")

kdm6b_group_levels <- c("Kdm6b-high", "Intermediate", "Kdm6b-low")
kdm6b_group_colors <- c(
  "Kdm6b-high" = "#B2182B",
  "Intermediate" = "#BDBDBD",
  "Kdm6b-low" = "#2166AC"
)
```

## Load data and map Kdm6b groups

```{r load-data}
obj <- readRDS(input_file)
kdm6b_ranking <- readr::read_tsv(ranking_file, show_col_types = FALSE) %>%
  dplyr::rename(subtype = dplyr::all_of(subtype_col)) %>%
  dplyr::mutate(
    subcluster = as.character(subcluster),
    kdm6b_group = factor(
      kdm6b_group,
      levels = kdm6b_group_levels
    )
  )

required_metadata <- c("sample", "seurat_clusters")
required_ranking_columns <- c("subcluster", "subtype", "kdm6b_group")
stopifnot(all(required_metadata %in% colnames(obj@meta.data)))
stopifnot(all(required_ranking_columns %in% colnames(kdm6b_ranking)))

if (anyDuplicated(kdm6b_ranking$subcluster)) {
  stop("Each subcluster must have one Kdm6b group assignment.")
}

observed_subclusters <- unique(as.character(obj$seurat_clusters))
if (!setequal(observed_subclusters, kdm6b_ranking$subcluster)) {
  stop("The ranking table and selected-compartment object contain different subclusters.")
}

subcluster_levels <- as.character(sort(as.integer(observed_subclusters)))
kdm6b_ranking <- kdm6b_ranking %>%
  dplyr::arrange(kdm6b_group, as.integer(subcluster))
# Keep violin plots ordered by Kdm6b group, then ranking within group.
subcluster_plot_levels <- kdm6b_ranking %>%
  dplyr::mutate(
    kdm6b_group = factor(
      kdm6b_group,
      levels = c("Kdm6b-high", "Intermediate", "Kdm6b-low")
    )
  ) %>%
  dplyr::arrange(kdm6b_group, kdm6b_rank) %>%
  dplyr::pull(subcluster)
kdm6b_group_by_subcluster <- setNames(
  as.character(kdm6b_ranking$kdm6b_group),
  kdm6b_ranking$subcluster
)
subcluster_axis_labels <- setNames(
  paste0(
    "C",
    subcluster_plot_levels,
    ":",
    kdm6b_group_by_subcluster[subcluster_plot_levels]
  ),
  subcluster_plot_levels
)
cluster_colors <- setNames(
  scales::hue_pal()(length(subcluster_levels)),
  subcluster_levels
)[subcluster_plot_levels]
obj$seurat_clusters <- factor(
  as.character(obj$seurat_clusters),
  levels = subcluster_plot_levels
)

group_index <- match(
  as.character(obj$seurat_clusters),
  kdm6b_ranking$subcluster
)
cell_kdm6b_group <- kdm6b_ranking$kdm6b_group[group_index]
if (anyNA(cell_kdm6b_group)) {
  stop("At least one cell could not be assigned to a Kdm6b group.")
}
obj$kdm6b_group <- factor(
  cell_kdm6b_group,
  levels = kdm6b_group_levels
)

SeuratObject::DefaultAssay(obj) <- "RNA"
log_expression <- SeuratObject::GetAssayData(
  obj,
  assay = "RNA",
  layer = "data"
)
raw_counts <- SeuratObject::GetAssayData(
  obj,
  assay = "RNA",
  layer = "counts"
)

missing_genes <- setdiff(unique(program_genes$gene), rownames(log_expression))
if (length(missing_genes) > 0) {
  stop("Program genes absent from the RNA assay: ", paste(missing_genes, collapse = ", "))
}
```

## Gene expression across selected-compartment subclusters

The violin plots show the distribution of log-normalized expression for each
gene across selected-compartment subclusters.

```{r summarize-gene-expression}
unique_program_genes <- unique(program_genes$gene)

gene_expression_by_subcluster <- dplyr::bind_rows(lapply(
  subcluster_levels,
  function(subcluster) {
    cell_index <- which(obj$seurat_clusters == subcluster)
    tibble::tibble(
      subcluster = subcluster,
      gene = unique_program_genes,
      mean_log_expression = as.numeric(
        Matrix::rowMeans(log_expression[unique_program_genes, cell_index, drop = FALSE])
      ),
      detection_fraction = as.numeric(
        Matrix::rowMeans(raw_counts[unique_program_genes, cell_index, drop = FALSE] > 0)
      )
    )
  }
)) %>%
  dplyr::left_join(program_genes, by = "gene", relationship = "many-to-many") %>%
  dplyr::left_join(
    kdm6b_ranking %>%
      dplyr::select(subcluster, subtype, kdm6b_group),
    by = "subcluster"
  ) %>%
  dplyr::mutate(
    subcluster = factor(subcluster, levels = subcluster_plot_levels),
    program = factor(program, levels = program_levels)
  ) %>%
  dplyr::arrange(program, gene, subcluster)

readr::write_tsv(
  gene_expression_by_subcluster %>%
    dplyr::select(-score_name),
  file.path(results_dir, paste0(output_prefix, "genes_by_subcluster.tsv"))
)
```

```{r plot-gene-expression-violins, fig.width=12, fig.height=6}
gene_expression_long <- tibble::as_tibble(
  t(as.matrix(log_expression[unique_program_genes, , drop = FALSE])),
  rownames = "cell"
) %>%
  dplyr::left_join(
    obj@meta.data %>%
      tibble::rownames_to_column("cell") %>%
      dplyr::transmute(
        cell,
        sample,
        subcluster = factor(
          as.character(seurat_clusters),
          levels = subcluster_plot_levels
        ),
        kdm6b_group
      ),
    by = "cell"
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(unique_program_genes),
    names_to = "gene",
    values_to = "log_normalized_expression"
  ) %>%
  dplyr::left_join(program_genes, by = "gene", relationship = "many-to-many") %>%
  dplyr::mutate(
    program = factor(program, levels = program_levels),
    subcluster = factor(subcluster, levels = subcluster_plot_levels),
    kdm6b_group = factor(kdm6b_group, levels = kdm6b_group_levels)
  )

gene_violin_file <- file.path(
  plot_dir,
  paste0(output_prefix, "gene_violin_by_subcluster.pdf")
)
grDevices::pdf(gene_violin_file, width = 12, height = 6, bg = "white")
for (gene_name in unique_program_genes) {
  p_gene_violin <- VlnPlot(
    obj,
    features = gene_name,
    group.by = "seurat_clusters",
    assay = "RNA",
    pt.size = 0,
    cols = cluster_colors
  ) &
    scale_x_discrete(
      limits = subcluster_plot_levels,
      labels = subcluster_axis_labels,
      drop = FALSE
    ) &
    theme_Publication() &
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      legend.background = element_rect(fill = "white", color = NA),
      legend.key = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) &
    labs(
      title = paste(gene_name, "expression by", compartment_label, "subcluster"),
      x = "Seurat cluster",
      y = "Log-normalized RNA expression"
    )
  print(p_gene_violin)
}
grDevices::dev.off()
```

## Kdm6b-high versus Kdm6b-low gene expression

Intermediate subclusters are excluded from this comparison. Gene expression is
first averaged within each sample and Kdm6b group, then compared with a paired
Wilcoxon signed-rank test across samples. This avoids treating individual cells
as independent biological replicates.

```{r summarize-gene-high-low}
gene_expression_by_sample_group <- gene_expression_long %>%
  dplyr::select(
    cell,
    sample,
    gene,
    log_normalized_expression,
    kdm6b_group
  ) %>%
  dplyr::distinct() %>%
  dplyr::filter(kdm6b_group != "Intermediate") %>%
  dplyr::group_by(gene, sample, kdm6b_group) %>%
  dplyr::summarise(
    mean_log_expression = mean(log_normalized_expression),
    n_cells = dplyr::n(),
    .groups = "drop"
  )

gene_high_low_tests <- gene_expression_by_sample_group %>%
  dplyr::select(gene, sample, kdm6b_group, mean_log_expression) %>%
  tidyr::pivot_wider(
    names_from = kdm6b_group,
    values_from = mean_log_expression
  ) %>%
  dplyr::filter(!is.na(`Kdm6b-high`), !is.na(`Kdm6b-low`)) %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    n_samples = dplyr::n(),
    median_high_expression = median(`Kdm6b-high`),
    median_low_expression = median(`Kdm6b-low`),
    median_paired_difference = median(`Kdm6b-high` - `Kdm6b-low`),
    p_value = stats::wilcox.test(
      `Kdm6b-high`,
      `Kdm6b-low`,
      paired = TRUE,
      exact = FALSE
    )$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(adjusted_p_value = stats::p.adjust(p_value, method = "BH"))

readr::write_tsv(
  gene_high_low_tests,
  file.path(results_dir, paste0(output_prefix, "gene_high_vs_low.tsv"))
)
```

```{r plot-gene-high-low-boxplots, fig.width=8, fig.height=6}
gene_boxplot_data <- gene_expression_by_sample_group %>%
  dplyr::mutate(
    kdm6b_group = factor(
      kdm6b_group,
      levels = c("Kdm6b-high", "Kdm6b-low")
    )
  )

gene_boxplot_file <- file.path(
  plot_dir,
  paste0(output_prefix, "gene_high_vs_low_boxplots.pdf")
)
grDevices::pdf(gene_boxplot_file, width = 8, height = 6, bg = "white")
gene_boxplot_labels <- gene_high_low_tests %>%
  dplyr::left_join(
    program_genes %>% dplyr::select(program, gene) %>% dplyr::distinct(),
    by = "gene"
  ) %>%
  dplyr::mutate(
    p_label = paste0(
      "p = ",
      format.pval(p_value, digits = 3, eps = 0.001)
    )
  )
gene_boxplot_y_positions <- gene_boxplot_data %>%
  dplyr::group_by(gene) %>%
  dplyr::summarise(
    y_position = max(mean_log_expression, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(y_position = ifelse(y_position > 0, y_position * 1.08, 0.1))

for (program_name in program_levels) {
  program_genes_to_plot <- program_genes %>%
    dplyr::filter(program == program_name) %>%
    dplyr::pull(gene) %>%
    unique()
  program_boxplot_data <- gene_boxplot_data %>%
    dplyr::filter(gene %in% program_genes_to_plot)
  program_boxplot_labels <- gene_boxplot_labels %>%
    dplyr::filter(program == program_name) %>%
    dplyr::left_join(gene_boxplot_y_positions, by = "gene")
  p_gene_boxplot <- ggplot(
    program_boxplot_data,
    aes(x = kdm6b_group, y = mean_log_expression, fill = kdm6b_group)
  ) +
    geom_line(
      aes(group = sample),
      color = "grey70",
      linewidth = 0.5
    ) +
    geom_boxplot(
      width = 0.55,
      outlier.shape = NA,
      color = "grey20"
    ) +
    geom_point(
      aes(fill = kdm6b_group),
      color = "black",
      shape = 21,
      alpha = 0.85,
      position = position_jitter(width = 0.05, height = 0),
      size = 1.2
    ) +
    scale_fill_manual(values = kdm6b_group_colors, drop = FALSE) +
    scale_x_discrete(
      labels = c(
        "Kdm6b-high" = "Kdm6b-high",
        "Kdm6b-low" = "Kdm6b-low"
      ),
      drop = FALSE
    ) +
    theme_Publication() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(size = 8),
      legend.position = "none"
    ) +
    facet_wrap(~ gene, scales = "free_y", ncol = 4) +
    geom_text(
      data = program_boxplot_labels,
      aes(x = "Kdm6b-low", y = y_position, label = p_label),
      inherit.aes = FALSE,
      size = 3
    ) +
    labs(
      title = paste(program_name, "gene expression"),
      x = NULL,
      y = "Sample-level mean log-normalized expression"
    )
  print(p_gene_boxplot)
}
grDevices::dev.off()
```

## UCell program scores

UCell scores are calculated for every selected-compartment cell. The object is not saved;
only sample- and subcluster-level summaries are written.

```{r calculate-ucell-scores, message=FALSE, warning=FALSE}
obj <- UCell::AddModuleScore_UCell(
  obj,
  features = program_signatures,
  name = ""
)

score_columns <- names(program_signatures)
if (!all(score_columns %in% colnames(obj@meta.data))) {
  stop("UCell did not create the expected score columns.")
}

cell_scores <- obj@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::select(
    cell,
    sample,
    subcluster = seurat_clusters,
    kdm6b_group,
    dplyr::all_of(score_columns)
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(score_columns),
    names_to = "score_name",
    values_to = "ucell_score"
  ) %>%
  dplyr::left_join(program_key, by = "score_name") %>%
  dplyr::mutate(
    subcluster = as.character(subcluster),
    kdm6b_group = factor(kdm6b_group, levels = kdm6b_group_levels)
  )
```

```{r summarize-ucell-scores}
ucell_by_subcluster <- cell_scores %>%
  dplyr::group_by(program, subcluster, kdm6b_group) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    median_ucell_score = median(ucell_score),
    q25_ucell_score = stats::quantile(ucell_score, 0.25, names = FALSE),
    q75_ucell_score = stats::quantile(ucell_score, 0.75, names = FALSE),
    .groups = "drop"
  ) %>%
  dplyr::left_join(
    kdm6b_ranking %>% dplyr::select(subcluster, subtype),
    by = "subcluster"
  ) %>%
  dplyr::arrange(program, as.integer(subcluster))

ucell_by_sample_group <- cell_scores %>%
  dplyr::filter(kdm6b_group != "Intermediate") %>%
  dplyr::group_by(program, sample, kdm6b_group) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    median_ucell_score = median(ucell_score),
    .groups = "drop"
  )

pair_check <- ucell_by_sample_group %>%
  dplyr::count(program, sample, name = "n_groups")
if (any(pair_check$n_groups != 2)) {
  stop("Every sample must contribute both Kdm6b-high and Kdm6b-low scores.")
}

ucell_high_low_tests <- ucell_by_sample_group %>%
  dplyr::select(program, sample, kdm6b_group, median_ucell_score) %>%
  tidyr::pivot_wider(
    names_from = kdm6b_group,
    values_from = median_ucell_score
  ) %>%
  dplyr::group_by(program) %>%
  dplyr::summarise(
    n_samples = dplyr::n(),
    median_high_score = median(`Kdm6b-high`),
    median_low_score = median(`Kdm6b-low`),
    median_paired_difference = median(`Kdm6b-high` - `Kdm6b-low`),
    p_value = stats::wilcox.test(
      `Kdm6b-high`,
      `Kdm6b-low`,
      paired = TRUE,
      exact = FALSE
    )$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(adjusted_p_value = stats::p.adjust(p_value, method = "BH"))

readr::write_tsv(
  ucell_by_subcluster,
  file.path(results_dir, paste0(output_prefix, "ucell_by_subcluster.tsv"))
)
readr::write_tsv(
  ucell_by_sample_group,
  file.path(results_dir, paste0(output_prefix, "ucell_by_sample_group.tsv"))
)
readr::write_tsv(
  ucell_high_low_tests,
  file.path(results_dir, paste0(output_prefix, "ucell_high_vs_low.tsv"))
)

ucell_high_low_tests
```

```{r plot-ucell-violins-by-subcluster, fig.width=12, fig.height=6}
ucell_violin_data <- cell_scores %>%
  dplyr::mutate(
    subcluster = factor(subcluster, levels = subcluster_plot_levels),
    program = factor(program, levels = program_levels)
  )

ucell_violin_file <- file.path(
  plot_dir,
  paste0(output_prefix, "ucell_violin_by_subcluster.pdf")
)
grDevices::pdf(ucell_violin_file, width = 12, height = 6, bg = "white")
for (program_name in program_levels) {
  p_ucell_violin <- ggplot(
    dplyr::filter(ucell_violin_data, program == program_name),
    aes(x = subcluster, y = ucell_score, fill = subcluster)
  ) +
    geom_violin(
      scale = "width",
      trim = TRUE,
      linewidth = 0.5,
      color = "black"
    ) +
    scale_x_discrete(
      limits = subcluster_plot_levels,
      labels = subcluster_axis_labels,
      drop = FALSE
    ) +
    scale_fill_manual(values = cluster_colors, drop = FALSE) +
    theme_Publication() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) +
    labs(
      title = paste(program_name, "UCell score by", compartment_label, "subcluster"),
      x = "Seurat cluster",
      y = "UCell score"
    )
  print(p_ucell_violin)
}
grDevices::dev.off()
```

## Kdm6b-high versus Kdm6b-low UCell scores

Intermediate subclusters are excluded. UCell scores are summarized within
each sample and group, then compared with the paired Wilcoxon signed-rank test.

```{r plot-ucell-high-low-boxplots, fig.width=6, fig.height=5}
ucell_boxplot_data <- ucell_by_sample_group %>%
  dplyr::mutate(
    kdm6b_group = factor(
      kdm6b_group,
      levels = c("Kdm6b-high", "Kdm6b-low")
    )
  )

ucell_boxplot_file <- file.path(
  plot_dir,
  paste0(output_prefix, "ucell_high_vs_low_boxplots.pdf")
)
grDevices::pdf(ucell_boxplot_file, width = 6, height = 5, bg = "white")
for (program_name in program_levels) {
  test_row <- ucell_high_low_tests %>% dplyr::filter(program == program_name)
  p_ucell_boxplot <- ggplot(
    dplyr::filter(ucell_boxplot_data, program == program_name),
    aes(x = kdm6b_group, y = median_ucell_score, fill = kdm6b_group)
  ) +
    geom_line(
      aes(group = sample),
      color = "grey70",
      linewidth = 0.5
    ) +
    geom_boxplot(
      width = 0.55,
      outlier.shape = NA,
      color = "grey20"
    ) +
    geom_point(
      aes(fill = kdm6b_group),
      color = "black",
      shape = 21,
      alpha = 0.85,
      position = position_jitter(width = 0.05, height = 0),
      size = 1.8
    ) +
    scale_fill_manual(values = kdm6b_group_colors, drop = FALSE) +
    scale_x_discrete(
      labels = c(
        "Kdm6b-high" = "Kdm6b-high",
        "Kdm6b-low" = "Kdm6b-low"
      ),
      drop = FALSE
    ) +
    theme_Publication() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "grey90"),
      panel.grid.minor = element_blank(),
      #axis.text.x = element_text(size = 8),
      legend.position = "none"
    ) +
    labs(
      title = paste(program_name, "UCell score"),
      subtitle = paste(
        "Paired Wilcoxon p =",
        format.pval(test_row$p_value, digits = 3, eps = 0.001)
      ),
      x = NULL,
      y = "Sample-level median UCell score"
    )
  print(p_ucell_boxplot)
}
grDevices::dev.off()
```

## Session info

```{r session-info}
sessionInfo()
```
