---
title: "01-tlr-analysis"
author: "Sam Chen, Bicna Song"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
---

## Expected output files

- `results/tlr_module_scores.rds`
- `plots/tlr_module_scores_umap.pdf`
- `plots/tlr_module_scores_umap_by_condition.pdf`
- `plots/tlr_module_scores_heatmap_by_condition.pdf`
- `plots/tlr_module_scores_heatmap_by_subtype.pdf`
- `plots/tlr_module_scores_heatmap_by_subtype_condition.pdf`
- `plots/tlr_module_scores_violin_by_condition.pdf`
- `plots/tlr_module_scores_violin_by_subtype.pdf`
- `plots/tlr_module_scores_violin_by_subtype_condition.pdf`

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(UCell)
library(pheatmap)
library(patchwork)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v4")
analysis_dir <- file.path(root_dir, "analyses", "tlr-analysis")
plot_dir <- file.path(analysis_dir, "plots")
results_dir <- file.path(analysis_dir, "results")
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# set random seed
set.seed(1234)
```

## Load data
```{r load-seurat-object}
obj <- readRDS(file.path(data_dir, "cart_myeloid_subtypes.rds"))

# Standardize subtype naming
obj$myeloid_subtype <- obj$myeloid_subtype %>%
  stringr::str_replace(
    "Perivascular macrophages.*",
    "Perivascular macrophages"
  )

unique(obj$myeloid_subtype)
```

```{r define-condition-order}
# Define condition order and colors once
condition_levels <- c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB", "STOP", "tumor")

condition_colors <- c(
  "41BB-L" = "#7570B3",
  "B7H3" = "#E7298A",
  "CD28-41BB" = "#1B9E77",
  "CD8-41BB" = "#D95F02",
  "STOP" = "#BDBDBD",
  "tumor" = "#66A61E"
)

obj$condition <- factor(obj$condition, levels = condition_levels)
```

## Define TLR gene sets
```{r define-tlr-genes}
# ---- Receptors ----
receptors <- c(
  "Tlr7","Tlr8"
)

# ---- Core Signaling Cascade ----
core_signaling_cascade <- c(
  "Myd88","Irak1","Irak2","Irak4","Traf6","Traf3",
  "Tab1","Tab2","Tab3","Map3k7","Ikbkb","Ikbkg",
  "Chuk","Tbk1","Ikbke"
)

# ---- IRF / Interferon Transcription Factors ----
irf_interferon_tfs <- c(
  "Irf1","Irf3","Irf5","Irf7","Irf8","Irf9",
  "Stat1","Stat2"
)

# ---- Type I Interferon / ISGs ----
typeI_ifn_isgs <- c(
  "Isg15","Isg20","Ifit1","Ifit2","Ifit3","Ifi27","Ifi35","Ifi44","Ifi44l","Oas1a",
  "Oas2","Oas3","Oasl1","Rsad2","Bst2","Xaf1","Usp18","Ifih1","Plscr1","Samhd1"
) # Warning: The following features are not present in the object: Ifit5, Ifi6, Mx1, Mx2, Ddx58

# ---- GBP Antimicrobial / IFN-Induced ----
#gbp_antimicrobial <- c(
#  "Gbp2","Gbp3","Gbp4","Gbp5"
#) # Warning: The following features are not present in the object: Gbp1

# ---- NF-kB / Inflammatory Cytokines ----
nfkb_inflammatory_cytokines <- c(
  "Tnf","Il1a","Il1b","Il6","Il12a","Il12b",
  "Il18","Nfkbia","Nfkbib","Nfkbiz","Rela","Relb",
  "Nfkb1","Nfkb2"
)

# ---- Chemokines ----
chemokines <- c(
  "Ccl2","Ccl3","Ccl4","Ccl5","Cxcl1","Cxcl9",
  "Cxcl10"
) # Warning: The following features are not present in the object: Cxcl11

# ---- Antigen Presentation / Co-stimulation ----
antigen_presentation_costimulation <- c(
  "Cd40","Cd80","Cd86","Ciita","H2-Aa","H2-Ab1",
  "H2-Eb1","H2-DMa","H2-DMb1"
)

# ---- Myeloid Activation / Polarization ----
myeloid_activation_polarization <- c(
  "Axl","Batf2","Ido1","Socs1","Socs3","Cd274"
) # Warning: The following features are not present in the object: Lilrb4

# ---- Inflammasome / Innate Amplification ----
inflammasome_innate_amplification <- c(
  "Nlrp3","Pycard","Casp1"
)

# ---- Antiviral / RNA Sensing Crosstalk ----
#antiviral_rna_sensing_crosstalk <- c(
#  "Ddx60","Dhx58"
#)

# ---- Additional Myeloid Response / Activation Markers ----
additional_myeloid_response_activation <- c(
  "Fcgr1","Fcgr2b","Fcgr3","Cd69","Ccr7","Ccr2"
)
```

## Calculate module scores
```{r calculate-module-scores}
# Combine all gene sets into a list
tlr_gene_sets <- list(
  Receptors = receptors,
  Core_Signaling_Cascade = core_signaling_cascade,
  IRF_Interferon_TFs = irf_interferon_tfs,
  TypeI_IFN_ISGs = typeI_ifn_isgs,
  #GBP_Antimicrobial = gbp_antimicrobial,
  NFkB_Inflammatory_Cytokines = nfkb_inflammatory_cytokines,
  Chemokines = chemokines,
  Antigen_Presentation_CoStimulation = antigen_presentation_costimulation,
  Myeloid_Activation_Polarization = myeloid_activation_polarization,
  Inflammasome_Innate_Amplification = inflammasome_innate_amplification,
  #Antiviral_RNA_Sensing_Crosstalk = antiviral_rna_sensing_crosstalk,
  Additional_Myeloid_Response_Activation = additional_myeloid_response_activation
)
# Calculate module scores for each gene set using UCell
obj <- AddModuleScore_UCell(
  obj,
  features = tlr_gene_sets,
  name = ""
)
```

```{r save-module-scores}
saveRDS(obj, file = file.path(results_dir, "tlr_module_scores.rds"))
```

## UMAP of all myeloid cells
```{r umap-myeloid, fig.width=12, fig.height=8}
DimPlot(obj, reduction = "umap", group.by = "myeloid_subtype") + 
  plot_annotation(title = "UMAP of Myeloid Cells")
```

## UMAP visualization of module scores
```{r umap-module-scores, fig.width=15, fig.height=16}
# Clean feature names for display
clean_features <- gsub("_", " ", names(tlr_gene_sets))

# Create UMAP plots
p <- FeaturePlot(
  obj,
  features = names(tlr_gene_sets),
  ncol = 3,
  reduction = "umap",
  cols = c("lightgrey", "#E65100"),
  order = TRUE,
  max.cutoff = "q95",
  pt.size = 1.5
) &
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

# Replace titles manually
for (i in seq_along(p)) {
  p[[i]] <- p[[i]] + ggtitle(clean_features[i])
}

print(p)

ggsave(
  filename = file.path(plot_dir, "tlr_module_scores_umap.pdf"),
  plot = p,
  width = 15,
  height = 16,
  units = "in"
)
```

```{r umap-module-scores-split-by-condition}
features <- names(tlr_gene_sets)
conditions <- levels(factor(obj$condition))

pdf(file.path(plot_dir, "tlr_module_scores_umap_by_condition.pdf"),
    width = 15, height = 8)

for (f in features) {

  plist <- FeaturePlot(
    obj,
    features = f,
    reduction = "umap",
    split.by = "condition",
    cols = c("lightgrey", "#E65100"),
    order = TRUE,
    max.cutoff = "q95",
    raster = TRUE,
    pt.size = 1.2,
    combine = FALSE
  )

  # Replace panel titles with condition names
  plist <- Map(function(p, cond) {
    p +
      ggtitle(cond) +
      theme_classic() +
      theme(
        plot.title = element_text(
          hjust = 0.5,
          size = 14,
          face = "bold"
        ),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.line.y.right = element_blank(),
        axis.title.y.right = element_blank(),
        axis.text.y.right  = element_blank(),
        axis.ticks.y.right = element_blank()
      )
  }, plist, conditions)

  p <- wrap_plots(plist, ncol = 3) +
    plot_annotation(
      title = gsub("_", " ", f),
      theme = theme(
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        plot.margin = margin(10, 20, 10, 20)
      )
    )

  print(p)
}

dev.off()
```

## Heatmaps of module scores by conditions
```{r heatmap-module-scores-condition, fig.width=10, fig.height=13}
modules <- names(tlr_gene_sets)

df <- obj@meta.data %>%
  select(condition, all_of(modules)) %>%
  group_by(condition) %>%
  summarise(across(all_of(modules), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

mat <- df %>%
  select(all_of(modules)) %>%
  as.matrix()

rownames(mat) <- df$condition
storage.mode(mat) <- "double"

mat_z <- scale(mat)
mat_t <- t(mat_z)
rownames(mat_t) <- gsub("_", " ", rownames(mat_t))

mat_t <- mat_t[, condition_levels]

annotation_col <- data.frame(
  condition = colnames(mat_t)
)
rownames(annotation_col) <- colnames(mat_t)

ann_colors <- list(
  condition = condition_colors
)

pheatmap(
  mat_t,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  breaks = seq(-2, 2, length.out = 101),
  border_color = NA,
  fontsize_row = 11,
  fontsize_col = 12,
  angle_col = 45,
  show_rownames = TRUE,
  show_colnames = FALSE,
  legend = TRUE,
  legend_breaks = c(-2, -1, 0, 1, 2),
  legend_labels = c("-2", "-1", "0", "1", "2"),
  filename = file.path(plot_dir, "tlr_module_scores_heatmap_by_condition.pdf"),
  width = 9,
  height = 7
)
```

## Heatmaps of module scores by subtype
```{r heatmap-module-scores-subtype, fig.width=12, fig.height=10}
obj_sub <- subset(
  obj,
  subset = myeloid_subtype != "Inflammatory monocytes"
)

df <- obj_sub@meta.data %>%
  select(myeloid_subtype, all_of(modules)) %>%
  group_by(myeloid_subtype) %>%
  summarise(across(all_of(modules), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  arrange(myeloid_subtype)

mat <- df %>%
  select(all_of(modules)) %>%
  as.matrix()

rownames(mat) <- df$myeloid_subtype
storage.mode(mat) <- "double"

mat_z <- scale(mat)
mat_t <- t(mat_z)
rownames(mat_t) <- gsub("_", " ", rownames(mat_t))

annotation_col <- data.frame(
  subtype = colnames(mat_t)
)
rownames(annotation_col) <- colnames(mat_t)

# cluster subtype columns and extract order
ph_subtype <- pheatmap(
  mat_t,
  annotation_col = annotation_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  breaks = seq(-2, 2, length.out = 101),
  border_color = NA,
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 45,
  show_rownames = TRUE,
  show_colnames = FALSE,
  legend = TRUE,
  legend_breaks = c(-2, -1, 0, 1, 2),
  legend_labels = c("-2", "-1", "0", "1", "2"),
  silent = TRUE
)

# get subtype clustering order from heatmap
subtype_order <- colnames(mat_t)[ph_subtype$tree_col$order]

# reorder annotation legend using clustering order
annotation_col$subtype <- factor(annotation_col$subtype, levels = subtype_order)

subtype_colors <- setNames(
  DiscretePalette(length(subtype_order)),
  subtype_order
)

ann_colors <- list(
  subtype = subtype_colors
)

# redraw with ordered legend/colors
ph_subtype <- pheatmap(
  mat_t[, subtype_order],
  annotation_col = annotation_col[subtype_order, , drop = FALSE],
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  breaks = seq(-2, 2, length.out = 101),
  border_color = NA,
  fontsize_row = 11,
  fontsize_col = 10,
  angle_col = 45,
  show_rownames = TRUE,
  show_colnames = FALSE,
  legend = TRUE,
  legend_breaks = c(-2, -1, 0, 1, 2),
  legend_labels = c("-2", "-1", "0", "1", "2"),
  filename = file.path(plot_dir, "tlr_module_scores_heatmap_by_subtype.pdf"),
  width = 12,
  height = 7
)
```

## Heatmap of module scores by subtype and condition
```{r heatmap-module-scores-subtype-condition, fig.width=14, fig.height=10}
obj_sub$myeloid_subtype <- factor(obj_sub$myeloid_subtype, levels = subtype_order)

df <- obj_sub@meta.data %>%
  select(myeloid_subtype, condition, all_of(modules)) %>%
  group_by(myeloid_subtype, condition) %>%
  summarise(across(all_of(modules), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  arrange(myeloid_subtype, condition)

mat <- df %>%
  select(all_of(modules)) %>%
  as.matrix()

rownames(mat) <- paste(df$myeloid_subtype, df$condition, sep = " | ")
storage.mode(mat) <- "double"

mat_z <- scale(mat)
mat_t <- t(mat_z)
rownames(mat_t) <- gsub("_", " ", rownames(mat_t))

ordered_cols <- paste(df$myeloid_subtype, df$condition, sep = " | ")
mat_t <- mat_t[, ordered_cols]

annotation_col <- data.frame(
  subtype = factor(df$myeloid_subtype, levels = subtype_order),
  condition = factor(df$condition, levels = condition_levels)
)
rownames(annotation_col) <- colnames(mat_t)

subtype_colors <- setNames(
  DiscretePalette(length(subtype_order)),
  subtype_order
)

ann_colors <- list(
  subtype = subtype_colors,
  condition = condition_colors
)

subtype_counts <- table(df$myeloid_subtype)
gaps_col <- cumsum(subtype_counts)

pheatmap(
  mat_t,
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  gaps_col = gaps_col[-length(gaps_col)],
  breaks = seq(-2, 2, length.out = 101),
  border_color = NA,
  fontsize_row = 10,
  fontsize_col = 8,
  angle_col = 90,
  show_rownames = TRUE,
  show_colnames = FALSE,
  legend = TRUE,
  legend_breaks = c(-2, -1, 0, 1, 2),
  legend_labels = c("-2", "-1", "0", "1", "2"),
  filename = file.path(plot_dir, "tlr_module_scores_heatmap_by_subtype_condition.pdf"),
  width = 14,
  height = 8
)
```

## Violin plots by condition
```{r violin-module-scores-condition, fig.height=8, fig.width=10}
pdf(file.path(plot_dir, "tlr_module_scores_violin_by_condition.pdf"),
    width = 6, height = 4)

for (f in features) {

  clean_title <- gsub("_", " ", f)

  p <- VlnPlot(
    obj,
    features = f,
    group.by = "condition",
    cols = condition_colors,
    pt.size = 0
  ) &
    ggtitle(clean_title) &
    theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        vjust = 1
      ),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.margin = margin(10, 20, 10, 20),
      legend.position = "none"
    )

  print(p)
}

dev.off()
```

## Violin plots excluding Inflammatory monocytes
```{r violin-module-scores-subtype, fig.height=8, fig.width=18}
pdf(file.path(plot_dir, "tlr_module_scores_violin_by_subtype.pdf"),
    width = 8, height = 6)

for (f in features) {

  clean_title <- gsub("_", " ", f)

  p <- VlnPlot(
    obj_sub,
    features = f,
    group.by = "myeloid_subtype",
    pt.size = 0
  ) &
    ggtitle(clean_title) &
    theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        vjust = 1,
        size = 10
      ),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.margin = margin(10, 30, 10, 40),
      legend.position = "none"
    )

  print(p)
}

dev.off()
```

## Violin plots excluding Inflammatory monocytes split by condition
```{r violin-module-scores-subtype-condition, fig.height=8, fig.width=18}
pdf(file.path(plot_dir, "tlr_module_scores_violin_by_subtype_condition.pdf"),
    width = 16, height = 6)

for (f in features) {

  clean_title <- gsub("_", " ", f)

  p <- VlnPlot(
    obj_sub,
    features = f,
    group.by = "myeloid_subtype",
    split.by = "condition",
    cols = condition_colors,
    pt.size = 0
  ) &
    ggtitle(clean_title) &
    theme(
      axis.text.x = element_text(
        angle = 45,
        hjust = 1,
        vjust = 1
      ),
      axis.title.x = element_blank(),
      plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.margin = margin(10, 30, 10, 40)
    )

  print(p)
}

dev.off()
```

```{r sessionInfo}
sessionInfo()
```

