---
title: "01-seurat-processing"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(presto)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v1")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))

# set random seed
set.seed(1234)
```

## Load Cell Ranger data
```{r load-data}
# All 14 sample names and conditions
samples <- c("Tumor_only_D1", "Tumor_only_D2", "STOP_pos_T_D1", "STOP_pos_T_D2",
             "B7H3_pos_T_D1", "B7H3_pos_T_D2",
             "41BB-L_pos_T_D1", "41BB-L_pos_T_D2", 
             "CD8-41BB_pos_T_D1", "CD8-41BB_pos_T_D2",
             "CD28-41BB_pos_T_D1", "CD28-41BB_pos_T_D2")

conditions <- c("tumor", "tumor", "STOP", "STOP",
                "B7H3", "B7H3",
                "41BB-L", "41BB-L",
                "CD8-41BB", "CD8-41BB",
                "CD28-41BB", "CD28-41BB")

### Load in data as a list of seurat objects
obj_list <- lapply(samples, function(sample) {
  data_path <- file.path(data_dir, sample)
  print(data_path)
  seurat_obj <- Read10X(data.dir = data_path) %>%
    # more lenient parameters for flex libraries
    CreateSeuratObject(project = sample, min.cells = 1, min.features = 50)
  seurat_obj[["sample"]] <- sample
  seurat_obj[["condition"]] <- conditions[which(samples == sample)]
  return(seurat_obj)
})

```

## Run doublet removal on individual samples
```{r run-doubletfinder, warning=FALSE, message=FALSE}
# Get Doublet/Singlet IDs by DoubletFinder()
doublet_res <- lapply(obj_list, run_doubletfinder)
doublet_res <- data.frame(bind_rows(doublet_res))
rownames(doublet_res) <- doublet_res$row_names
doublet_res$row_names <- NULL
table(doublet_res$doublet_finder)
```

## Merge Seurat objects
```{r merge-objects}
# Merge all seurat objects into one
combined_obj <- Reduce(function(x, y) merge(x, y), obj_list)
```

## Visualize and filter doublets
```{r visualize-filter-doublets, fig.width=8, fig.height=6, out.width='100%', fig.align='center', message=FALSE, warning=FALSE}
# Add Doublet/Singlet info to metadata
combined_obj <- AddMetaData(combined_obj, metadata = doublet_res, col.name = 'doublet_finder')

# Check how doublets singlets differ in QC measures per sample
VlnPlot(combined_obj, split.by = "doublet_finder",
        features = c("nFeature_RNA", "nCount_RNA"), 
        ncol = 2, pt.size = 0) + theme(legend.position = 'right')

# Filter out predicted doublets to keep only singlets for downstream analysis.
combined_obj <- subset(x = combined_obj, subset = doublet_finder == 'Singlet')

# Save the combined singlet object so I don't need to rerun doublet removal
saveRDS(combined_obj, file = file.path(results_dir, "cart_combined.rds"))
```

## Basic QC metrics
```{r visualize-qc-metrics, fig.width=10, fig.height=6, out.width='100%', fig.align='center', message=FALSE, warning=FALSE}

combined_obj <- readRDS(file.path(results_dir, "cart_combined.rds"))

# add mitochondrial percentage
combined_obj[["percent.mt"]] <- PercentageFeatureSet(combined_obj, pattern = "^mt-")
# Visualize QC metrics
p1 <- QC_Plots_Genes(seurat_object = combined_obj, alpha = 0.2, pt.size=0)
p2 <- QC_Plots_UMIs(seurat_object = combined_obj,  y_axis_log = TRUE, alpha = 0.2, pt.size=0)
p3 <- QC_Plots_Mito(seurat_object = combined_obj, mito_name = "percent.mt", high_cutoff = 10, alpha = 0.2, pt.size = 0)

plots <- list(p1, p2, p3)
plots <- lapply(plots, function(p)
  p + theme(
    plot.title = element_text(size = 10, margin = margin(b = 4)),
    plot.margin = margin(t = 6, r = 25, b = 6, l = 12)  # add column gutters
  )
)
wrap_plots(plots, ncol = 3)
```

## Perform adaptive QC filtering
```{r adaptive-qc-filtering}

# Before filtering
before_filtering <- table(combined_obj$sample)

# Filter on mitochondrial percentage first
combined_obj <- subset(combined_obj, subset = percent.mt < 10)

# Adaptive filtering on nFeature_RNA and nCount_RNA per sample to avoid removing too many cells
# because Flex libraries are generally lower quality than standard libraries
flex_qc <- function(seurat_obj, 
                    q_low=0.02, q_high=0.995,
                    floor_counts=100, floor_features=50) {
  mat <- GetAssayData(seurat_obj, slot="counts")
  nCount   <- Matrix::colSums(mat)
  nFeature <- Matrix::colSums(mat > 0)

  loC <- max(floor_counts,   quantile(nCount,   q_low,  na.rm=TRUE))
  loF <- max(floor_features, quantile(nFeature, q_low,  na.rm=TRUE))
  hiC <- quantile(nCount,   q_high, na.rm=TRUE)
  hiF <- quantile(nFeature, q_high, na.rm=TRUE)

  keep <- (nCount >= loC & nCount <= hiC &
           nFeature >= loF & nFeature <= hiF)

  subset(seurat_obj, cells = colnames(seurat_obj)[keep])
}

# split to perform per sample filtering
objs <- SplitObject(combined_obj, split.by = "sample")
objs_qc <- lapply(objs, function(x) {flex_qc(x)})

# Merge back into one object
combined_obj <- Reduce(function(x, y) merge(x, y), objs_qc)

# After filtering
after_filtering <- table(combined_obj$sample)

# Plot number of cells before and after filtering
qc_df <- data.frame(
  sample = rep(names(before_filtering), 2),
  n_cells = c(as.numeric(before_filtering), as.numeric(after_filtering)),
  status = rep(c("Before Filtering", "After Filtering"), each = length(before_filtering))
)

ggplot(qc_df, aes(x = sample, y = n_cells, fill = status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of Cells Before and After QC Filtering",
       x = "Sample",
       y = "Number of Cells") +
  scale_fill_manual(values = c("Before Filtering" = "lightblue", "After Filtering" = "salmon"))
```

## Data Normalization and scaling
```{r normalize-and-scale}
combined_obj <- NormalizeData(combined_obj, normalization.method = "LogNormalize", scale.factor = 10000)
combined_obj <- FindVariableFeatures(combined_obj, selection.method = "vst", nfeatures = 2000)
combined_obj <- ScaleData(combined_obj)
```

## Dimensional reduction with naive merging
```{r dimension-reduction-naive-merge}
combined_obj <- RunPCA(combined_obj)
# Check elbow plot
ElbowPlot(combined_obj, ndims = 50) # use 30 PCs for downstream analysis
```

## Clustering and UMAP with naive merging
```{r naive-merge-cluster-and-umap}
# clustering
combined_obj <- FindNeighbors(combined_obj, dims = 1:30, reduction = "pca")
combined_obj <- FindClusters(combined_obj, resolution = 0.5, cluster.name = "merged_clusters")

# umap 
combined_obj <- RunUMAP(combined_obj, dims = 1:30, reduction = "pca", reduction.name = "merged_umap")
```

## Visualize UMAP after naive merging
```{r visualize-naive-merge-umap, fig.width=12, fig.height=6, out.width='100%', fig.align='center'}
DimPlot(combined_obj, reduction = "merged_umap", 
        group.by = c("merged_clusters", "sample"), alpha = 0.25)
```
```{r split-native-merge-umap, fig.width=8, fig.height=16, out.width='100%', fig.align='center'}}
DimPlot(combined_obj, reduction = "merged_umap", 
        group.by = "sample", split.by = "sample", ncol = 4, alpha = 0.5) + 
  theme(
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    axis.text = element_text(size = 6),
    strip.text = element_text(size = 10, face = "bold")
  )
```

## Save naive merged object
```{r save-merged-object}
saveRDS(combined_obj, file = file.path(results_dir, "cart_merged.rds"))
```

## CCA integration
Run CCA-based integration to mitigate sample/batch effects and re-compute neighbors/clusters
```{r cca-integration, message=FALSE, warning=FALSE}
combined_obj <- readRDS(file.path(results_dir, "cart_merged.rds"))

integrated_obj <- IntegrateLayers(
  object = combined_obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca",
  verbose = FALSE
)

# re-join layers after integration
integrated_obj[["RNA"]] <- JoinLayers(integrated_obj[["RNA"]])
```

## Clustering and UMAP of integrated object
```{r cluster-and-umap-integrated}
# clustering
integrated_obj <- FindNeighbors(integrated_obj, reduction = "integrated.cca", dims = 1:30)
integrated_obj <- FindClusters(integrated_obj, resolution = 0.5)

# umap
integrated_obj <- RunUMAP(integrated_obj, reduction = "integrated.cca", dims = 1:30)
```

## Visualize UMAP of integrated object
```{r visualize-integrated-umap, fig.width=12, fig.height=6, out.width='100%', fig.align='center'}
DimPlot(integrated_obj, reduction = "umap", 
        group.by = c("seurat_clusters", "sample"), alpha = 0.25)
```

```{r split-integrated-umap, fig.width=8, fig.height=16, out.width='100%', fig.align='center'}}
DimPlot(integrated_obj, reduction = "umap", 
        group.by = "sample", split.by = "sample", ncol = 4, alpha = 0.5) + 
  theme(
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    axis.text = element_text(size = 6),
    strip.text = element_text(size = 10, face = "bold")
  )
```

## Save integrated object
```{r save-integrated-object}
saveRDS(integrated_obj, file = file.path(results_dir, "cart_integrated.rds"))
```

## Session Info
```{r session-info}
sessionInfo()
```
