---
title: "04-myeloid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(AUCell)
library(msigdbr)
library(sccomp)
library(pheatmap)
library(fgsea)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v2")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(results_dir, "cart_myeloid_subclusters.rds"))
```

## Re-run clusters in higher resolution
```{r rerun-clusters, message=FALSE, warning=FALSE}
myeloid_obj <- FindNeighbors(myeloid_obj, reduction = "integrated.cca", dims = 1:20)
myeloid_obj <- FindClusters(myeloid_obj, resolution = 0.7)
```

## Visualize myeloid subclusters
```{r visualize-myeloid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Cell Lineages")

p1 + p2
```

## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

write.csv(cluster_markers, file.path(results_dir, "myeloid_subcluster_markers.csv"), row.names = FALSE)
```

```{r microglia-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Sparc", "Hexb", "Csf1r", "Tmem119"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Microglia Core")

FeaturePlot_scCustom(myeloid_obj, features = c("Il1a", "Tnf", "Ccl3", "Cd83"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Activated microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Apoe", "Tyrobp", "C1qa", "Lpl"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Phagocytic microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Ifit3", "Isg15", "Rsad2", "Irf7"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Interferon microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Hspa1a", "Hspa1b", "Dnaja1", "Hsp90aa1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Stress response")
```

```{r macrophage-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Lyz2", "Ms4a7", "Cd14", "Ctss"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Macrophage Core")

FeaturePlot_scCustom(myeloid_obj, features = c("Gpnmb", "Fabp5", "Plin2", "Lipa"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Lipid-associated TAM")

FeaturePlot_scCustom(myeloid_obj, features = c("Cd74", "H2-Eb1", "H2-Ab1", "Ccr2"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Antigen presentation")

FeaturePlot_scCustom(myeloid_obj, features = c("Mrc1", "Dab2", "Gas6", "Cd163"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Perivascular macrophages")

FeaturePlot_scCustom(myeloid_obj, features = c("Plac8", "Msrb1", "S100a4", "Alox5ap"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Monocytes")
```

## Reannotate clusters by their expression profiles
```{r reannotate-clusters}
new_cluster_ids <- c(
  "0" = "Homeostatic/stress-response microglia",
  "4" = "Homeostatic/stress-response microglia",
  "5" = "Homeostatic/stress-response microglia",
  "2" = "Activated microglia",
  "7" = "Activated microglia",
  "11" = "Activated microglia",
  "1" = "Phagocytic microglia",
  "3" = "Lipid-associated macrophages",
  "8" = "Lipid-associated macrophages",
  "9" = "Lipid-associated macrophages",
  "6" = "Antigen-presenting macrophages",
  "10" = "Perivascular macrophages",
  "12" = "Interferon-responsive macrophages",
  "13" = "Monocytes",
  "14" = "Neurons"
)

myeloid_obj$myeloid_subtype <- plyr::mapvalues(
  x = myeloid_obj$seurat_clusters,
  from = names(new_cluster_ids),
  to = new_cluster_ids
)

DimPlot(myeloid_obj, reduction = "umap", group.by = "myeloid_subtype") + 
  ggtitle("Myeloid Subsets")
```

## Marker genes on subsets
```{r subset-markers, message=FALSE, warning=FALSE}
subtype_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                 group.by = "myeloid_subtype",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

write.csv(subtype_markers, file.path(results_dir, "myeloid_subtype_markers.csv"), row.names = FALSE)
```

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- subtype_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(myeloid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

## Composition analysis using sccomp

```{r visualize-split-umap, fig.height=8, fig.width=12}
DimPlot(myeloid_obj, reduction = "umap", group.by = "myeloid_subtype", split.by = "condition", ncol = 3) + 
  ggtitle("Myeloid Subclusters by Condition")
```

```{r composition-analysis-sccomp, message=FALSE, warning=FALSE}
sccomp_input <- myeloid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = myeloid_subtype,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('tumor', 'STOP', '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result = 
  sccomp_input |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r visualize-sccomp-results}
sccomp_result |> plot()
```

``` {r sccomp-pairwise-fc}
sccomp_result  |> 
   sccomp_proportional_fold_change(
     formula_composition = ~  condition,
     from =  "tumor", 
     to = "B7H3"
    ) |> 
  select(cell_group, statement)
```

## plot relative proportions of clusters between conditions
```{r plot-relative-proportions}

## stacked barplot
prop_df <- myeloid_obj@meta.data %>%
  count(condition, myeloid_subtype) %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(prop_df, aes(x = condition, y = prop, fill = myeloid_subtype)) +
  geom_col(position = position_dodge(width = 0.8)) +
  ylab("Proportion") +
  theme_classic()

## dotplot of proportions
ggplot(prop_df, aes(x = condition, y = myeloid_subtype, size = prop, color = prop)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_y_discrete(limits = rev) + 
  theme_classic() +
  ylab("Cluster") + 
  xlab("Condition")
```

# Gene set enrichment analysis

## AUCell scoring on MSigDB Hallmark Gene sets
```{r get-genesets, message=FALSE, warning=FALSE}
# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c5 onology
msigdb_c5 <- msigdbr::msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c7 immune
msigdb_c7 <- msigdbr::msigdbr(species = "Mus musculus", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by myeloid subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(myeloid_obj, assay = "RNA", slot = "data"), nCores = 4, plotStats = TRUE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)
```

```{r visualize-hallmark-activity, fig.height=20, fig.width=16}
# add hallmark scores to seurat object
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
myeloid_obj <- AddMetaData(myeloid_obj, metadata = hallmark_scores_df)
# visualize selected hallmark pathways
hallmark_pathways_to_plot <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                "HALLMARK_COAGULATION",
                                "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                "HALLMARK_COMPLEMENT",
                               "HALLMARK_GLYCOLYSIS",
                               "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",
                               "HALLMARK_E2F_TARGETS",
                               "HALLMARK_MYC_TARGETS_V1")
FeaturePlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, reduction = "umap",
                     pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_plasma_light_high) + 
  plot_annotation(title = "Hallmark Pathway Activity in Myeloid Subtypes")


VlnPlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, group.by = "myeloid_subtype",
                 pt.size = 0, log = FALSE, colors_use = viridis::viridis(9)) + 
  plot_annotation(title = "Hallmark Pathway Activity by Myeloid Subtype")
```

## Subtype vs pathway heatmap of AUCell scores
```{r subtype-pathway-heatmap, fig.height=8, fig.width=16}
# z scores of AUCell scores across clusters per gene set
hallmark_scores_df$cluster = myeloid_obj$myeloid_subtype

avg_hallmark_scores <- hallmark_scores_df %>%
  group_by(cluster) %>%
  summarise(across(starts_with("HALLMARK"), mean)) %>%
  column_to_rownames("cluster")

hallmark_z_scores <- scale(avg_hallmark_scores)


# Color scale
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)
breaks <- seq(-3, 3, length.out = 101)

pheatmap::pheatmap(mat = hallmark_z_scores,
                   color = colors.use,
                   border_color = "white",
                   breaks = breaks,
                   clustering_distance_rows = "correlation",
                  clustering_distance_cols = "correlation") 
```

## GSEA analysis of Hallmark gene sets
```{r gsea-hallmark}
# get foldchanges for each gene per cluster
clusters <- levels(myeloid_obj$myeloid_subtype)
fc_list <- lapply(clusters, function(cl){
  df <- FoldChange(
    object = myeloid_obj,
    group.by = "myeloid_subtype",
    ident.1 = cl,
    ident.2 = NULL,
    assay = "RNA",
    slot = "data"
  )
  df <- as.data.frame(df)
  df$gene <- rownames(df)

  # Filter by pct.1
  df <- df %>% filter(pct.1 >= 0.05)
  
  df$cluster <- cl
  df
})
names(fc_list) <- clusters  

# rank genes by product of log FCs and -log10 p values
make_rank_vector <- function(df) {
  stat <- df$avg_log2FC
  names(stat) <- df$gene
  stat[is.finite(stat)]
}

fgsea_list_hallmark <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))

  fgsea(
    pathways = msigdb_hallmark,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_hallmark <- imap_dfr(fgsea_list_hallmark, ~ mutate(.x, cluster = .y))
fgsea_res_hallmark_sig <- fgsea_res_hallmark %>% 
  filter(padj < 0.1 & NES > 0) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_hallmark_sig, file.path(results_dir, "myeloid_subtype_hallmark_gsea_results.csv"), row.names = FALSE)
```

## Visualize select significant genesets
```{r visualize-gsea-results, fig.height=8, fig.width=12}
c0_ranks <- make_rank_vector(fc_list[["Activated microglia"]])
c7_ranks <- make_rank_vector(fc_list[["Interferon-responsive macrophages"]])

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]], c0_ranks) +
  ggtitle("Activated microglia: TNFA Signaling via NFKB")

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]], c7_ranks) +
  ggtitle("Interferon-responsive macrophages: Interferon Gamma Response")
```

## GSEA on C5 GO gene sets
```{r gsea-c5}
fgsea_list_c5 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c5,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c5 <- imap_dfr(fgsea_list_c5, ~ mutate(.x, cluster = .y))
fgsea_res_c5_sig <- fgsea_res_c5 %>% 
  filter(padj < 0.1) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c5_sig, file.path(results_dir, "myeloid_subtype_c5_gsea_results.csv"), row.names = FALSE)
```

## GSEA on C7 immune gene sets
```{r gsea-c7}
fgsea_list_c7 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c7,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c7 <- imap_dfr(fgsea_list_c7, ~ mutate(.x, cluster = .y))
fgsea_res_c7_sig <- fgsea_res_c7 %>% 
       filter(padj < 0.1) %>%
       arrange(cluster, desc(NES)) %>%
       select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c7_sig, file.path(results_dir, "myeloid_subtype_c7_gsea_results.csv"), row.names = FALSE)
```

# Summary of the myeloid subtypes
AI-assisted summerization of the full marker lists for each myeloid subtype:

### Homeostatic / Stress-Response Microglia
- Key markers: Tmem119, Siglech, Sparc, Hspa1a/b, Hsp90aa1, Txnip
- Enriched in tumor and STOP samples
- A core microglial population maintaining tissue surveillance, enriched for canonical homeostatic genes (Tmem119, Siglech, Sparc) but showing a mild stress-response program marked by heat-shock proteins (Hspa1a/b, Hsp90aa1). Reflects stable microglia experiencing environmental or handling-related cellular stress.


### Activated microglia
- Key markers: Il1a, Tnf, Ccl3, Egr1, Atf3
- Enriched in tumor and STOP samples
- This population represents danger-sensing, cytokine-producing, immune-interactive microglia with strong NF-κB/AP-1 activity and maintained microglial identity. It aligns with injury-induced, tumor-associated, or cytokine-exposed activated microglia.

### Phagocytic microglia
- Key markers: Apoe, Trem2, C1qa, Tyrobp, Cst7
- Slightly enriched in CD28-41BB samples
- This population represents actively engulfing, lipid-processing phagocytic microglia with strong lysosomal and lipid-catabolic machinery. Their profile is consistent with DAM/LAM microglia responding to debris, dying cells, myelin, or tumor-derived material.

### Lipid-associated macrophages
- Key markers: Gpnmb, Fabp5, Lipa, Lgals3
- Enriched in B7H3, 41BBL, and CD28-41BB samples
- A highly lipid-scavenging, lysosome-expanded, Trem2-Tyrobp–driven macrophage population analogous to LAM/TAM signatures in cancer. These cells manage lipid-rich debris, remodel extracellular matrix, buffer oxidative stress, and maintain persistent low-grade antigen presentation.

### Antigen presenting macrophages
- Key markers: H2-Aa, H2-Ab1, Cd74, Vim, Srgn
- Enriched in CD8-41BB samples
- A highly activated, MHC-II–intense antigen-presenting macrophage population enriched for inflammatory cytokines and migration machinery.

### Interferon-responsive macrophages
- Key markers: Ifit2, Ifit3, Rsad2, Isg15, Irf7
- Enriched in B7H3 samples
- Highly antiviral, type I interferon–driven macrophages expressing strong ISG modules (Ifit/Isg15/Oasl/Rsad2), chemokines (Cxcl9/10), antigen-presentation machinery, and nucleic-acid sensors. Represents macrophages engaged in innate immune amplification, viral-mimicry sensing, or tumor-associated interferon signaling.

### Perivascular macrophages
- Key markers: Mrc1, Dab2, Gas6, Cd163, Pf4
- Depleted in B7H3 and 41BB-L samples
- A population specializing in endothelial interaction, chemokine-mediated leukocyte recruitment, oxidative-stress buffering, and vascular remodeling. These cells form a vessel-associated, tissue-resident macrophage compartment with a balanced inflammatory and tolerogenic signature.

### Monocytes: 
- Key markers: Plac8, Ccr2, S100a4, Alox5ap
- Enriched in B7H3 samples
- Represents the major infiltrating monocyte pool that seeds the macrophage and inflammatory myeloid states within the tissue.

### Neurons:
- Key markers: Ryr1, Syt7, Ank1, Itpr1
- Contamination

## Save final myeloid object
```{r save-final-object}
saveRDS(myeloid_obj, file.path(results_dir, "cart_myeloid_subtypes.rds"))
```

## Session Info
```{r session-info}
sessionInfo()
```
  
