---
title: "04-myeloid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(AUCell)
library(msigdbr)
library(sccomp)
library(pheatmap)
library(fgsea)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v3")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(results_dir, "cart_myeloid_subclusters.rds"))
```

## Visualize myeloid subclusters
```{r visualize-myeloid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Cell Lineages")

p1 + p2
```

## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.5,
                                  logfc.threshold = 0.5)

write.csv(cluster_markers, file.path(results_dir, "myeloid_subcluster_markers.csv"), row.names = FALSE)
```

```{r microglia-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Sparc", "Hexb", "Csf1r", "Tmem119"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Microglia Core")

FeaturePlot_scCustom(myeloid_obj, features = c("Il1a", "Tnf", "Ccl3", "Cd83"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Activated microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Apoe", "Tyrobp", "C1qa", "Lpl"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Phagocytic microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Ifit3", "Isg15", "Rsad2", "Irf7"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Interferon microglia")

FeaturePlot_scCustom(myeloid_obj, features = c("Hspa1a", "Hspa1b", "Dnaja1", "Hsp90aa1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Stress response")
```

```{r macrophage-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Lyz2", "Ms4a7", "Cd14", "Ctss"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Macrophage Core")

FeaturePlot_scCustom(myeloid_obj, features = c("Gpnmb", "Fabp5", "Plin2", "Lipa"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Lipid-associated TAM")

FeaturePlot_scCustom(myeloid_obj, features = c("Cd74", "H2-Eb1", "H2-Ab1", "Ccr2"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Antigen presentation")

FeaturePlot_scCustom(myeloid_obj, features = c("Mrc1", "Dab2", "Gas6", "Cd163"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Perivascular macrophages")

FeaturePlot_scCustom(myeloid_obj, features = c("Plac8", "Msrb1", "S100a4", "Alox5ap"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Monocytes")

FeaturePlot_scCustom(myeloid_obj, features = c("Ifit2", "Ifit3", "Rsad2", "Isg15"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + plot_annotation(title = "Interferon-producing macrophages")
```

## Reannotate clusters by their expression profiles
```{r reannotate-clusters}
new_cluster_ids <- c(
  "0" = "Early activated microglia",
  "1" = "TAM-like phagocytic microglia",
  "2" = "Stress-response microglia",
  "3" = "Stress-response microglia",
  "4" = "Inflammatory activated microglia",
  "5" = "Angiogenic TAMs",
  "6" = "Antigen-presenting TAMs",
  "7" = "Homeostatic microglia",
  "8" = "Perivascular macrophages",
  "9" = "Lipid-associated TAMs",
  "10" = "Perivascular macrophages (chemokine-producing)",
  "11" = "Antigen-presenting monocytes",
  "12" = "Inflammatory TAMs",
  "13" = "Homeostatic microglia",
  "14" = "cDC1",
  "15" = "Inflammatory monocytes"
)

myeloid_obj$myeloid_subtype <- plyr::mapvalues(
  x = myeloid_obj$seurat_clusters,
  from = names(new_cluster_ids),
  to = new_cluster_ids
)
```

```{r umap-visualization, fig.height=6, fig.width=9}
DimPlot(myeloid_obj, reduction = "umap", group.by = "myeloid_subtype") + 
  ggtitle("Myeloid Subtypes")

ggsave(file.path(plot_dir, "myeloid_subtypes_umap.png"), width = 9, height = 6)
```

## Marker genes on subsets
```{r subset-markers, message=FALSE, warning=FALSE}
subtype_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                 group.by = "myeloid_subtype",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.5,
                                  logfc.threshold = 0.5)

write.csv(subtype_markers, file.path(results_dir, "myeloid_subtype_markers.csv"), row.names = FALSE)
```

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- subtype_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(myeloid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high, order = FALSE) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

```{r visualize-top-markers-heatmap, fig.height=16, fig.width=8}
subtype_markers <- read.csv(file.path(results_dir, "myeloid_subtype_markers.csv"))

top50_markers <- subtype_markers %>%
  group_by(cluster) %>%
  top_n(n = 50, wt = avg_log2FC)

# scale top markers in RNA assay
myeloid_obj <- ScaleData(myeloid_obj, features = unique(top50_markers$gene), assay = "RNA")

# separate heatmap for each cluster, save in one pdf file
pdf(file = file.path(plot_dir, "myeloid_subtype_heatmap_top_markers.pdf"), width = 16, height = 8)
for (lineage in unique(top50_markers$cluster)) {
  markers <- top50_markers %>%
    filter(cluster == lineage) %>%
    pull(gene)
  
  p <- DoHeatmap(myeloid_obj, features = markers, group.by = "myeloid_subtype", assay = "RNA", cells = 1:2000) +
    ggtitle(paste("Top Markers for", lineage)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  print(p)
}
dev.off()
```

## Composition analysis using sccomp

```{r visualize-split-umap, fig.height=6, fig.width=14}
DimPlot(myeloid_obj, reduction = "umap", group.by = "myeloid_subtype", split.by = "condition", ncol = 3) + 
  ggtitle("Myeloid Subclusters by Condition")

ggsave(file.path(plot_dir, "myeloid_subtypes_by_condition_umap.png"), width = 16, height = 8)
```

```{r composition-analysis-sccomp, message=FALSE, warning=FALSE}
sccomp_input <- myeloid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = myeloid_subtype,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('tumor', 'STOP', '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result = 
  sccomp_input |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r visualize-sccomp-results, fig.height=6, fig.width=12}
sccomp_result |> 
  plot_1D_intervals()
ggsave(file.path(plot_dir, "myeloid_subtype_composition_sccomp.png"), width = 12, height = 6)
```

``` {r sccomp-pairwise-fc}
sccomp_result  |> 
   sccomp_proportional_fold_change(
     formula_composition = ~  condition,
     from =  "tumor", 
     to = "B7H3"
    ) |> 
  select(cell_group, statement)
```

## plot relative proportions of clusters between conditions
```{r plot-relative-proportions}
## stacked barplot
prop_df <- myeloid_obj@meta.data %>%
  count(condition, myeloid_subtype) %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(prop_df, aes(x = condition, y = prop, fill = myeloid_subtype)) +
  geom_col(position = position_dodge(width = 0.8)) +
  ylab("Proportion") +
  theme_classic()

ggsave(file.path(plot_dir, "myeloid_subtype_proportions_barplot.png"), width = 12, height = 6)

## dotplot of proportions
ggplot(prop_df, aes(x = condition, y = myeloid_subtype, size = prop, color = prop)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_y_discrete(limits = rev) + 
  theme_classic() +
  ylab("Cluster") + 
  xlab("Condition")

ggsave(file.path(plot_dir, "myeloid_subtype_proportions_dotplot.png"), width = 12, height = 6)
```

# Gene set enrichment analysis

## AUCell scoring on MSigDB Hallmark Gene sets
```{r get-genesets, message=FALSE, warning=FALSE}
# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c5 onology
msigdb_c5 <- msigdbr::msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c7 immune
msigdb_c7 <- msigdbr::msigdbr(species = "Mus musculus", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by myeloid subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(myeloid_obj, assay = "RNA", slot = "data"), nCores = 4, plotStats = TRUE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)
```

```{r visualize-hallmark-activity, fig.height=20, fig.width=18}
# add hallmark scores to seurat object
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
myeloid_obj <- AddMetaData(myeloid_obj, metadata = hallmark_scores_df)
# visualize selected hallmark pathways
hallmark_pathways_to_plot <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                "HALLMARK_COAGULATION",
                                "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                "HALLMARK_COMPLEMENT",
                               "HALLMARK_GLYCOLYSIS",
                               "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",
                               "HALLMARK_E2F_TARGETS",
                               "HALLMARK_MYC_TARGETS_V1")
FeaturePlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, reduction = "umap",
                     pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_plasma_light_high) + 
  plot_annotation(title = "Hallmark Pathway Activity in Myeloid Subtypes")


VlnPlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, group.by = "myeloid_subtype",
                 pt.size = 0, log = FALSE, colors_use = viridis::viridis(14)) + 
  plot_annotation(title = "Hallmark Pathway Activity by Myeloid Subtype")
```

## Subtype vs pathway heatmap of AUCell scores
```{r subtype-pathway-heatmap, fig.height=8, fig.width=16}
# z scores of AUCell scores across clusters per gene set
hallmark_scores_df$cluster = myeloid_obj$myeloid_subtype

avg_hallmark_scores <- hallmark_scores_df %>%
  group_by(cluster) %>%
  summarise(across(starts_with("HALLMARK"), mean)) %>%
  column_to_rownames("cluster")

hallmark_z_scores <- scale(avg_hallmark_scores)


# Color scale
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)
breaks <- seq(-3, 3, length.out = 101)

# save heatmap as pdf
pdf(file = file.path(plot_dir, "myeloid_subtype_hallmark_auc_heatmap.pdf"), width = 16, height = 8)

hm <- pheatmap::pheatmap(mat = hallmark_z_scores,
                   color = colors.use,
                   border_color = "white",
                   breaks = breaks,
                   clustering_distance_rows = "correlation",
                  clustering_distance_cols = "correlation",
                  main = "Hallmark Pathway AUCell Activity across Myeloid Subtypes")
print(hm)
dev.off()
```

## GSEA analysis of Hallmark gene sets
```{r gsea-hallmark}
# get foldchanges for each gene per cluster
clusters <- levels(myeloid_obj$myeloid_subtype)
fc_list <- lapply(clusters, function(cl){
  print(cl)
  df <- FoldChange(
    object = myeloid_obj,
    group.by = "myeloid_subtype",
    ident.1 = cl,
    ident.2 = NULL,
    assay = "RNA",
    slot = "data"
  )
  df <- as.data.frame(df)
  df$gene <- rownames(df)

  # Filter by pct.1
  df <- df %>% filter(pct.1 >= 0.05)
  
  df$cluster <- cl
  df
})
names(fc_list) <- clusters  

# rank genes by product of log FCs and -log10 p values
make_rank_vector <- function(df) {
  stat <- df$avg_log2FC
  names(stat) <- df$gene
  stat[is.finite(stat)]
}

fgsea_list_hallmark <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))

  fgsea(
    pathways = msigdb_hallmark,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_hallmark <- imap_dfr(fgsea_list_hallmark, ~ mutate(.x, cluster = .y))
fgsea_res_hallmark_sig <- fgsea_res_hallmark %>% 
  filter(padj < 0.1 & NES > 0) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_hallmark_sig, file.path(results_dir, "myeloid_subtype_hallmark_gsea_results.csv"), row.names = FALSE)
```

## Visualize select significant genesets
```{r visualize-gsea-results, fig.height=8, fig.width=12}
inflam_TAM_ranks <- make_rank_vector(fc_list[["Inflammatory TAMs"]])
phag_microglia_ranks <- make_rank_vector(fc_list[["TAM-like phagocytic microglia"]])

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_OXIDATIVE_PHOSPHORYLATION"]], phag_microglia_ranks) +
  ggtitle("TAM-like phagocytic microglia: Oxidative Phosphorylation")

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]], inflam_TAM_ranks) +
  ggtitle("Inflammatory TAMs: Interferon Gamma Response")
```

## GSEA on C5 GO gene sets
```{r gsea-c5}
fgsea_list_c5 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c5,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c5 <- imap_dfr(fgsea_list_c5, ~ mutate(.x, cluster = .y))
fgsea_res_c5_sig <- fgsea_res_c5 %>% 
  filter(padj < 0.1) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c5_sig, file.path(results_dir, "myeloid_subtype_c5_gsea_results.csv"), row.names = FALSE)
```

## GSEA on C7 immune gene sets
```{r gsea-c7}
fgsea_list_c7 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c7,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c7 <- imap_dfr(fgsea_list_c7, ~ mutate(.x, cluster = .y))
fgsea_res_c7_sig <- fgsea_res_c7 %>% 
       filter(padj < 0.1) %>%
       arrange(cluster, desc(NES)) %>%
       select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c7_sig, file.path(results_dir, "myeloid_subtype_c7_gsea_results.csv"), row.names = FALSE)
```

# Summary of the myeloid subtypes
AI-assisted summerization of the full marker lists for each myeloid subtype:

### Homeostatic  Microglia
- Key markers: Tmem119, Siglech, Sparc, Sall1 
- Enriched in tumor and STOP samples
- Canonical homeostatic microglia maintaining tissue surveillance and baseline immune functions in the brain. These cells express core microglial identity genes and show minimal activation signatures, representing a quiescent resident microglial population.
- Literature support: 
  - Butovsky et al., Nat Neurosci. 2014: "Identification of a unique TGF-β–dependent molecular and functional signature in microglia"
  - Bennett et al., Nat Neurosci. 2016: "New tools for studying microglia in the mouse and human CNS"
  - Konishi et al., Glia. 2017: "Siglec-H is a microglia-specific marker that discriminates microglia from CNS-associated macrophages and CNS-infiltrating monocytes"

### Stress-Response Microglia
- Key markers: Hspa1a/b, Hsp90aa1, Txnip
- Depleted in B7H3 samples
- Microglia retaining a homeostatic identity but exhibiting a stress-response transcriptional program, marked by heat-shock and oxidative stress genes. Likely reflects microglia responding to tumor-associated stress, hypoxia, or tissue processing rather than a distinct lineage state.
- Literature support:
  - Dukay et al., Front. Pharmacol. 2019: Heat-Shock Proteins in Neuroinflammation
  - Sequeira & Bolton, Endocrinology. 2023: Stressed Microglia: Neuroendocrine–Neuroimmune Interactions in the Stress Response

### Early Activated Microglia
- Key markers: Tnf, Il1a, Nfkbia, Junb, Egr1, Icam1, Cd83
- Similar across conditions
- Acutely activated microglia characterized by immediate-early gene expression and NFκB signaling. Represents a rapid-response microglial state likely induced by tumor-derived cues or CAR-T–associated inflammatory signals.
- Literature support: 
  - Gao et al., Signal Transduction and Targeted Therapy. 2023: Microglia in neurodegenerative diseases: mechanism and potential therapeutic targets
  - Chen et al., Blood. 2009: Identification of early growth response protein 1 (EGR-1) as a novel target for JUN-induced apoptosis in multiple myeloma

### Inflammatory Activated Microglia
- Key markers: Nlrp3, Il1a, Nfkb1, Jak1, Stat3, Trib1, Icam1
- Similar across conditions
- A more sustained inflammatory microglial state marked by inflammasome components and cytokine signaling pathways. These cells likely participate in chronic neuroinflammation within the tumor microenvironment.
- Literature support: 
  - Xu et al., Cellular & Molecular Immunology. 2025: NLRP3 inflammasome in neuroinflammation and central nervous system diseases
  - Zheng et al., Stroke and Vascular Neurology. 2022: Novel role of STAT3 in microglia-dependent neuroinflammation after experimental subarachnoid haemorrhage

### TAM-like Phagocytic Microglia
- Key markers: Apoe, Cst7, C1qa/b/c, Axl, Cd9, Grn
- Similar across conditions
- Microglia adopting a tumor-associated macrophage–like phenotype with enhanced phagocytic and lysosomal activity. This population reflects disease-associated microglia engaged in debris clearance and tumor interaction.
- Literature support:
  - Tran et al., Molecular Neurodegeneration. 2025: APOE Christchurch enhances a disease-associated microglial response to plaque but suppresses response to tau pathology
  - Huang et al., Nat. Immunol. 2021: Microglia use TAM receptors to detect and engulf amyloid beta plaques
  - Wang et al., iScience. 2024: Microglial apolipoprotein E particles contribute to neuronal dysfunction by impairing neurite growth and synapse integrity and promoting neuronal senescence

### Angiogenic Tumor-Associated Macrophages (TAMs)
- Key markers: Vegfa, Thbs1, Tgm2, Plaur, Cd44, Vim
- Enriched across CAR-T designs
- Tumor-associated macrophages specialized for angiogenesis and tissue remodeling. These cells likely localize to hypoxic or perivascular tumor regions and contribute to vascular remodeling and tumor progression.
- Literature support:
  - Lee et al., Signal Transduction and Targeted Therapy. 2025: Vascular endothelial growth factor signaling in health and disease: from molecular mechanisms to therapeutic perspectives
  - Cao et al., Am. J. Pathol. 2006: Involvement of endothelial CD44 during in vivo angiogenesis
  - Kaur et al., Int. J. Mol. Sci. 2021: Functions of thrombospondin-1 in the tumor microenvironment

### Antigen-Presenting TAMs
- Key markers: Cd74, H2-Aa, H2-Ab1, Ciita, Ms4a7, Lyz2
- Enriched across CAR-T designs
- TAMs with strong antigen-presentation capacity, expressing MHC class II machinery alongside macrophage lineage markers. These cells may serve as intermediaries between innate and adaptive immune responses within the tumor.
- Literature support:
  - Polania et al., Immunity. 2025: Antigen presentation by tumor-associated macrophages drives T cells from a progenitor exhaustion state to terminal exhaustion
  - Muraoka et al., J. Clin. Invest. 2019: Antigen delivery targeted to tumor-associated macrophages overcomes tumor immune resistance

### Inflammatory TAMs
- Key markers: C3, Cfb, Clec4e, Vegfa, Tgfbi, Slamf7
- Enriched across CAR-T designs, most strongly in CD8-41BB samples
- Inflammatory TAMs characterized by complement activation, innate immune receptors, interferon production, and tumor-adaptive outputs. This population likely represents macrophages responding to chronic inflammatory and necrotic signals in the tumor microenvironment.
- Literature support:
  - Ma et al., Trends in Immunology. 2022: Macrophage diversity in cancer revisited in the era of single-cell omics
  - Peng et al., Theranostics. 2022: TGFBI secreted by tumor-associated macrophages promotes glioblastoma stem cell-driven tumor growth via integrin αvβ5-Src-Stat3 signaling
  - Suzuki et al., J. Thorac. Oncol. 2025: Tumor-Derived Complement Factor B Drives Tumor Growth and Anti-PD-1 Resistance in STK11-Mutant Lung Adenocarcinoma
  
### Lipid-associated TAMs
- Key markers: Fabp4, Fabp5, Lpl, Gpnmb, Acp5, Atp6v0d2
- Enriched across CAR-T designs
- Macrophages specialized for lipid metabolism and phagocytosis, resembling lipid-associated or foamy macrophages. These cells may arise in response to cell death and lipid-rich debris within tumors.
- Literature support:
  - Timperi et al., Cancer Research. 2022: Lipid-Associated Macrophages Are Induced by Cancer-Associated Fibroblasts and Mediate Immune Suppression in Breast Cancer
  - Yalcin et al., Acta Neuropathologica Communications. 2024: Tumor-associated microglia/macrophages utilize GPNMB to promote tumor growth and alter immune cell infiltration in glioma

### Perivascular Macrophages
- Key markers: Mrc1, Cd163, F13a1, Stab1, Pf4, Dab2
- Depleted across CAR-T designs
- Macrophages associated with the vasculature, displaying immunoregulatory and tissue-supportive features. These cells are often linked to blood–brain barrier maintenance and immune suppression in tumors.
- Literature support:
  - Zhan et al., Cell Death and Disease. 2025: Perivascular macrophages in the central nervous system: insights into their roles in health and disease
  - Wen et al., Brain. 2024: Brain perivascular macrophages: current understanding and future prospects

### Perivascular Macrophages (Chemokine producing)
- Key markers: Ccl2, Ccl7, Cxcl2, Mrc1, Cd163, Pf4
- Depleted across CAR-T designs
- A chemokine-enriched subset of perivascular macrophages that actively recruit additional immune cells. This population may amplify myeloid infiltration and inflammation, particularly under CAR-T–induced immune pressure.
- Literature support:
  - Kim et al., Blood Advances. 2024: Increased CCL2/CCR2 axis promotes tumor progression by increasing M2 macrophages in MYC/BCL2 double-expressor DLBCL
  - Nie et al., Cell Death & Disease. 2024: The role of CXCL2-mediated crosstalk between tumor cells and macrophages in Fusobacterium nucleatum-promoted oral squamous cell carcinoma progression

### Antigen-Presenting Monocytes
- Key markers: Ciita, Cd74, H2-Aa, H2-Ab1, H2-Eb1, Lsp1
- Peripheral monocytes exhibiting a strong antigen-presentation program. These cells likely represent monocytes transitioning toward macrophage differentiation within the tumor while retaining a distinct APC-like activation state.

### Inflammatory Monocytes
- Key markers: Ccr2, S100a8, S100a9, Plac8
- Recruited inflammatory monocytes characterized by chemokine receptor expression and interferon- or stress-associated programs. These cells represent newly infiltrating peripheral myeloid cells responding to tumor and CAR-T–associated inflammation.

### cDC1 (Cross-Presenting Dendritic Cells)
- Key markers: Xcr1, Clec9a, Wdfy4, Itgae, Btla
- Contamination 

## Save final myeloid object
```{r save-final-object}
saveRDS(myeloid_obj, file.path(results_dir, "cart_myeloid_subtypes.rds"))
```

## Session Info
```{r session-info}
sessionInfo()
```
  
