---
title: "04-myeloid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(presto)
library(decoupleR)
library(AUCell)
library(msigdbr)
library(sccomp)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v2")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(results_dir, "cart_myeloid_subclusters.rds"))
```

## Visualize myeloid subclusters
```{r visualize-myeloid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Cell Lineages")

p1 + p2
```
## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.5,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")
```

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(myeloid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

```{r other-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Lyz2", "Apoc2", "Ifngr2", "Irf7"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Inflammation")

FeaturePlot_scCustom(myeloid_obj, features = c("Arg1", "Ccl8", "Spp1", "Gpnmb"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Immune regulation")

FeaturePlot_scCustom(myeloid_obj, features = c("Arg1", "Mrc1", "Chil3", "Fn1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Immunosuppressive")

FeaturePlot_scCustom(myeloid_obj, features = c("Cxcl2", "Thbs1", "Pf4", "Isg15"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Proinflammatory")
```

## Composition analysis using sccomp

```{r visualize-split-umap, fig.height=8, fig.width=12}
DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3, label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Subclusters by Condition")
```

```{r composition-analysis-sccomp, message=FALSE, warning=FALSE}
sccomp_input <- myeloid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = seurat_clusters,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('tumor', 'STOP', '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result = 
  sccomp_input |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r visualize-sccomp-results}
sccomp_result |> plot()
```
``` {r sccomp-pairwise-fc}
sccomp_result  |> 
   sccomp_proportional_fold_change(
     formula_composition = ~  condition,
     from =  "tumor", 
     to = "B7H3"
    ) |> 
  select(cell_group, statement)
```
## Pathway and geneset analysis using decoupler
```{r get-genesets, message=FALSE, warning=FALSE}
# progeny
progeny_pathways <- decoupleR::get_progeny(organism = "mouse", top = 500)

# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# msigdb c7 (immunologic signatures)
msigdb_c7 <- msigdbr::msigdbr(species = "Mus musculus", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by myeloid subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(myeloid_obj, assay = "RNA", slot = "data"), nCores = 4, plotStats = TRUE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)
hallmark_thresholds <- AUCell_exploreThresholds(hallmark_AUC, plotHist=TRUE, assign=TRUE) 
```

```{r visualize-hallmark-activity, fig.height=10, fig.width=16}
# add hallmark scores to seurat object
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
myeloid_obj <- AddMetaData(myeloid_obj, metadata = hallmark_scores_df)
# visualize selected hallmark pathways
hallmark_pathways_to_plot <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                "HALLMARK_COAGULATION",
                                "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                "HALLMARK_COMPLEMENT",
                               "HALLMARK_GLYCOLYSIS",
                               "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",
                               "HALLMARK_E2F_TARGETS",
                               "HALLMARK_MYC_TARGETS_V1")
FeaturePlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, reduction = "umap",
                     pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_plasma_light_high) + 
  plot_annotation(title = "Hallmark Pathway Activity in Myeloid Subclusters")


VlnPlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, group.by = "seurat_clusters",
                 pt.size = 0, log = FALSE, colors_use = viridis::viridis(8)) + 
  plot_annotation(title = "Hallmark Pathway Activity by Myeloid Subcluster")
```

## DE genes analysis between conditions within each myeloid subcluster
Use pseudobulk + EdgeR as it is more statistically sound compared to the default Wilcox test. 

```{r pseudobulk-edgeR-de-analysis}
# extract raw counts and meta data
counts <- GetAssayData(myeloid_obj, assay = "RNA", slot = "counts")
meta <- myeloid_obj@meta.data
meta$cluster <- paste0("c", meta$seurat_clusters)
samples_df <- meta %>% distinct(.data[['sample']], .data[['condition']])

pseudobulk_by_cluster <- function(counts, cell_meta,
                                  cluster_col = "cluster",
                                  sample_col = "sample_id") {
  # Make a combined grouping label: "sample_cluster"
  group_ids <- paste(cell_meta[[sample_col]], cell_meta[[cluster_col]], sep = "__")
  
  # For each group, sum counts across its cells
  # This uses Matrix's efficient rowSums on sparse matrices
  groups <- split(seq_along(group_ids), group_ids)
  
  pb_list <- lapply(groups, function(idx) {
    if (length(idx) == 1L) {
      counts[, idx, drop = FALSE]
    } else {
      Matrix::rowSums(counts[, idx, drop = FALSE])
    }
  })
  
  pb_mat <- do.call(cbind, pb_list)
  colnames(pb_mat) <- names(groups)
  round(pb_mat)
}

pb_counts <- pseudobulk_by_cluster(counts, meta,
                                   cluster_col = "cluster",
                                   sample_col = "sample")
dim(pb_counts)



pb_samples <- data.frame(sample_cluster = colnames(pb_counts)) |>
  tidyr::separate(sample_cluster, into = c("sample", "cluster"), sep = "__")

# Map condition (and other covariates) from original cell-level metadata
sample_level <- meta %>%
  dplyr::select(sample, condition) %>%
  distinct()

pb_samples <- pb_samples %>%
  left_join(sample_level, by = "sample")

head(pb_samples)

run_edger_multi_condition <- function(cluster_id,
                                      pb_counts,
                                      pb_samples,
                                      condition_col = "condition",
                                      min_reps = 2,
                                      do_pairwise = TRUE) {
  # subset to cluster
  keep_lib   <- pb_samples$cluster == cluster_id
  sub_counts <- pb_counts[, keep_lib, drop = FALSE]
  sub_samples <- pb_samples[keep_lib, , drop = FALSE]
  
  # need ≥2 conditions
  cond <- factor(sub_samples[[condition_col]])
  if (length(levels(cond)) < 2L) {
    message("Cluster ", cluster_id, ": only one condition present, skipping.")
    return(NULL)
  }
  # need enough reps per condition
  if (any(table(cond) < min_reps)) {
    message("Cluster ", cluster_id, ": not enough replicates per condition, skipping.")
    return(NULL)
  }
  
  # DGEList
  y <- DGEList(counts = sub_counts, samples = sub_samples)
  
  # Filter lowly expressed genes using full design
  design <- model.matrix(~ 0 + cond)
  colnames(design) <- levels(cond)   # nicer names
  
  keep_genes <- filterByExpr(y, design = design)
  y <- y[keep_genes, , keep.lib.sizes = FALSE]
  
  # Normalization + dispersion
  y <- calcNormFactors(y)
  y <- estimateDisp(y, design)
  fit <- glmQLFit(y, design)
  
  out_list <- list()
  
  ## 3a. Overall test: any condition difference (F-test)
  overall_test <- glmQLFTest(fit, coef = 1:ncol(design))
  overall_res <- topTags(overall_test, n = Inf)$table
  overall_res$gene    <- rownames(overall_res)
  overall_res$cluster <- cluster_id
  overall_res$contrast <- "overall"
  out_list[["overall"]] <- overall_res
  
  ## 3b. Pairwise contrasts between conditions
  if (do_pairwise && length(levels(cond)) >= 2L) {
    levs <- levels(cond)
    for (i in seq_along(levs)) {
      for (j in seq_along(levs)) {
        if (j <= i) next
        c1 <- levs[j]
        c0 <- levs[i]
        contrast_vec <- rep(0, length(levs))
        names(contrast_vec) <- levs
        contrast_vec[c1] <-  1
        contrast_vec[c0] <- -1
        
        qlf <- glmQLFTest(fit, contrast = contrast_vec)
        res <- topTags(qlf, n = Inf)$table
        res$gene     <- rownames(res)
        res$cluster  <- cluster_id
        res$contrast <- paste0(c1, "_vs_", c0)
        out_list[[paste0(c1, "_vs_", c0)]] <- res
      }
    }
  }
  
  dplyr::bind_rows(out_list)
}
```
