---
title: "04-myeloid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(AUCell)
library(msigdbr)
library(sccomp)
library(pheatmap)
library(fgsea)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v2")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(results_dir, "cart_myeloid_subclusters.rds"))
```

## Visualize myeloid subclusters
```{r visualize-myeloid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(myeloid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Cell Lineages")

p1 + p2
```
## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(myeloid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

write.csv(cluster_markers, file.path(results_dir, "myeloid_subcluster_markers.csv"), row.names = FALSE)
```

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(myeloid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

```{r other-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(myeloid_obj, features = c("Lyz2", "Apoc2", "Ifngr2", "Irf7"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Inflammation")

FeaturePlot_scCustom(myeloid_obj, features = c("Arg1", "Ccl8", "Spp1", "Gpnmb"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Immune regulation")

FeaturePlot_scCustom(myeloid_obj, features = c("Arg1", "Mrc1", "Chil3", "Fn1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Immunosuppressive")

FeaturePlot_scCustom(myeloid_obj, features = c("Cxcl2", "Thbs1", "Pf4", "Isg15"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Proinflammatory")
```

## Composition analysis using sccomp

```{r visualize-split-umap, fig.height=8, fig.width=12}
DimPlot(myeloid_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3, label = TRUE, label.size = 4) + 
  ggtitle("Myeloid Subclusters by Condition")
```

```{r composition-analysis-sccomp, message=FALSE, warning=FALSE}
sccomp_input <- myeloid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = seurat_clusters,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('tumor', 'STOP', '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result = 
  sccomp_input |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r visualize-sccomp-results}
sccomp_result |> plot()
```

``` {r sccomp-pairwise-fc}
sccomp_result  |> 
   sccomp_proportional_fold_change(
     formula_composition = ~  condition,
     from =  "tumor", 
     to = "B7H3"
    ) |> 
  select(cell_group, statement)
```
## plot relative proportions of clusters between conditions
```{r plot-relative-proportions}

## stacked barplot
prop_df <- myeloid_obj@meta.data %>%
  count(condition, seurat_clusters) %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(prop_df, aes(x = condition, y = prop, fill = seurat_clusters)) +
  geom_col(position = position_dodge(width = 0.8)) +
  ylab("Proportion") +
  theme_classic()

## dotplot of proportions
ggplot(prop_df, aes(x = condition, y = seurat_clusters, size = prop, color = prop)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_y_discrete(limits = rev) + 
  theme_classic() +
  ylab("Cluster") + 
  xlab("Condition")
```

# Single Cell gene set analysis

## AUCell scoring on MSigDB Hallmark Gene sets
```{r get-genesets, message=FALSE, warning=FALSE}
# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c5 onology
msigdb_c5 <- msigdbr::msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c7 immune
msigdb_c7 <- msigdbr::msigdbr(species = "Mus musculus", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by myeloid subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(myeloid_obj, assay = "RNA", slot = "data"), nCores = 4, plotStats = TRUE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)
```

```{r visualize-hallmark-activity, fig.height=10, fig.width=16}
# add hallmark scores to seurat object
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
myeloid_obj <- AddMetaData(myeloid_obj, metadata = hallmark_scores_df)
# visualize selected hallmark pathways
hallmark_pathways_to_plot <- c("HALLMARK_INFLAMMATORY_RESPONSE",
                                "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
                                "HALLMARK_COAGULATION",
                                "HALLMARK_INTERFERON_ALPHA_RESPONSE",
                                "HALLMARK_COMPLEMENT",
                               "HALLMARK_GLYCOLYSIS",
                               "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY",
                               "HALLMARK_E2F_TARGETS",
                               "HALLMARK_MYC_TARGETS_V1")
FeaturePlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, reduction = "umap",
                     pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_plasma_light_high) + 
  plot_annotation(title = "Hallmark Pathway Activity in Myeloid Subclusters")


VlnPlot_scCustom(myeloid_obj, features = hallmark_pathways_to_plot, group.by = "seurat_clusters",
                 pt.size = 0, log = FALSE, colors_use = viridis::viridis(8)) + 
  plot_annotation(title = "Hallmark Pathway Activity by Myeloid Subcluster")
```

## Cluster vs pathway heatmap of AUCell scores
```{r cluster-pathway-heatmap, fig.height=8, fig.width=16}
# z scores of AUCell scores across clusters per gene set
hallmark_scores_df$cluster = myeloid_obj$seurat_clusters

avg_hallmark_scores <- hallmark_scores_df %>%
  group_by(cluster) %>%
  summarise(across(starts_with("HALLMARK"), mean)) %>%
  column_to_rownames("cluster")

hallmark_z_scores <- scale(avg_hallmark_scores)


# Color scale
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)
breaks <- seq(-3, 3, length.out = 101)

pheatmap::pheatmap(mat = hallmark_z_scores,
                   color = colors.use,
                   border_color = "white",
                   breaks = breaks,
                   clustering_distance_rows = "correlation",
                  clustering_distance_cols = "correlation") 
```
## GSEA analysis of Hallmark gene sets
```{r gsea-hallmark}
# get foldchanges for each gene per cluster
clusters <- levels(Idents(myeloid_obj))
fc_list <- lapply(clusters, function(cl){
  df <- FoldChange(
    object = myeloid_obj,
    ident.1 = cl,
    ident.2 = NULL,
    assay = "RNA",
    slot = "data"
  )
  df <- as.data.frame(df)
  df$gene <- rownames(df)

  # Filter by pct.1
  df <- df %>% filter(pct.1 >= 0.05)
  
  df$cluster <- cl
  df
})
names(fc_list) <- clusters  

# rank genes by product of log FCs and -log10 p values
make_rank_vector <- function(df) {
  stat <- df$avg_log2FC
  names(stat) <- df$gene
  stat[is.finite(stat)]
}

fgsea_list_hallmark <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))

  fgsea(
    pathways = msigdb_hallmark,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_hallmark <- imap_dfr(fgsea_list_hallmark, ~ mutate(.x, cluster = .y))
fgsea_res_hallmark_sig <- fgsea_res_hallmark %>% 
  filter(padj < 0.1 & NES > 0) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_hallmark_sig, file.path(results_dir, "myeloid_subcluster_hallmark_gsea_results.csv"), row.names = FALSE)
```

Visualize select significant genesets
```{r visualize-gsea-results, fig.height=8, fig.width=12}
c0_ranks <- make_rank_vector(fc_list[["0"]])
c7_ranks <- make_rank_vector(fc_list[["7"]])

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]], c0_ranks) +
  ggtitle("Cluster 0: TNFA Signaling via NFKB")

fgsea::plotEnrichment(msigdb_hallmark[["HALLMARK_INTERFERON_GAMMA_RESPONSE"]], c7_ranks) +
  ggtitle("Cluster 7: Interferon Gamma Response")
```
## GSEA on C5 GO gene sets
```{r gsea-c5}
fgsea_list_c5 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c5,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c5 <- imap_dfr(fgsea_list_c5, ~ mutate(.x, cluster = .y))
fgsea_res_c5_sig <- fgsea_res_c5 %>% 
  filter(padj < 0.1) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c5_sig, file.path(results_dir, "myeloid_subcluster_c5_gsea_results.csv"), row.names = FALSE)
```

## GSEA on C7 immune gene sets
```{r gsea-c7}
fgsea_list_c7 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c7,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c7 <- imap_dfr(fgsea_list_c7, ~ mutate(.x, cluster = .y))
fgsea_res_c7_sig <- fgsea_res_c7 %>% 
       filter(padj < 0.1) %>%
       arrange(cluster, desc(NES)) %>%
       select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c7_sig, file.path(results_dir, "myeloid_subcluster_c7_gsea_results.csv"), row.names = FALSE)
```

# Summary of the 8 myeloid clusters
## Cluster 0: Inflammatory activated microglia
- Key markers:
  - Inflammatory/chemokine: Il1a, Tnf, Ccl3, Icam1
  - Activation markers: Egr1, Kdm6b
  - Core microlia: Sparc, Hexb, Csf1r
- Enriched in tumor and STOP samples

## Cluster 1: Homeostatic/stress-response microglia
- Key markers:
  - Core microlia: Sparc, Hexb, Csf1r
  - Metababolic: Selenop, Sgk1
  - Stress response: Hspa1a, Hspa1b, Hsp90aa1
- Enriched in tumor and STOP samples

## Cluster 2: Complement/phagocytic microglia
- Key markers:
  - Complement: C1qa, C1qb, Tyrobp
  - Lysosomal/phagocytic: Ctsd, Ctsb
- Slightly enriched in CD28-41BB samples

## Cluster 3: TAM (Tumor associated macrophages)
- Key markers:
  - TAM markers: Apoe, Lyz2, Ms4a7, Cd14
- Enriched in 41BB-L and B7H3 samples

## Cluster 4: MHCII+ antigen presenting macrophages
- Key markers:
  - MHCII: H2-Aa, H2-Ab1, Cd74
- Enriched in CD8-41BB samples

## Cluster 5: Lipid-loaded TAMs
- Key markers:
  - Lipid metabolism: Gpnmb, Fabp5, Lipa
  - TAM markers: Lgals3, Lyz2
- Enriched in B7H3 samples

## Cluster 6: Perivascular macrophages
- Key markers:
  - Perivascular markers: Mrc1, Dab2, Gas6
- Enriched in B7H3 and 41BB-L samples

## Cluster 7: Interferon-responsive macrophages
- Key markers:
  - Interferon response: Ifit2, Ifit3, Rsad2, Isg15, Irf7
- Enriched in B7H3 samples

## Session Info
```{r session-info}
sessionInfo()
```
  
