---
title: "05-lymphoid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(AUCell)
library(msigdbr)
library(sccomp)
library(pheatmap)
library(fgsea)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v2")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
lymphoid_obj <- readRDS(file.path(results_dir, "cart_lymphoid_subclusters.rds"))
#lymphoid_obj <- readRDS(file.path(results_dir, "cart_tcell_subclusters.rds"))
```

## Visualize lymphoid subclusters
```{r visualize-lymphoid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(lymphoid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(lymphoid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("lymphoid Cell Lineages")

p1 + p2
```

## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(lymphoid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.5,
                                  logfc.threshold = 0.5)

write.csv(cluster_markers, file.path(results_dir, "nkt_subcluster_markers.csv"), row.names = FALSE)
```

Clusters at the top of UMAP are still driven by myeloid genes

## T cell and macrophage genes
```{r visualize-select-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(lymphoid_obj, features =  c("Cd3d", "Cd4", "Cd8a", "Trac"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
  plot_annotation(title = "T Cell Core")

FeaturePlot_scCustom(lymphoid_obj, features = c("Lyz2", "Ms4a7", "Cd14", "Ctss"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Macrophage Core")
```

## Confirm with module scores to see whether they're really T cells with ambient myeloid RNA
```{r module-scores}
# T cell core program (CD4/CD8 agnostic)
t_genes <- c(
  "Cd3d","Cd3e","Cd3g",
  "Cd4","Cd8a","Cd8b1",
  "Trac","Trbc1","Trbc2",
  "Lck","Lat",
  "Il7r","Ccr7","Tcf7","Lef1",   # naive/memory
  "Gzmb","Prf1","Ifng"           # effector
)

lymphoid_obj <- AddModuleScore(
  lymphoid_obj,
  features = list(t_genes),
  name = "T_sig",
  assay = "RNA",
)

# Generic macrophage / monocyte / TAM-ish
mac_genes <- c(
  "Lyz2","Lyz1","C1qa","C1qb","C1qc",
  "Tyrobp","Aif1","Cst3",
  "Csf1r","Mrc1","Cd68","Cd163",
  "Lgals3","Apoe","Spp1",
  "Ccl3","Ccl4"
)

lymphoid_obj <- AddModuleScore(
  lymphoid_obj,
  features = list(mac_genes),
  name = "Mac_sig",
  assay = "RNA",
)

# Microglia-enriched (CNS context)
microglia_genes <- c(
  "P2ry12","P2ry13","Tmem119",
  "Cx3cr1","Gpr34","Fcrls",
  "Hexb","Trem2","Olfml3",
  "Sall1"
)

lymphoid_obj <- AddModuleScore(
  lymphoid_obj,
  features = list(microglia_genes),
  name = "Mic_sig",
  assay = "RNA",
)

# B cell genes
b_genes <- c(
  "Cd79a","Cd79b","Cd19","Ms4a1",      # BCR complex / CD19 / CD20
  "Cd22","Bank1","Blnk","Pax5", "Igkc"         # B identity / signaling
)

lymphoid_obj <- AddModuleScore(
  lymphoid_obj,
  features = list(b_genes),
  name = "B_sig",
  assay = "RNA",
)
```

## Visualize module scores
```{r visualize-module-scores, fig.height=6, fig.width=8}
FeaturePlot_scCustom(lymphoid_obj, features = c("T_sig1", "Mac_sig1", "Mic_sig1", "B_sig1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
  plot_annotation(title = "Lineage module scores")
```
```{r module-score-violin, fig.height=6, fig.width=8}
VlnPlot(lymphoid_obj, features = c("T_sig1", "Mac_sig1", "Mic_sig1", "B_sig1"), group.by = "seurat_clusters", pt.size = 0, ncol = 2) + 
  plot_annotation(title = "Lineage module scores by cluster")
```

Compared to before resequencing, the contamination is much reduced, but still present.
This suggest that further subsetting is probably not necessary anymore.

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(lymphoid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```


```{r other-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(lymphoid_obj, features = c("Tcf7", "Sell", "Lef1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "Naive-like")

FeaturePlot_scCustom(lymphoid_obj, features = c("S1pr1", "Klf2", "Slfn1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "Quiescent")

FeaturePlot_scCustom(lymphoid_obj, features = c("Cd28", "C1qa", "Lat"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "T-cell activation")

FeaturePlot_scCustom(lymphoid_obj, features = c("Zap70", "Ifitm1", "Ifngr1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "Cytokine signaling")

FeaturePlot_scCustom(lymphoid_obj, features = c("Gzmb", "Gzma", "Prf1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "Cytotoxicity")

FeaturePlot_scCustom(lymphoid_obj, features = c("Pdcd1", "Ctla4", "Lag3", "Havcr2", "Eomes", "Tox"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, order=FALSE, colors_use = viridis_light_high) + plot_annotation(title = "Inhibitory receptors")

FeaturePlot_scCustom(lymphoid_obj, features = c("Foxp3", "Tnfrsf9"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Regulatory CD4")

FeaturePlot_scCustom(lymphoid_obj, features = c("Ifng", "Ccl1", "Ccl4"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Effector CD4")
```

## Reannotate clusters by their expression profiles
```{r reannotate-clusters}
new_cluster_ids <- c(
  "0" = "Heat-shocked T cells",
  "1" = "Myeloid cells",
  "2" = "CD4+ Regulatory T cells",
  "3" = "CD8+ Cytotoxic T cells",
  "4" = "CD4+ Helper T cells",
  "5" = "Proliferating",
  "6" = "iNKT1 cells",
  "7" = "NK cells",
  "8" = "Naive T cells",
  "9" = "Proliferating",
  "10" = "Central memory T cells",
  "11" = "NK cells",
  "12" = "IFN-stimulated T cells",
  "13" = "iNKT17 cells",
  "14" = "iNKT2 cells"
)

lymphoid_obj$tcell_subtype <- plyr::mapvalues(
  x = lymphoid_obj$seurat_clusters,
  from = names(new_cluster_ids),
  to = new_cluster_ids
)

DimPlot(lymphoid_obj, reduction = "umap", group.by = "tcell_subtype") + 
  ggtitle("T Cell Subtypes")
ggsave(filename = file.path(plot_dir, "tcell_subtypes_umap.png"), width = 9, height = 6)
```

## Marker genes on subsets
```{r subset-markers, message=FALSE, warning=FALSE}
subtype_markers <- FindAllMarkers(lymphoid_obj,
                                  assay = "RNA",
                                 group.by = "tcell_subtype",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.5,
                                  logfc.threshold = 0.5)

write.csv(subtype_markers, file.path(results_dir, "tcell_subtype_markers.csv"), row.names = FALSE)
```

```{r visualize-top-markers-heatmap, fig.height=16, fig.width=8}
subtype_markers <- read.csv(file.path(results_dir, "tcell_subtype_markers.csv"))

top50_markers <- subtype_markers %>%
  group_by(cluster) %>%
  top_n(n = 50, wt = avg_log2FC)

# scale top markers in RNA assay
lymphoid_obj <- ScaleData(lymphoid_obj, features = unique(top50_markers$gene), assay = "RNA")

# separate heatmap for each cluster, save in one pdf file
pdf(file = file.path(plot_dir, "tcell_subtype_heatmap_top_markers.pdf"), width = 16, height = 8)
for (lineage in unique(top50_markers$cluster)) {
  markers <- top50_markers %>%
    filter(cluster == lineage) %>%
    pull(gene)
  
  p <- DoHeatmap(lymphoid_obj, features = markers, group.by = "tcell_subtype", assay = "RNA", cells = 1:2000) +
    ggtitle(paste("Top Markers for", lineage)) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  print(p)
}
dev.off()
```

# Composition analysis using sccomp

## Split by condition
```{r visualize-lymphoid-subclusters-by-condition, fig.height=6, fig.width=8}
DimPlot(lymphoid_obj, reduction = "umap", group.by = "tcell_subtype", split.by = "condition", ncol = 3) + 
  ggtitle("T Cell Subclusters by Condition")
ggsave(filename = file.path(plot_dir, "tcell_subtypes_by_condition.png"), width = 16, height = 6)
```

```{r composition-analysis-sccomp_tumor, message=FALSE, warning=FALSE}
sccomp_input_tumor <- lymphoid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = tcell_subtype,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('tumor', 'STOP', '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result_tumor = 
  sccomp_input_tumor |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r composition-analysis-sccomp_STOP, message=FALSE, warning=FALSE}
sccomp_input_STOP <- lymphoid_obj@meta.data |>
  tibble::rownames_to_column("cell") |>
  dplyr::select(
    cell,
    sample = sample,
    cell_group = tcell_subtype,
    condition
  ) %>% mutate(condition = factor(condition,
               levels = c('STOP', "tumor", '41BB-L', 'B7H3', 'CD28-41BB', 'CD8-41BB'))) %>%  
  as_tibble()

sccomp_result_STOP = 
  sccomp_input_STOP |>
  sccomp_estimate( 
    formula_composition = ~ condition, 
    .sample = sample, 
    .cell_group = cell_group,
    bimodal_mean_variability_association  = TRUE,
    cores = 4,
    verbose = FALSE
  ) |> 
  sccomp_test()

```

```{r visualize-sccomp-results, fig.height=6, fig.width=12}
sccomp_result_tumor |> 
  plot_1D_intervals()
ggsave(file.path(plot_dir, "tcell_subtype_composition_sccomp_tumor.png"), width = 12, height = 6)

sccomp_result_STOP |> 
  plot_1D_intervals()
ggsave(file.path(plot_dir, "tcell_subtype_composition_sccomp_STOP.png"), width = 12, height = 6)
```
## plot relative proportions of clusters between conditions
```{r plot-relative-proportions}
## stacked barplot
prop_df <- lymphoid_obj@meta.data %>%
  count(condition, tcell_subtype) %>%
  group_by(condition) %>%
  mutate(prop = n / sum(n))

ggplot(prop_df, aes(x = condition, y = prop, fill = tcell_subtype)) +
  geom_col(position = position_dodge(width = 0.8)) +
  ylab("Proportion") +
  theme_classic()
ggsave(filename = file.path(plot_dir, "tcell_subtype_proportions_barplot.png"), width =9, height = 6)

## dotplot of proportions
ggplot(prop_df, aes(x = condition, y = tcell_subtype, size = prop, color = prop)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_y_discrete(limits = rev) + 
  theme_classic() +
  ylab("Cluster") + 
  xlab("Condition")
ggsave(filename = file.path(plot_dir, "tcell_subtype_proportions_dotplot.png"), width =9, height = 6)
```

# Gene set enrichment analysis

## AUCell scoring on MSigDB Hallmark Gene sets
```{r get-genesets, message=FALSE, warning=FALSE}
# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c5 onology
msigdb_c5 <- msigdbr::msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()

# c7 immune
msigdb_c7 <- msigdbr::msigdbr(species = "Mus musculus", category = "C7") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by myeloid subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(lymphoid_obj, assay = "RNA", slot = "data"), nCores = 4, plotStats = TRUE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)
```

## Subtype vs pathway heatmap of AUCell scores
```{r subtype-pathway-heatmap, fig.height=8, fig.width=16}
# z scores of AUCell scores across clusters per gene set
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
hallmark_scores_df$cluster = lymphoid_obj$tcell_subtype

avg_hallmark_scores <- hallmark_scores_df %>%
  group_by(cluster) %>%
  summarise(across(starts_with("HALLMARK"), mean)) %>%
  column_to_rownames("cluster")

hallmark_z_scores <- scale(avg_hallmark_scores)


# Color scale
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)
breaks <- seq(-3, 3, length.out = 101)

pdf(file = file.path(plot_dir, "tcell_subtype_hallmark_auc_heatmap.pdf"), width = 16, height = 8)
hm <- pheatmap::pheatmap(mat = hallmark_z_scores,
                   color = colors.use,
                   border_color = "white",
                   breaks = breaks,
                   clustering_distance_rows = "correlation",
                  clustering_distance_cols = "correlation") 
print(hm)
dev.off()
```

## GSEA analysis of Hallmark gene sets
```{r gsea-hallmark}
# get foldchanges for each gene per cluster
clusters <- levels(lymphoid_obj$tcell_subtype)
fc_list <- lapply(clusters, function(cl){
  df <- FoldChange(
    object = lymphoid_obj,
    group.by = "tcell_subtype",
    ident.1 = cl,
    ident.2 = NULL,
    assay = "RNA",
    slot = "data"
  )
  df <- as.data.frame(df)
  df$gene <- rownames(df)

  # Filter by pct.1
  df <- df %>% filter(pct.1 >= 0.05)
  
  df$cluster <- cl
  df
})
names(fc_list) <- clusters  

# rank genes by product of log FCs and -log10 p values
make_rank_vector <- function(df) {
  stat <- df$avg_log2FC
  names(stat) <- df$gene
  stat[is.finite(stat)]
}

fgsea_list_hallmark <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))

  fgsea(
    pathways = msigdb_hallmark,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_hallmark <- imap_dfr(fgsea_list_hallmark, ~ mutate(.x, cluster = .y))
fgsea_res_hallmark_sig <- fgsea_res_hallmark %>% 
  filter(padj < 0.1 & NES > 0) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_hallmark_sig, file.path(results_dir, "tcell_subtype_hallmark_gsea_results.csv"), row.names = FALSE)
```

## GSEA on C5 GO gene sets
```{r gsea-c5}
fgsea_list_c5 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c5,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c5 <- imap_dfr(fgsea_list_c5, ~ mutate(.x, cluster = .y))
fgsea_res_c5_sig <- fgsea_res_c5 %>% 
  filter(padj < 0.1) %>%
  arrange(cluster, desc(NES)) %>%
  select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c5_sig, file.path(results_dir, "tcell_subtype_c5_gsea_results.csv"), row.names = FALSE)
```

## GSEA on C7 immune gene sets
```{r gsea-c7}
fgsea_list_c7 <- map(fc_list, function(df_cluster) {
  stats_vec <- make_rank_vector(df_cluster)
  print(length(stats_vec))
  
  fgsea(
    pathways = msigdb_c7,
    stats    = stats_vec,
    minSize  = 15,
    maxSize  = 500,
  )
})

fgsea_res_c7 <- imap_dfr(fgsea_list_c7, ~ mutate(.x, cluster = .y))
fgsea_res_c7_sig <- fgsea_res_c7 %>% 
       filter(padj < 0.1) %>%
       arrange(cluster, desc(NES)) %>%
       select(cluster, pathway, NES, pval, padj)
write.csv(fgsea_res_c7_sig, file.path(results_dir, "tcell_subtype_c7_gsea_results.csv"), row.names = FALSE)
```

## Save T cell object
```{r save-tcell-object}
saveRDS(lymphoid_obj, file.path(results_dir, "cart_tcell_subclusters.rds"))
```

# Summary of the T cell subtypes
AI-assisted summerization of the full marker lists for each T cell subtype.

### Heat-shock Activated T cells
- Key markers: Hspa1a, Hspa1b, Dnajb1, Jun, Cd69, Ms4a4b, Rasal3, Btg2
- Recently activated or acutely stressed T cells, often induced by tissue dissociation, tumor microenvironment stress, or strong stimulation. Represents a state, not a stable lineage.

### Myeloid cells (contamination)
- Key markers: Csf1r, Tmem119, Hexb, Cst3, Ccl3
- Microglia/macrophage-like cells. Clear non-lymphoid contamination and should be excluded from T/NKT-focused analyses.Potentially central memory T cells, but not clean

### CD4+ Regulatory T cells
- Key markers: Foxp3, Ctla4, Il2ra (CD25), Ikzf2, Tnfrsf4
- Immunosuppressive CD4+ T cells involved in immune tolerance and checkpoint regulation. Likely dampen effector responses within the tissue

### CD8+ Cytotoxic T cells
- Key markers: Cd8a, Cd8b1, Gzmb, Nkg7, Prf1, Pdcd1
- Activated cytotoxic effector T cells with killing capacity. PD-1 expression suggests recent activation and/or early exhaustion.

### CD4+ Helper T cells
- Key markers: Cd4, Icos, Tnfrsf4 (OX40), Bhlhe40
- Activated helper T cells supporting immune coordination and cytokine signaling. Less cytotoxic, more regulatory/organizing in function.

## Early-activation T cells
- Key markers: Egr1, Jun, Fos, Atf3, Zfp36, Dusp1, Hspa1a/b, Klf2
- “Shock response” T cells—very typical after stimulation or tissue dissociation.

## Proliferating
- Key markers: Mki67, Cdk1, Ccnb1, Ccnb2, Top2a
- Actively cycling lymphocytes. Identity is secondary; represents a cell-cycle state shared across multiple T/NKT subsets.

# iNKT1 cells
- Key markers: Eomes, Tbx21, Prf1, Gzma, Nkg7
- Cytotoxic, Th1-like invariant NKT cells. Specialized for IFN-γ production and effector responses against tumors or infections.

# NK cells
- Key markers: Ncr1, Klrb1c, Klrd1, Tyrobp, Fcgr3
- Innate cytotoxic lymphocytes lacking TCR signaling. Functionally distinct from T and iNKT cells, with strong NK receptor programs.

# Naive T cells
- Key markers: Satb1, Txk, Cd7, Tcf7, Lef1
- Quiescent or early-differentiation T cells with minimal activation or effector signatures. Likely precursors to activated or memory states.

# Central memory T cells
- Key markers: Sell (CD62L), S1pr1, Lef1, Foxp1
- Recirculating memory T cells with lymphoid-homing capacity. Less cytotoxic, optimized for long-term immune surveillance.

# IFN-stimulated T cells
- Key markers: Ifit3, Isg15, Oas3, Zbp1
- T cells responding to type I interferon signaling. Reflects cytokine exposure or viral/inflammatory sensing rather than lineage differences.

# iNKT17 cells
- Key markers: Rorc, Il23r, Il1r1, Cxcr6, Zbtb16
- Pro-inflammatory invariant NKT subset associated with IL-17 biology and tissue residency, particularly in barrier or inflamed tissues.

# iNKT2 cells
- Key markers: Gata3, Il17rb, Il1rl1 (ST2), Areg
- Th2-like invariant NKT cells involved in cytokine support and tissue repair. Amphiregulin suggests regenerative or homeostatic roles.


## Session Info
```{r session-info}
sessionInfo()
```
  