---
title: "05-lymphoid-characterization.Rmd"
author: "Sam Chen"
date: "`r Sys.Date()`"
output: 
  html_document:
  toc: TRUE
---

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(tidyverse)
library(SeuratObject)
library(scCustomize)
library(AUCell)
library(msigdbr)
library(sccomp)
library(pheatmap)
library(fgsea)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v2")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Source function
source(file.path(analysis_dir, "util", "run_doubletfinder.R"))
source(file.path(analysis_dir, "util", "run_soupx.R"))

# set random seed
set.seed(1234)
```

## Load Data
```{r load-data}
# Load Seurat object
lymphoid_obj <- readRDS(file.path(results_dir, "cart_lymphoid_subclusters.rds"))
```

## Visualize lymphoid subclusters
```{r visualize-lymphoid-subclusters, fig.height=6, fig.width=12}
p1 <- DimPlot(lymphoid_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) + 
  ggtitle("Subclusters")
p2 <- DimPlot(lymphoid_obj, reduction = "umap", group.by = "cell_type", label = TRUE, label.size = 4) + 
  ggtitle("lymphoid Cell Lineages")

p1 + p2
```

## Cluster marker genes
```{r cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(lymphoid_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

write.csv(cluster_markers, file.path(results_dir, "lymphoid_subcluster_markers.csv"), row.names = FALSE)
```

## Visualize top markers
```{r visualize-top-markers, fig.height=6, fig.width=6}
top_markers_table <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(lymphoid_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

## Check T cell genes
```{r check-t-cell-genes, fig.height=6, fig.width=12}
t_cell_genes <- c("Cd3d", "Cd3e", "Cd4", "Cd8a", "Cd8b1", "Trac", "Trbc2", "Gzmb", "Ifng", "Pdcd1", "Ctla4")
FeaturePlot_scCustom(lymphoid_obj, features = t_cell_genes, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
  plot_annotation(title = "T Cell Marker Genes")
```

## Check macrophage genes
```{r check-macrophage-genes, fig.height=6, fig.width=12}
macrophage_genes <- c("Lyz2", "C1qc", "Csf1r", "Mrc1", "Cd14", "Itgam")
FeaturePlot_scCustom(lymphoid_obj, features = macrophage_genes, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
  plot_annotation(title = "Macrophage Marker Genes")
```
```{r other-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(lymphoid_obj, features = c("Tcf7", "Sell", "Lef1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Naive-like")

FeaturePlot_scCustom(lymphoid_obj, features = c("S1pr1", "Klf2", "Slfn1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Quiescent")

FeaturePlot_scCustom(lymphoid_obj, features = c("Cd28", "C1qa", "Lat"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "T-cell activation")

FeaturePlot_scCustom(lymphoid_obj, features = c("Zap70", "Ifitm1", "Ifngr1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Cytokine signaling")

FeaturePlot_scCustom(lymphoid_obj, features = c("Gzmb", "Gzma", "Prf1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Cytotoxicity")

FeaturePlot_scCustom(lymphoid_obj, features = c("Pdcd1", "Ctla4", "Lag3", "Havcr2", "Eomes", "Tox"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Inhibitory receptors")

FeaturePlot_scCustom(lymphoid_obj, features = c("Foxp3", "Tnfrsf9"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Regulatory CD4")

FeaturePlot_scCustom(lymphoid_obj, features = c("Ifng", "Ccl1", "Ccl4"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Effector CD4")
```


## Further subsetting of the object to only T cells
Upon inspection of top marker genes in each cluster:
- clusters 0, 2, 3, 7, 9 likely represent myeloid cells
- cluster 6 is B cells
- cluster 10 is neurons

```{r subset-t-cells}
tcell_obj <- subset(lymphoid_obj, idents = c(1, 4, 5, 8), invert = FALSE)
```

## Reprocess T cell object
```{r reprocess-tcell-object, message=FALSE, warning=FALSE}
tcell_obj[['RNA']] <- split(tcell_obj[['RNA']], tcell_obj$sample)
tcell_obj <- NormalizeData(tcell_obj)
tcell_obj <- FindVariableFeatures(tcell_obj, selection.method = "vst", nfeatures = 2000)
tcell_obj <- ScaleData(tcell_obj)
tcell_obj <- RunPCA(tcell_obj, npcs = 30)
ElbowPlot(tcell_obj, ndims = 30) # use 15 dims

tcell_obj <- IntegrateLayers(
  object = tcell_obj, method = CCAIntegration,
  orig.reduction = "pca", new.reduction = "integrated.cca", verbose=FALSE, k.weight=30
)
tcell_obj [["RNA"]] <- JoinLayers(tcell_obj[["RNA"]])

tcell_obj <- FindNeighbors(tcell_obj, reduction = "integrated.cca", dims = 1:15)
tcell_obj <- FindClusters(tcell_obj, resolution = 0.5)

tcell_obj <- RunUMAP(tcell_obj, reduction = "integrated.cca", dims = 1:15)
```

## Visualize T cell subclusters
```{r visualize-lymphoid-subclusters, fig.height=6, fig.width=7}
DimPlot(tcell_obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE, label.size = 4) +
  ggtitle("Subclusters")
```

## Split by condition
```{r visualize-lymphoid-subclusters-by-condition, fig.height=6, fig.width=8}
DimPlot(tcell_obj, reduction = "umap", group.by = "seurat_clusters", split.by = "condition", ncol = 3) + 
  ggtitle("T Cell Subclusters by Condition")
```

## Save T cell object
```{r save-tcell-object}
saveRDS(tcell_obj, file.path(results_dir, "cart_tcell_subclusters.rds"))
```

## Cluster marker genes
```{r t-cell-cluster-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(tcell_obj,
                                  assay = "RNA",
                                  slot = "data",
                                  only.pos = TRUE,
                                  min.pct = 0.25,
                                  logfc.threshold = 0.25,
                                  test.use = "MAST")

write.csv(cluster_markers, file.path(results_dir, "tcell_subcluster_markers.csv"), row.names = FALSE)
```

## Visualize top markers
```{r visualize-top-tcell-markers, fig.height=6, fig.width=6}
top_markers_table <- cluster_markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 4) %>%
  ungroup() %>%
  arrange(cluster, desc(avg_log2FC))

for (clust in unique(top_markers_table$cluster)) {
  clust_markers <- top_markers_table %>% filter(cluster == clust) %>% pull(gene)
  p <- FeaturePlot_scCustom(tcell_obj, features = clust_markers, reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + 
    plot_annotation(title = paste0("Cluster ", clust))
  print(p)
}
```

```{r other-markers, fig.height=6, fig.width=6}
FeaturePlot_scCustom(tcell_obj, features = c("Tcf7", "Sell", "Lef1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Naive-like")

FeaturePlot_scCustom(tcell_obj, features = c("S1pr1", "Klf2", "Slfn1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Quiescent")

FeaturePlot_scCustom(tcell_obj, features = c("Cd28", "C1qa", "Lat"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "T-cell activation")

FeaturePlot_scCustom(tcell_obj, features = c("Zap70", "Ifitm1", "Ifngr1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Cytokine signaling")

FeaturePlot_scCustom(tcell_obj, features = c("Gzmb", "Gzma", "Prf1"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Cytotoxicity")

FeaturePlot_scCustom(tcell_obj, features = c("Pdcd1", "Ctla4", "Lag3", "Havcr2", "Eomes", "Tox"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Inhibitory receptors")

FeaturePlot_scCustom(tcell_obj, features = c("Foxp3", "Tnfrsf9"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Regulatory CD4")

FeaturePlot_scCustom(tcell_obj, features = c("Ifng", "Ccl1", "Ccl4"), reduction = "umap", pt.size = 0.1, alpha_exp = 0.5, colors_use = viridis_light_high) + plot_annotation(title = "Effector CD4")
```
