---
title: "11-myeloid-gsea.Rmd"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---

## Overview

This notebook performs **gene set enrichment analysis (GSEA)** in myeloid cells across treatment conditions using MSigDB Hallmark gene sets.

We provide **two complementary workflows**:

1) **Cell-level DE → fgsea (Seurat)**  
   - Uses Seurat `FindMarkers()` (Wilcoxon test) to compute differential expression between conditions at the single-cell level.  
   - Ranks **all genes** by `avg_log2FC` (no expression/FC filtering) and runs `fgsea()` on the ranked list.

2) **Pseudobulk DE → fgsea (DESeq2, recommended when replicates exist)**  
   - Aggregates raw counts per **sample** to create a pseudobulk count matrix (genes × samples).  
   - Runs DESeq2 to obtain a ranked statistic (Wald `stat`) for each contrast.  
   - Runs `fgsea()` on the DESeq2 ranking.  
   - This approach is generally more appropriate for between-condition inference because it respects sample-level replication.

### Contrasts computed

**Cell-level (Seurat):**  

- 41BB-L vs rest (all other conditions)  
- 41BB-L vs other CARs (B7H3 + CD28-41BB + CD8-41BB; excludes tumor/STOP)  
- 41BB-L vs each CAR individually  

**Pseudobulk (DESeq2):**  

- 41BB-L vs rest (all other conditions)  
- 41BB-L vs other CARs (B7H3 + CD28-41BB + CD8-41BB; excludes tumor/STOP)  
- 41BB-L vs each CAR individually  

### Outputs

Results are written to `results_dir` as CSV files:  

- DE tables: `DE_*.csv` (Seurat) and `DESeq2_*.csv` (pseudobulk)  
- GSEA tables: `GSEA_*.csv`  

Plots are written to `plot_dir` as PDF files:  
- Volcano plots: `volcano_*.pdf`  
- GSEA dot plots: `gsea_dot_*.pdf`  

Note: `fgsea` returns list-columns (e.g., `leadingEdge`). These are serialized (semicolon-separated) before writing CSV.  

## Setup
```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(tibble)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(scCustomize)
library(msigdbr)
library(fgsea)
library(Matrix)
library(DESeq2)
library(stringr)
library(EnhancedVolcano)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v4")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# set random seed
set.seed(1234)
```

## Load data and define identities
```{r load-data}
# Load Seurat object
myeloid_obj <- readRDS(file.path(results_dir, "cart_myeloid_subtypes.rds"))

# Make condition a factor with explicit order so contrasts are stable/reproducible
myeloid_obj$condition <- factor(
  myeloid_obj$condition,
  levels = c("tumor","STOP","B7H3","41BB-L","CD8-41BB","CD28-41BB")
)

Idents(myeloid_obj) <- "condition"

# Quick sanity check
table(myeloid_obj$condition)
```

## Load MSigDB Hallmark gene sets

```{r load-msigdb-hallmark}
hallmark_pathways <- msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)

length(hallmark_pathways)
```

## Helper functions

```{r helper-functions}
# Write fgsea results to CSV
write_fgsea_csv <- function(df, out_path) {
  df2 <- df %>%
    mutate(
      leadingEdge = if ("leadingEdge" %in% colnames(df)) {
        vapply(leadingEdge, function(x) paste(x, collapse = ";"), character(1))
      } else {
        NA_character_
      }
    )
  write.csv(df2, out_path, row.names = FALSE)
}

# Volcano plot
plot_volcano <- function(df,
                         gene_col = "gene",
                         lfc_col,
                         p_col,
                         title,
                         subtitle = "",
                         caption = NULL,
                         p_cutoff = 0.05,
                         lfc_cutoff = 1,
                         label_top_n = 15) {
  stopifnot(all(c(gene_col, lfc_col, p_col) %in% colnames(df)))
  
  df2 <- df %>%
    mutate(.p = .data[[p_col]], .lfc = .data[[lfc_col]]) %>%
    filter(is.finite(.p), is.finite(.lfc), !is.na(.p), !is.na(.lfc))
  
  label_df <- df2 %>%
  filter(.p <= p_cutoff, abs(.lfc) >= lfc_cutoff) %>%
  arrange(.p) %>%
  slice_head(n = label_top_n)
  
  lab_vec <- ifelse(df2[[gene_col]] %in% label_df[[gene_col]], df2[[gene_col]], "")
  
  if (is.null(caption)) {
    caption <- paste0("total = ", nrow(df2), " variables")
  }
  
  EnhancedVolcano::EnhancedVolcano(
    df2,
    lab = lab_vec,
    x = lfc_col,
    y = p_col,
    title = title,
    subtitle = subtitle,
    caption = caption,
    pCutoff = p_cutoff,
    FCcutoff = lfc_cutoff,
    legendPosition = "bottom",
    pointSize = 1.5,
    colAlpha = 0.8,
    labSize = 3,
    drawConnectors = TRUE,
    widthConnectors = 0.4,
    colConnectors = "black",
    max.overlaps = Inf
  )
}

# GSEA dotplot (activated vs suppressed) from fgsea output
plot_gsea_dot <- function(gsea_df,
                          title = NULL,
                          p_cutoff = 0.05,
                          top_n = 10,
                          use_p = c("pval", "padj")) {
  use_p <- match.arg(use_p)
  stopifnot(all(c("pathway", "NES", use_p, "size") %in% colnames(gsea_df)))
  
  df <- gsea_df %>%
    mutate(
      p = .data[[use_p]],
      direction = ifelse(NES >= 0, "activated", "suppressed"),
      Count = if ("leadingEdge" %in% colnames(gsea_df)) lengths(leadingEdge) else NA_integer_,
      GeneRatio = if ("leadingEdge" %in% colnames(gsea_df)) (lengths(leadingEdge) / size) else NA_real_,
      pathway_clean = str_remove(pathway, "^HALLMARK_")
    ) %>%
    filter(is.finite(NES), is.finite(p))
  
  df2 <- df %>%
    filter(p <= p_cutoff) %>%
    group_by(direction) %>%
    arrange(p) %>%
    slice_head(n = top_n) %>%
    ungroup()
  
  if (nrow(df2) == 0) {
    message("No pathways with ", use_p, " <= ", p_cutoff, " for ", title, ".")
    return(NULL)
  }
  
  df2 <- df2 %>%
    group_by(direction) %>%
    mutate(pathway_clean = factor(pathway_clean, levels = rev(unique(pathway_clean)))) %>%
    ungroup()
  
  ggplot(df2, aes(x = GeneRatio, y = pathway_clean)) +
    geom_point(aes(size = Count, color = p), alpha = 0.9) +
    facet_grid(. ~ direction, scales = "free_y", space = "free_y") +
    scale_color_gradient(low = "red", high = "blue", trans = "reverse") +
    labs(
      x = "GeneRatio",
      y = NULL,
      size = "Count",
      color = use_p,
      title = title
    ) +
    guides(
      size = guide_legend(order = 1),
      color = guide_colorbar(order = 2)
    ) +
    theme_bw() +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Run Seurat DE (FindMarkers) and fgsea in one step for a given contrast
run_celllevel_fgsea <- function(seu,
                                ident1,
                                ident2,
                                hallmark_pathways,
                                out_prefix,
                                results_dir,
                                plot_dir,
                                minSize = 15,
                                maxSize = 500,
                                volcano_lfc_cutoff = 1,
                                volcano_p_cutoff = 0.05) {
  de <- FindMarkers(
    seu,
    ident.1 = ident1,
    ident.2 = ident2,
    assay = "RNA",
    slot = "data",
    test.use = "wilcox",
    logfc.threshold = 0,
    min.pct = 0
  ) %>%
    tibble::rownames_to_column("gene") %>%
    filter(is.finite(avg_log2FC), !is.na(avg_log2FC)) %>%
    group_by(gene) %>%
    slice_max(avg_log2FC, n = 1, with_ties = FALSE) %>%
    ungroup()
  
  ranked_genes <- setNames(de$avg_log2FC, de$gene)
  ranked_genes <- sort(ranked_genes, decreasing = TRUE)
  
  set.seed(1234)
  gsea <- fgsea(
    pathways = hallmark_pathways,
    stats = ranked_genes,
    minSize = minSize,
    maxSize = maxSize
  ) %>%
    arrange(pval)
  
  # Save tables
  write.csv(de, file.path(results_dir, paste0("DE_", out_prefix, ".csv")), row.names = FALSE)
  write_fgsea_csv(gsea, file.path(results_dir, paste0("GSEA_", out_prefix, "_hallmark.csv")))
  
  de_volcano <- de %>%
    filter(pmax(pct.1, pct.2) >= 0.05)
  
  # Volcano plot
  pvol <- plot_volcano(
    df = de_volcano,
    gene_col = "gene",
    lfc_col = "avg_log2FC",
    p_col = "p_val",
    title = out_prefix,
    p_cutoff = volcano_p_cutoff,
    lfc_cutoff = volcano_lfc_cutoff,
    label_top_n = 15
  )
  ggsave(file.path(plot_dir, paste0("volcano_", out_prefix, ".pdf")),
         pvol, width = 10, height = 7)
  
  # GSEA dotplot (activated vs suppressed)
  pgsea <- plot_gsea_dot(
    gsea_df = gsea,
    title = paste0("GSEA: ", out_prefix),
    top_n = 10,
    use_p = "pval"
  )
  
  if (!is.null(pgsea)) {
    ggsave(file.path(plot_dir, paste0("gsea_dot_", out_prefix, ".pdf")),
           pgsea, width = 8, height = 6)
  }
  
  list(
    de = de,
    gsea = gsea,
    volcano = pvol,
    gsea_dot = pgsea
  )
}

# Run pseudobulk DESeq2 + fgsea for a 2-group contrast defined by factor cond2
run_pseudobulk_fgsea <- function(pb_counts, coldata, cond2_levels,
                                 hallmark_pathways, out_prefix, results_dir, plot_dir,
                                 minSize = 15, maxSize = 500,
                                 volcano_lfc_cutoff = 1, volcano_p_cutoff = 0.05) {
  
  coldata$cond2 <- factor(coldata$cond2, levels = cond2_levels)
  
  dds <- DESeqDataSetFromMatrix(
    countData = round(pb_counts),
    colData   = coldata,
    design    = ~ cond2
  )
  
  # NOTE: no additional per-contrast gene filtering here
  dds <- DESeq(dds)
  
  res <- results(dds, contrast = c("cond2", cond2_levels[2], cond2_levels[1]))
  
  res_df <- as.data.frame(res) %>%
    tibble::rownames_to_column("gene") %>%
    filter(!is.na(stat), is.finite(stat)) %>%
    arrange(padj)
  
  # Save DE table
  write.csv(res_df,
            file.path(results_dir, paste0("DESeq2_", out_prefix, ".csv")),
            row.names = FALSE)
  
  # fgsea ranks: DESeq2 Wald stat
  ranks <- setNames(res_df$stat, res_df$gene)
  ranks <- sort(ranks, decreasing = TRUE)
  
  set.seed(1234)
  gsea <- fgsea(
    pathways = hallmark_pathways,
    stats    = ranks,
    minSize  = minSize,
    maxSize  = maxSize
  ) %>%
    arrange(pval)
  
  write_fgsea_csv(gsea,
                  file.path(results_dir, paste0("GSEA_", out_prefix, "_hallmark.csv")))
  
  # Volcano plot
  pvol <- plot_volcano(
    df = res_df,
    gene_col = "gene",
    lfc_col = "log2FoldChange",
    p_col = "pvalue",
    title = out_prefix,
    p_cutoff = volcano_p_cutoff,
    lfc_cutoff = volcano_lfc_cutoff,
    label_top_n = 15
  )
  ggsave(file.path(plot_dir, paste0("volcano_", out_prefix, ".pdf")),
         pvol, width = 10, height = 7)
  
  # GSEA dotplot: only if significant pathways exist
  pgsea <- plot_gsea_dot(
    gsea_df = gsea,
    title = paste0("GSEA: ", out_prefix),
    p_cutoff = 0.05,
    top_n = 10,
    use_p = "pval"
  )
  
  if (!is.null(pgsea)) {
    ggsave(file.path(plot_dir, paste0("gsea_dot_", out_prefix, ".pdf")),
           pgsea, width = 8, height = 6)
  }
  
  list(de = res_df, gsea = gsea, volcano = pvol, gsea_dot = pgsea)
}
```

## Part I — Cell-level GSEA (Seurat FindMarkers → fgsea)

### 1) 41BB-L vs rest (all other conditions)
This contrast compares 41BB-L against all remaining conditions in the dataset, including tumor and STOP.

```{r 41BBL-vs-rest}
rest_levels <- setdiff(levels(myeloid_obj$condition), "41BB-L")

out_cell_rest <- run_celllevel_fgsea(
  seu = myeloid_obj,
  ident1 = "41BB-L",
  ident2 = rest_levels,
  hallmark_pathways = hallmark_pathways,
  out_prefix = "41BB-L_vs_rest",
  results_dir = results_dir,
  plot_dir = plot_dir,
  volcano_lfc_cutoff = 1,
  volcano_p_cutoff = 0.05
)

out_cell_rest$gsea %>%
  filter(pval < 0.05) %>%
  select(pathway, NES, pval, padj, size) %>%
  head(20) %>%
  print()
```

### 2) 41BB-L vs other CARs only (exclude tumor and STOP)
This is typically the most biologically interpretable CAR-specific contrast if you want to isolate differences between CAR constructs.

```{r 41BBL-vs-other-CARs}
other_cars <- setdiff(levels(myeloid_obj$condition), c("41BB-L", "tumor", "STOP"))

out_cell_othercar <- run_celllevel_fgsea(
  seu = myeloid_obj,
  ident1 = "41BB-L",
  ident2 = other_cars,
  hallmark_pathways = hallmark_pathways,
  out_prefix = "41BB-L_vs_otherCAR",
  results_dir = results_dir,
  plot_dir = plot_dir,
  volcano_lfc_cutoff = 1,
  volcano_p_cutoff = 0.05
)

out_cell_othercar$gsea %>%
  filter(pval < 0.05) %>%
  select(pathway, NES, pval, padj, size) %>%
  head(20) %>%
  print()
```

### 3) 41BB-L vs each CAR individually

This section runs pairwise comparisons:

- 41BB-L vs B7H3
- 41BB-L vs CD28-41BB
- 41BB-L vs CD8-41BB

Pairwise contrasts can be more sensitive than “vs pooled other CARs” when the pooled group is heterogeneous.

```{r 41BBL-vs-each-CAR}
cars_to_compare <- c("B7H3", "CD28-41BB", "CD8-41BB")
gsea_cell_each <- list()

for (car in cars_to_compare) {
  out <- run_celllevel_fgsea(
    seu = myeloid_obj,
    ident1 = "41BB-L",
    ident2 = car,
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("41BB-L_vs_", car),
    results_dir = results_dir,
    plot_dir = plot_dir,
    volcano_lfc_cutoff = 1,
    volcano_p_cutoff = 0.05
  )
  gsea_cell_each[[car]] <- out$gsea %>% mutate(contrast = paste0("41BB-L vs ", car))

  cat("\n==== Cell-level: 41BB-L vs ", car, " ====\n", sep = "")
  print(
    out$gsea %>%
      filter(pval < 0.05) %>%
      select(pathway, NES, pval, padj, size) %>%
      head(10)
  )
}

gsea_cell_each_df <- bind_rows(gsea_cell_each)
write_fgsea_csv(gsea_cell_each_df,
                file.path(results_dir, "GSEA_41BB-L_vs_eachCAR_hallmark.csv"))
```

## Part II — Pseudobulk GSEA (sample-level counts → DESeq2 → fgsea)

### 1) Build pseudobulk counts per sample

This section aggregates raw UMI counts (RNA assay, slot="counts") across all myeloid cells
belonging to the same sample, yielding a genes × samples count matrix.

```{r build-pseudobulk}
stopifnot(all(c("sample","condition") %in% colnames(myeloid_obj@meta.data)))

counts <- GetAssayData(myeloid_obj, assay = "RNA", layer = "counts")

meta <- myeloid_obj@meta.data %>%
  dplyr::select(sample, condition) %>%
  mutate(
    sample = as.character(sample),
    condition = factor(condition, levels = levels(myeloid_obj$condition))
  )

# Aggregate counts by sample (genes x samples)
pb_counts <- Matrix::t(rowsum(Matrix::t(counts), group = meta$sample))

# Sample-level metadata (one row per sample)
coldata <- meta %>%
  as.data.frame() %>%
  { rownames(.) <- NULL; . } %>%      # remove existing rownames (cell barcodes)
  distinct(sample, condition) %>%
  tibble::column_to_rownames("sample")

# Align matrices
pb_counts <- pb_counts[, rownames(coldata)]

# Replicate counts per group
table(coldata$condition)
```

### 2) Pseudobulk: 41BB-L vs rest

This mirrors the cell-level “vs rest” contrast, but at the sample level.

```{r pseudobulk-41BBL-vs-rest}
coldata_pb <- coldata
coldata_pb$cond2 <- ifelse(coldata_pb$condition == "41BB-L", "41BB-L", "rest")

out_pb_rest <- run_pseudobulk_fgsea(
  pb_counts = pb_counts,
  coldata = coldata_pb,
  cond2_levels = c("rest", "41BB-L"),
  hallmark_pathways = hallmark_pathways,
  out_prefix = "pseudobulk_41BB-L_vs_rest",
  results_dir = results_dir,
  plot_dir = plot_dir,
  volcano_lfc_cutoff = 1,
  volcano_p_cutoff = 0.05
)

out_pb_rest$gsea %>%
  filter(pval < 0.05) %>%
  select(pathway, NES, pval, padj, size) %>%
  head(20) %>%
  print()
```

### 3) Pseudobulk: 41BB-L vs other CARs only (exclude tumor/STOP)

This is the pseudobulk version of the main CAR-specific contrast:
41BB-L vs (B7H3 + CD28-41BB + CD8-41BB).

```{r pseudobulk-41BBL-vs-other-CARs}
car_levels <- c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB")

coldata_car <- coldata %>%
  filter(condition %in% car_levels) %>%
  droplevels()

pb_counts_car <- pb_counts[, rownames(coldata_car)]

coldata_car$cond2 <- ifelse(coldata_car$condition == "41BB-L", "41BB-L", "other_CAR")

out_pb_othercar <- run_pseudobulk_fgsea(
  pb_counts = pb_counts_car,
  coldata = coldata_car,
  cond2_levels = c("other_CAR", "41BB-L"),
  hallmark_pathways = hallmark_pathways,
  out_prefix = "pseudobulk_41BB-L_vs_otherCAR",
  results_dir = results_dir,
  plot_dir = plot_dir
)

out_pb_othercar$gsea %>%
  filter(pval < 0.05) %>%
  select(pathway, NES, pval, padj, size) %>%
  head(20) %>%
  print()
```

### 4) Pseudobulk: 41BB-L vs each CAR (pairwise)

This section runs pseudobulk DESeq2 + fgsea for:

- 41BB-L vs B7H3
- 41BB-L vs CD28-41BB
- 41BB-L vs CD8-41BB

```{r pseudobulk-41BBL-vs-each-CAR}
cars_to_compare <- c("B7H3", "CD28-41BB", "CD8-41BB")
gsea_pb_each <- list()

for (car in cars_to_compare) {

  keep_samples <- rownames(coldata_car)[coldata_car$condition %in% c("41BB-L", car)]
  coldata_2 <- coldata_car[keep_samples, , drop = FALSE]
  pb_2 <- pb_counts_car[, keep_samples]

  coldata_2$cond2 <- as.character(coldata_2$condition)

  out <- run_pseudobulk_fgsea(
    pb_counts = pb_2,
    coldata = coldata_2,
    cond2_levels = c(car, "41BB-L"),
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_41BB-L_vs_", car),
    results_dir = results_dir,
    plot_dir = plot_dir,
    volcano_lfc_cutoff = 1,
    volcano_p_cutoff = 0.05
  )

  gsea_pb_each[[car]] <- out$gsea %>% mutate(contrast = paste0("41BB-L vs ", car))

  cat("\n==== Pseudobulk: 41BB-L vs ", car, " ====\n", sep = "")
  print(
    out$gsea %>%
      filter(pval < 0.05) %>%
      select(pathway, NES, pval, padj, size) %>%
      head(10)
  )
}

gsea_pb_each_df <- bind_rows(gsea_pb_each)
write_fgsea_csv(gsea_pb_each_df,
                file.path(results_dir, "GSEA_pseudobulk_41BB-L_vs_eachCAR_hallmark.csv"))
```

```{r session-info}
sessionInfo()
```
