---
title: "11-subtype-pseudobulk-gsea.Rmd"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
params:
  input_rds: "cart_myeloid_subtypes.rds"
  subtype_label: "myeloid"
  condition_levels: !r c("tumor", "STOP", "B7H3", "41BB-L", "CD8-41BB", "CD28-41BB")
  car_levels: !r c("41BB-L", "B7H3", "CD28-41BB", "CD8-41BB")
  minSize: 15
  maxSize: 500
  volcano_lfc_cutoff: 1
  volcano_p_cutoff: 0.05
  gsea_p_cutoff: 0.05
---

## Overview

This notebook performs **gene set enrichment analysis (GSEA)** for a user-specified cell subtype across treatment conditions using **pseudobulk differential expression** and MSigDB Hallmark gene sets.

The notebook is designed to be **subtype-agnostic**:  

- the input Seurat object is provided through `params$input_rds`  
- the subtype name used in output files is provided through `params$subtype_label`  

### Workflow

**Pseudobulk DE → fgsea (DESeq2)**  

- Aggregates raw counts per **sample** to create a pseudobulk count matrix (genes × samples).  
- Runs DESeq2 to obtain a ranked statistic (Wald `stat`) for each contrast.  
- Runs `fgsea()` on the DESeq2 ranking.  
- This approach is more sample-aware because it compares **samples**, not individual cells.

### Contrasts computed

For the selected subtype, the script runs:

1. **Each condition vs rest**  
   - each condition compared against all other conditions

2. **Each condition vs other CARs**  
   - if the condition is a CAR condition, it is compared against the other CAR conditions  
   - if the condition is a non-CAR condition (e.g. tumor, STOP), it is compared against all CAR conditions

### Outputs

Results are written to `results_dir` as CSV files:  

- `<subtype>_DESeq2_*.csv` (pseudobulk DESeq2 DE)  
- `<subtype>_GSEA_*.csv` (fgsea results)  

Plots are written to `plot_dir` as PDF files:

- `<subtype>_volcano_*.pdf`  
- `<subtype>_gsea_dot_*.pdf`  

Note: `fgsea` returns list-columns (e.g., `leadingEdge`). These are serialized (semicolon-separated) before writing CSV files.  

## Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load libraries
library(rprojroot)
library(dplyr)
library(tibble)
library(Seurat)
library(SeuratObject)
library(ggplot2)
library(msigdbr)
library(fgsea)
library(Matrix)
library(DESeq2)
library(stringr)
library(EnhancedVolcano)
library(purrr)

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v4")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# set random seed
set.seed(1234)
```

## Load data and define identities

```{r load-data}
subtype_label <- params$subtype_label

obj <- readRDS(file.path(results_dir, params$input_rds))

# Make condition a factor with explicit order so contrasts are stable/reproducible
condition_levels <- intersect(params$condition_levels, unique(as.character(obj$condition)))

obj$condition <- factor(obj$condition, levels = condition_levels)
Idents(obj) <- "condition"

all_conditions <- levels(obj$condition)
car_levels <- intersect(params$car_levels, all_conditions)

cat("Subtype label:", subtype_label, "\n")
cat("Input object:", params$input_rds, "\n")
cat("Conditions:", paste(all_conditions, collapse = ", "), "\n")
cat("CAR conditions:", paste(car_levels, collapse = ", "), "\n")

table(obj$condition)
```

## Load MSigDB Hallmark gene sets

```{r load-msigdb-hallmark}
hallmark_pathways <- msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(x = .$gene_symbol, f = .$gs_name)

length(hallmark_pathways)
```

## Helper functions

```{r helper-functions}
# Write fgsea results to CSV
write_fgsea_csv <- function(df, out_path) {
  df2 <- df %>%
    mutate(
      leadingEdge = if ("leadingEdge" %in% colnames(df)) {
        vapply(leadingEdge, function(x) paste(x, collapse = ";"), character(1))
      } else {
        NA_character_
      }
    )
  write.csv(df2, out_path, row.names = FALSE)
}

# Volcano plot
plot_volcano <- function(df,
                         gene_col = "gene",
                         lfc_col,
                         p_col,
                         title,
                         subtitle = "",
                         caption = NULL,
                         p_cutoff = 0.05,
                         lfc_cutoff = 1,
                         label_top_n = 10) {
  stopifnot(all(c(gene_col, lfc_col, p_col) %in% colnames(df)))

  df2 <- df %>%
    mutate(.p = .data[[p_col]], .lfc = .data[[lfc_col]]) %>%
    filter(is.finite(.p), is.finite(.lfc), !is.na(.p), !is.na(.lfc))

  label_df <- df2 %>%
    filter(.p <= p_cutoff, abs(.lfc) >= lfc_cutoff) %>%
    arrange(.p) %>%
    slice_head(n = label_top_n)

  lab_vec <- ifelse(df2[[gene_col]] %in% label_df[[gene_col]], df2[[gene_col]], "")

  if (is.null(caption)) {
    caption <- paste0("total = ", nrow(df2), " genes")
  }

  EnhancedVolcano::EnhancedVolcano(
    df2,
    lab = lab_vec,
    x = lfc_col,
    y = p_col,
    title = title,
    subtitle = subtitle,
    caption = caption,
    pCutoff = p_cutoff,
    FCcutoff = lfc_cutoff,
    legendPosition = "bottom",
    pointSize = 1.5,
    colAlpha = 0.8,
    labSize = 3,
    drawConnectors = TRUE,
    widthConnectors = 0.4,
    colConnectors = "black",
    max.overlaps = Inf
  )
}

# GSEA dotplot (activated vs suppressed)
plot_gsea_dot <- function(gsea_df,
                          title = NULL,
                          p_cutoff = 0.05,
                          top_n = 10,
                          use_p = c("pval", "padj")) {
  use_p <- match.arg(use_p)
  stopifnot(all(c("pathway", "NES", use_p, "size") %in% colnames(gsea_df)))

  df <- gsea_df %>%
    mutate(
      p = .data[[use_p]],
      direction = ifelse(NES >= 0, "activated", "suppressed"),
      Count = if ("leadingEdge" %in% colnames(gsea_df)) lengths(leadingEdge) else NA_integer_,
      GeneRatio = if ("leadingEdge" %in% colnames(gsea_df)) (lengths(leadingEdge) / size) else NA_real_,
      pathway_clean = str_remove(pathway, "^HALLMARK_")
    ) %>%
    filter(is.finite(NES), is.finite(p))

  df2 <- df %>%
    filter(p <= p_cutoff) %>%
    group_by(direction) %>%
    arrange(p) %>%
    slice_head(n = top_n) %>%
    ungroup()

  if (nrow(df2) == 0) {
    message("No pathways with ", use_p, " <= ", p_cutoff, " for ", title, ".")
    return(NULL)
  }

  df2 <- df2 %>%
    group_by(direction) %>%
    mutate(pathway_clean = factor(pathway_clean, levels = rev(unique(pathway_clean)))) %>%
    ungroup()

  ggplot(df2, aes(x = GeneRatio, y = pathway_clean)) +
    geom_point(aes(size = Count, color = p), alpha = 0.9) +
    facet_grid(. ~ direction, scales = "free_y", space = "free_y") +
    scale_color_gradient(low = "red", high = "blue", trans = "reverse") +
    labs(
      x = "GeneRatio",
      y = NULL,
      size = "Count",
      color = use_p,
      title = title
    ) +
    guides(
      size = guide_legend(order = 1),
      color = guide_colorbar(order = 2)
    ) +
    theme_bw() +
    theme(
      strip.background = element_rect(fill = "grey90", color = NA),
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Run pseudobulk DESeq2 + fgsea for one contrast
run_pseudobulk_fgsea <- function(pb_counts,
                                 coldata,
                                 cond2_levels,
                                 hallmark_pathways,
                                 out_prefix,
                                 subtype_label,
                                 results_dir,
                                 plot_dir,
                                 minSize = 15,
                                 maxSize = 500,
                                 volcano_lfc_cutoff = 1,
                                 volcano_p_cutoff = 0.05,
                                 gsea_p_cutoff = 0.05) {
  coldata$cond2 <- factor(coldata$cond2, levels = cond2_levels)

  dds <- DESeqDataSetFromMatrix(
    countData = round(pb_counts),
    colData = coldata,
    design = ~ cond2
  )

  dds <- DESeq(dds)

  res <- results(dds, contrast = c("cond2", cond2_levels[2], cond2_levels[1]))

  res_df <- as.data.frame(res) %>%
    tibble::rownames_to_column("gene") %>%
    filter(!is.na(stat), is.finite(stat)) %>%
    arrange(padj)

  write.csv(
    res_df,
    file.path(results_dir, paste0(subtype_label, "_DESeq2_", out_prefix, ".csv")),
    row.names = FALSE
  )

  ranks <- setNames(res_df$stat, res_df$gene)
  ranks <- sort(ranks, decreasing = TRUE)

  set.seed(1234)
  gsea <- fgsea(
    pathways = hallmark_pathways,
    stats = ranks,
    minSize = minSize,
    maxSize = maxSize
  ) %>%
    arrange(pval)

  write_fgsea_csv(
    gsea,
    file.path(results_dir, paste0(subtype_label, "_GSEA_", out_prefix, "_hallmark.csv"))
  )

  pvol <- plot_volcano(
    df = res_df,
    gene_col = "gene",
    lfc_col = "log2FoldChange",
    p_col = "pvalue",
    title = out_prefix,
    p_cutoff = volcano_p_cutoff,
    lfc_cutoff = volcano_lfc_cutoff,
    label_top_n = 10
  )
  ggsave(
    file.path(plot_dir, paste0(subtype_label, "_volcano_", out_prefix, ".pdf")),
    pvol, width = 8, height = 6
  )

  pgsea <- plot_gsea_dot(
    gsea_df = gsea,
    title = paste0("GSEA: ", out_prefix),
    p_cutoff = gsea_p_cutoff,
    top_n = 10,
    use_p = "pval"
  )

  if (!is.null(pgsea)) {
    ggsave(
      file.path(plot_dir, paste0(subtype_label, "_gsea_dot_", out_prefix, ".pdf")),
      pgsea, width = 8, height = 6
    )
  }

  list(de = res_df, gsea = gsea, volcano = pvol, gsea_dot = pgsea)
}

# Generic pseudobulk contrast wrapper
run_pseudobulk_contrast <- function(pb_counts,
                                    coldata,
                                    case_condition,
                                    ref_conditions,
                                    ref_label,
                                    hallmark_pathways,
                                    out_prefix,
                                    subtype_label,
                                    results_dir,
                                    plot_dir,
                                    minSize,
                                    maxSize,
                                    volcano_lfc_cutoff,
                                    volcano_p_cutoff,
                                    gsea_p_cutoff) {
  keep_conditions <- c(case_condition, ref_conditions)

  keep_samples <- rownames(coldata)[coldata$condition %in% keep_conditions]
  coldata_sub <- coldata[keep_samples, , drop = FALSE]
  pb_sub <- pb_counts[, keep_samples, drop = FALSE]

  coldata_sub$cond2 <- ifelse(coldata_sub$condition == case_condition, case_condition, ref_label)

  run_pseudobulk_fgsea(
    pb_counts = pb_sub,
    coldata = coldata_sub,
    cond2_levels = c(ref_label, case_condition),
    hallmark_pathways = hallmark_pathways,
    out_prefix = out_prefix,
    subtype_label = subtype_label,
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = minSize,
    maxSize = maxSize,
    volcano_lfc_cutoff = volcano_lfc_cutoff,
    volcano_p_cutoff = volcano_p_cutoff,
    gsea_p_cutoff = gsea_p_cutoff
  )
}
```

## Build pseudobulk counts per sample
This section aggregates raw UMI counts across all cells in the input subtype object belonging to the same sample, yielding a genes × samples count matrix.

```{r build-pseudobulk, message=FALSE, warning=FALSE}
stopifnot(all(c("sample", "condition") %in% colnames(obj@meta.data)))

counts <- GetAssayData(obj, assay = "RNA", layer = "counts")

meta <- obj@meta.data %>%
  dplyr::select(sample, condition) %>%
  mutate(
    sample = as.character(sample),
    condition = factor(condition, levels = levels(obj$condition))
  )

# Aggregate counts by sample (genes x samples)
pb_counts <- Matrix::t(rowsum(Matrix::t(counts), group = meta$sample))

# Sample-level metadata (one row per sample)
coldata <- meta %>%
  as.data.frame() %>%
  { rownames(.) <- NULL; . } %>%
  distinct(sample, condition) %>%
  tibble::column_to_rownames("sample")

# Align matrices
pb_counts <- pb_counts[, rownames(coldata)]

table(coldata$condition)
```

## 1) Each condition vs rest

```{r each-condition-vs-rest, message=FALSE, warning=FALSE}
pb_rest_results <- list()

for (cond in all_conditions) {
  ref_conditions <- setdiff(all_conditions, cond)
  out_prefix <- paste0(cond, "_vs_rest")

  cat("\n===== Running each-vs-rest:", cond, "vs rest =====\n")

  pb_rest_results[[cond]] <- run_pseudobulk_contrast(
    pb_counts = pb_counts,
    coldata = coldata,
    case_condition = cond,
    ref_conditions = ref_conditions,
    ref_label = "rest",
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_", out_prefix),
    subtype_label = subtype_label,
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = params$minSize,
    maxSize = params$maxSize,
    volcano_lfc_cutoff = params$volcano_lfc_cutoff,
    volcano_p_cutoff = params$volcano_p_cutoff,
    gsea_p_cutoff = params$gsea_p_cutoff
  )
}
```

## 2) Each condition vs other CARs

For CAR conditions:

- compare to the other CAR conditions

For non-CAR conditions:

- compare to all CAR conditions

```{r each-condition-vs-other-cars, message=FALSE, warning=FALSE}
pb_car_results <- list()

for (cond in all_conditions) {

  if (cond %in% car_levels) {
    ref_conditions <- setdiff(car_levels, cond)
    ref_label <- "otherCAR"
    out_prefix <- paste0(cond, "_vs_otherCAR")
  } else {
    ref_conditions <- car_levels
    ref_label <- "allCAR"
    out_prefix <- paste0(cond, "_vs_allCAR")
  }

  if (length(ref_conditions) == 0) next

  cat("\n===== Running condition-vs-CARs:", cond, "=====\n")

  pb_car_results[[cond]] <- run_pseudobulk_contrast(
    pb_counts = pb_counts,
    coldata = coldata,
    case_condition = cond,
    ref_conditions = ref_conditions,
    ref_label = ref_label,
    hallmark_pathways = hallmark_pathways,
    out_prefix = paste0("pseudobulk_", out_prefix),
    subtype_label = subtype_label,
    results_dir = results_dir,
    plot_dir = plot_dir,
    minSize = params$minSize,
    maxSize = params$maxSize,
    volcano_lfc_cutoff = params$volcano_lfc_cutoff,
    volcano_p_cutoff = params$volcano_p_cutoff,
    gsea_p_cutoff = params$gsea_p_cutoff
  )
}
```

```{r session-info}
sessionInfo()
```
