---
title: "14-dc-reclustering.Rmd"
author: "Bicna Song"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
---

## Load packages
```{r load-packages, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(rprojroot)
  library(dplyr)
  library(Seurat)
  library(patchwork)
  library(ggplot2)
  library(tidyverse)
  library(SeuratObject)
  library(scCustomize)
  library(AUCell)
  library(msigdbr)
  library(pheatmap)
})

# Define directories
root_dir <- find_root(has_dir(".git"))
data_dir <- file.path(root_dir, "data", "v4")
analysis_dir <- file.path(root_dir, "analyses", "murine-cart-tumor")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

# Source custom theme
source(file.path(root_dir, "figures", "theme.R"))

# Set seed for reproducibility
set.seed(1234)
```

## Load dc object
```{r load-data}
dc_obj <- readRDS(file.path(data_dir, "cart_dc_subtypes.rds"))

table(dc_obj$dc_subtype)
```

## Exclude myeloid and lymphocytes cells
```{r subset-dc}
exclude_subtypes <- c("Myeloid cells", "Lymphocytes")

dc_obj <- subset(
  dc_obj,
  subset = !dc_subtype %in% exclude_subtypes
)

dc_obj$dc_subtype <- droplevels(dc_obj$dc_subtype)

table(dc_obj$dc_subtype)
```

## Reprocess and re-cluster filtered cells (No integration)
```{r recluster-dc, message=FALSE, warning=FALSE}
set.seed(1234)

## IMPORTANT:
## Do NOT split RNA by sample when not integrating
## (splitting is only needed for IntegrateLayers)
DefaultAssay(dc_obj) <- "RNA"

## Standard preprocessing
dc_obj <- NormalizeData(dc_obj)
dc_obj <- FindVariableFeatures(dc_obj, selection.method = "vst", nfeatures = 2000)
dc_obj <- ScaleData(dc_obj)
dc_obj <- RunPCA(dc_obj, npcs = 30)

## Inspect dimensionality
ElbowPlot(dc_obj, ndims = 30)

## PCA-based UMAP (no integration)
dc_obj <- RunUMAP(dc_obj, reduction = "pca", dims = 1:20)

DimPlot(dc_obj, reduction = "umap", group.by = "sample")

## Graph construction and clustering (PCA space)
dc_obj <- FindNeighbors(
  dc_obj,
  reduction = "pca",
  dims = 1:20
)

dc_obj <- FindClusters(
  dc_obj,
  resolution = 0.5
)
```

## UMAP of re-clustered filtered cells
```{r umap-reclustered, fig.height=6, fig.width=7}
p_umap_clusters <- DimPlot(
  dc_obj,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  label.size = 4
) +
  ggtitle("Re-clustered DC subclusters")

p_umap_clusters

ggsave(
  filename = file.path(plot_dir, "filtered_dc_reclustered_umap.pdf"),
  plot = p_umap_clusters,
  width = 7,
  height = 6,
  device = cairo_pdf
)
```

## UMAP split by treatment
```{r umap-by-treatment, fig.height=6, fig.width=12}
p_umap_split <- DimPlot(
  dc_obj,
  reduction = "umap",
  group.by = "seurat_clusters",
  split.by = "condition",
  ncol = 3,
  label = FALSE
) +
  ggtitle("Re-clustered DC subclusters by treatment")

p_umap_split

ggsave(
  filename = file.path(plot_dir, "filtered_dc_reclustered_umap_by_treatment.pdf"),
  plot = p_umap_split,
  width = 12,
  height = 6,
  device = cairo_pdf
)
```

## Generate cluster markers
```{r find-markers, message=FALSE, warning=FALSE}
cluster_markers <- FindAllMarkers(
  dc_obj,
  assay = "RNA",
  slot = "data",
  only.pos = TRUE,
  min.pct = 0.1,
  logfc.threshold = 0.5,
  random.seed = 1234,
  max.cells.per.ident = Inf
)

write.csv(
  cluster_markers,
  file.path(results_dir, "filtered_dc_reclustered_markers.csv"),
  row.names = FALSE
)
```

## Top 30 markers per cluster
```{r top30-markers}
top30_markers <- cluster_markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 30, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(cluster, desc(avg_log2FC))

write.csv(
  top30_markers,
  file.path(results_dir, "filtered_dc_reclustered_top30_markers_per_cluster.csv"),
  row.names = FALSE
)

top30_markers
```

## UMAP colored by previous subtype labels
```{r umap-old-labels, fig.height=6, fig.width=8}
p_old_labels <- DimPlot(
  dc_obj,
  reduction = "umap",
  group.by = "dc_subtype",
  label = TRUE,
  label.size = 3
) +
  ggtitle("Filtered cells colored by previous subtype labels")

p_old_labels

ggsave(
  filename = file.path(plot_dir, "filtered_dc_old_subtype_labels_on_reclustered_umap.pdf"),
  plot = p_old_labels,
  width = 8,
  height = 6,
  device = cairo_pdf
)
```

## New clusters vs old subtype labels
```{r cluster-vs-old-labels}
cluster_label_df <- dc_obj@meta.data %>%
  dplyr::filter(!is.na(seurat_clusters), !is.na(dc_subtype)) %>%
  dplyr::count(seurat_clusters, dc_subtype, name = "n") %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::mutate(prop_within_cluster = n / sum(n)) %>%
  dplyr::ungroup()

cluster_label_df
```

## Top old label per new cluster
```{r top-old-label-per-cluster}
cluster_label_summary <- cluster_label_df %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::slice_max(order_by = prop_within_cluster, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::rename(
    matched_old_label = dc_subtype,
    matched_count = n,
    matched_prop = prop_within_cluster
  ) %>%
  dplyr::arrange(as.numeric(as.character(seurat_clusters)))

write.table(
  cluster_label_summary,
  file = file.path(results_dir, "filtered_dc_top_old_label_per_new_cluster.tsv"),
  sep = "	",
  quote = FALSE,
  row.names = FALSE
)

cluster_label_summary
```

## Barplot of top old label match per new cluster
```{r top-old-label-barplot, fig.height=5, fig.width=9}
p_cluster_label_summary <- ggplot(
  cluster_label_summary,
  aes(
    x = seurat_clusters,
    y = matched_prop,
    fill = matched_old_label
  )
) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(matched_prop, accuracy = 0.1)),
    vjust = -0.3,
    size = 3
  ) +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  ) +
  xlab("New Seurat cluster") +
  ylab("Matched proportion") +
  labs(fill = "Top old label") +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1.05)
  ) +
  ggtitle("Top old subtype label match per new cluster")

p_cluster_label_summary

ggsave(
  filename = file.path(plot_dir, "filtered_dc_top_old_label_per_new_cluster_barplot.pdf"),
  plot = p_cluster_label_summary,
  width = 9,
  height = 5,
  device = cairo_pdf
)
```

## Heatmap of new clusters vs old subtype labels
```{r cluster-vs-old-label-heatmap, fig.height=6, fig.width=10}
p_cluster_label_heatmap <- ggplot(
  cluster_label_df,
  aes(x = dc_subtype, y = seurat_clusters, fill = prop_within_cluster)
) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#2C7FB8") +
  theme_Publication() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  xlab("Previous subtype label") +
  ylab("New Seurat cluster") +
  labs(fill = "Within-cluster
proportion") +
  ggtitle("Mapping of new clusters to previous subtype labels")

p_cluster_label_heatmap

ggsave(
  filename = file.path(plot_dir, "filtered_dc_cluster_vs_old_subtype_heatmap.pdf"),
  plot = p_cluster_label_heatmap,
  width = 8,
  height = 6,
  device = cairo_pdf
)
```

## Old subtype key markers
```{r}
## DC subtype key markers (updated)
old_subtype_markers <- list(
  "cDC1" = c("Xcr1", "Clec9a", "Wdfy4", "Mycl", "Itgae"),
  "cDC2" = c("Ciita", "H2-Ab1", "H2-Oa", "Mgl2","Il1b"),
  "Migratory DC" = c("Ccr7", "Cd80", "Fscn1", "Ccl22", "Il15ra"),
  "Inflammatory DC (IL12+)" = c("Il12b", "Ly75", "Fscn1","Scin")
)
```

## Marker validation plots for each old subtype
```{r old-subtype-marker-validation, fig.height=6, fig.width=8}
vln_plot_list <- list()

for (subtype_name in names(old_subtype_markers)) {
  markers_use <- old_subtype_markers[[subtype_name]]
  markers_use <- markers_use[markers_use %in% rownames(dc_obj)]

  if (length(markers_use) == 0) next

  p_vln <- VlnPlot(
    dc_obj,
    features = markers_use,
    group.by = "seurat_clusters",
    pt.size = 0,
    ncol = 2
  ) +
    plot_annotation(title = paste0(subtype_name, " key markers across new clusters"))

  vln_plot_list[[subtype_name]] <- p_vln
}

pdf(
  file = file.path(plot_dir, "filtered_dc_old_subtype_marker_violinplots.pdf"),
  width = 12,
  height = 8,
  onefile = TRUE
)
for (plot_name in names(vln_plot_list)) {
  print(vln_plot_list[[plot_name]])
}
dev.off()

# Print the last plot in the knitted document as a visual check
if (length(vln_plot_list) > 0) {
  vln_plot_list[[length(vln_plot_list)]]
}
```

## Stacked barplot of subtype proportions
```{r subtype-stacked-barplot}
prop_df <- dc_obj@meta.data %>%
  dplyr::filter(!is.na(condition), !is.na(dc_subtype)) %>%
  dplyr::count(condition, cell_group = dc_subtype, name = "n") %>%
  dplyr::group_by(condition) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

write.table(
  prop_df,
  file = file.path(results_dir, "filtered_dc_reclustered_proportions.tsv"),
  sep = "	",
  quote = FALSE,
  row.names = FALSE
)

p_prop <- ggplot(prop_df, aes(x = condition, y = prop, fill = cell_group)) +
  geom_bar(stat = "identity") +
  theme_Publication() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background  = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_manual(
    values = DiscretePalette(n = length(unique(prop_df$cell_group)))
  ) +
  labs(fill = NULL) +
  ylab("Proportion") +
  xlab("Condition") +
  scale_y_continuous(labels = scales::percent_format())

p_prop

ggsave(
  filename = file.path(plot_dir, "filtered_dc_reclustered_proportions_stacked_barplot.pdf"),
  plot = p_prop,
  width = 8,
  height = 5,
  device = cairo_pdf
)
```

## Save object
```{r save-object}
saveRDS(
  dc_obj,
  file.path(results_dir, "cart_dc_reclustered.rds")
)
```

# Gene set enrichment analysis

## AUCell scoring on MSigDB Hallmark Gene sets
```{r get-genesets, message=FALSE, warning=FALSE}
# msigdb hallmark
msigdb_hallmark <- msigdbr::msigdbr(species = "Mus musculus", category = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  dplyr::group_by(gs_name) %>%
  dplyr::summarise(genes = list(gene_symbol)) %>%
  deframe()
```

## Hallmark gene set activity by dc subcluster
```{r hallmark-activity, message=FALSE, warning=FALSE}
# AUCell on hallmark genesets
cells_rankings <- AUCell_buildRankings(GetAssayData(dc_obj, assay = "RNA", layer = "data"), nCores = 1, plotStats = FALSE)
hallmark_AUC <- AUCell_calcAUC(msigdb_hallmark, cells_rankings)

# add hallmark scores to seurat object
hallmark_scores_df <- as.data.frame(t(getAUC(hallmark_AUC)))
dc_obj <- AddMetaData(dc_obj, metadata = hallmark_scores_df)
```

# Subcluster vs pathway heatmap of AUCell scores
```{r subcluster-pathway-heatmap, fig.height=8, fig.width=16}
# z scores of AUCell scores across clusters per gene set
hallmark_scores_df$cluster = dc_obj$seurat_clusters

avg_hallmark_scores <- hallmark_scores_df %>%
  group_by(cluster) %>%
  summarise(across(starts_with("HALLMARK"), mean)) %>%
  column_to_rownames("cluster")

hallmark_z_scores <- scale(avg_hallmark_scores)

rownames(hallmark_z_scores) <- paste0("Cluster", rownames(hallmark_z_scores))

# Color scale
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)
breaks <- seq(-3, 3, length.out = 101)

# save heatmap as pdf
pheatmap::pheatmap(
  mat = hallmark_z_scores,
  color = colors.use,
  border_color = "white",
  breaks = breaks,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  main = "Hallmark Pathway AUCell Activity across DC Subclusters",
  filename = file.path(plot_dir, "filtered_dc_subcluster_hallmark_auc_heatmap.pdf"),
  width = 16,
  height = 8
)
```

## Session info
```{r session-info}
sessionInfo()
```
